---
title: "clustering and sub-clustering with each main group"
author: "lily yang"
date: "`r Sys.Date()`"
output: html_document
---

```{r, setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

work_dir <- Sys.getenv("work_dir")
bsp_num <- Sys.getenv("bsp_num")
sampSet <- Sys.getenv("sampSet")
clustK <- Sys.getenv("clustK")
reference_genome_version <- Sys.getenv("reference_genome_version")

knitr::opts_knit$set(root.dir = work_dir)

current_run  <- sampSet

if (clustK == "") {
   clustK <- 20          # defaulty
}
print(clustK)

print(paste('work_dir', work_dir, 'sampSet', sampSet, bsp_num))              

```



This script picks up where "differential_lily_uniform.Rmd" left off.

We will load the data, subset into different main cell type, re-cluster, remove doublet clusters, evaluate cluster quality, etc, in order to come up with the best dataset for annotation.

This script is usually run on the hpc, see "code/3_html/muci_html_cluster_subset.sh".


# first Clustering and evaluation

## Reading in the data


Load the data.

```{r load processed data and subset overview cell number, fig.width=10}
library(SingleCellExperiment)
library(DropletUtils)
library(scran)
library(scater)
library(uwot)
library(biomaRt)
library(scds)
library(batchelor)
library(ggplot2)
library(ggbeeswarm)
library(bluster)
library(scran)
library(scater)


sce_all <- readRDS("sce_all_with_named_subcluster.rds")
sce_all_corrected <- readRDS("sce_all_corrected_with_named_subcluster.rds")

# ensure sample20 has been deleted 
library(readxl)
population_divide <- read_excel("Population devision.xlsx")
selected_main_group <-  as.character(population_divide[1, 1])
selected_cluster <- as.character(population_divide[1, 3])
selected_cluster <- strsplit(selected_cluster, split = ",")[[1]]

library(stringr)
selected_cluster <- str_replace_all(selected_cluster, fixed(" "), "")
# save.image('cluster_subset_uniform_debug.RData')
# load('cluster_subset_uniform_debug.RData')
sce_main_group <- sce_all[ , (sce_all$cluster %in% selected_cluster) ]
table(sce_main_group$Sample)
overview_cell_number_per_sample <- function(sce, flag)
{
  library(ggplot2)
  df <- as.data.frame(table(sce$Sample)) 
  colnames(df) <- c("Sample_id","n_cells")
  
  
  dat_2 <- data.frame(Sample_id = sce$Sample,
                      cull_date = sce$cull_date,
                      condition = sce$condition)
  dat_2 <- unique(dat_2)  
  dat  <- base:::merge(df, dat_2, all.x = TRUE)
  
  ggplot(data=dat, aes(x=Sample_id, y=n_cells,fill = Sample_id)) +
    geom_bar(stat="identity") +
    geom_text(aes(label=n_cells), vjust=1, size=3.5) +
    ggtitle(paste0(flag, ", cell number summary")) +
    facet_wrap(cull_date ~ condition,  scales="free_x") +
    theme(axis.text.x = element_text(angle = 90)) 
  
}

# my_bar_plot(sce_run_x,"before batch effect correction")
overview_cell_number_per_sample(sce_main_group, flag = paste('main group:', selected_main_group))

```

```{r re-normarlize sce subset and batch effect correction, then visulize doublet situation in UMAP, fig.width=10}


clust <- quickCluster(sce_main_group)
size_factors <- calculateSumFactors(sce_main_group, cluster=clust)
# log transfrom gene counts  
sce_main_group <- logNormCounts(sce_main_group, size_factors=size_factors)
varmod <- modelGeneVar(sce_main_group, block=sce_main_group$run)
# Model the variance of the log-expression profiles for each gene, decomposing it into technical and biological components based on a fitted mean-variance trend.
vargenes <- getTopHVGs(varmod, n=5000)

sce_main_group <- fixedPCA(sce_main_group, subset.row=vargenes)

set.seed(100)
sce_main_group <- runUMAP(sce_main_group, dimred='PCA')
sce_main_group <- runTSNE(sce_main_group, dimred='PCA')
library(batchelor)
set.seed(100)
sce_main_group_corrected <- fastMNN(sce_main_group, batch=sce_main_group$run, subset.row = NULL, get.variance = TRUE)
sce_main_group_corrected$Sample <- sce_main_group$Sample
sce_main_group_corrected$condition <- sce_main_group$condition
sce_main_group_corrected$run <- sce_main_group$run
sce_main_group_corrected$cull_date <- sce_main_group$cull_date
sce_main_group_corrected$animal_id <- sce_main_group$animal_id  
sce_main_group_corrected$cxds_score <- sce_main_group$cxds_score
sce_main_group_corrected$query_doublet <- sce_main_group$query_doublet
sce_main_group_corrected$sum_counts <- sce_main_group$sum_counts
sce_main_group_corrected$genes_detected <- sce_main_group$genes_detected
sce_main_group_corrected$mito_proportion <- sce_main_group$mito_proportion
sce_main_group_corrected$sum_counts_level <- sce_main_group$sum_counts_level

rowdat <- rowData(sce_main_group_corrected)
rowdat <- cbind(rowData(sce_main_group)[rownames(sce_main_group_corrected),], rowdat )
rowData(sce_main_group_corrected) <- rowdat
set.seed(100)
sce_main_group_corrected <- runUMAP(sce_main_group_corrected, dimred="corrected")
sce_main_group_corrected <- runTSNE(sce_main_group_corrected, dimred="corrected")


```


## Clustering the data

Now we re cluster the data, look at sample contribution to each cluster, and find marker genes for each cluster, which might be used in cell type annotation step.

We'll use the graph-based clustering implemented in bluster, specifically using jaccard index to weight edges and the louvain method for community detection, with a k value of `r clustK`. This is largely because this combination gave us a reasonable number of clusters with decent quality metrics after trying a variety of different options. Thus we will run this on the batch-corrected data and look at distribution of samples across clusters and top marker genes for each cluster.


### Cluster, check and save DE genes


```{r re-cluster subset}
library(bluster)
library(scran)
library(scater)
# scran library need to load after some packages, otherwise could not find function "clusterCells"
 
sce_corrected_clust <- clusterCells(sce_main_group_corrected,
                                          use.dimred="corrected",
                                          BLUSPARAM=SNNGraphParam(k=clustK, type="jaccard",cluster.fun="louvain"))

# k An integer scalar specifying the number of nearest neighbors to consider during graph construction.                                                          



table(sce_corrected_clust)
sce_main_group_corrected$cluster <- sce_corrected_clust
sce_main_group$cluster <- sce_corrected_clust


```


### Basic dim reduction plots
```{r create basic UMAP plots, fig.width=10}
UMAP_color_by_cluster <- function(selected_sce, title_flag = "batch effect corrected" )
{
  set.seed(55698)
  umap_plot <- plotUMAP(selected_sce, colour_by="cluster", point_size=0.3,
                        text_by = "cluster") + 
    guides(colour = guide_legend(override.aes = list(size=6))) +
    ggtitle(paste(sampSet, title_flag))
  # Adjust the legend text and title size
  
  
  # Print the plot
  print(umap_plot)
  
  
  TSNE_plot <- plotTSNE(selected_sce, colour_by="cluster", point_size=0.3,
                        text_by = "cluster") + guides(colour = guide_legend(override.aes = list(size=6))) + ggtitle(paste(sampSet, title_flag))
  print(TSNE_plot)  
}

UMAP_color_by_cluster(sce_main_group_corrected, "batch effect corrected" )
UMAP_color_by_cluster(sce_main_group, "original, no batch effect correction")

ReducedDim_plot <-  plotReducedDim(sce_main_group_corrected, dimred='corrected', 
                                   colour_by="cluster", point_size=0.1, ncomponents=5) +
  guides(colour = guide_legend(override.aes = list(size=6)))
print(ReducedDim_plot)


```

```{r draw bar plot to check sample/condition contribution within each cluster, fig.width=10}
assays(sce_main_group_corrected)$counts <- assays(sce_main_group)$counts[rownames(sce_main_group_corrected),]
assays(sce_main_group_corrected)$logcounts <- assays(sce_main_group)$logcounts[rownames(sce_main_group_corrected),]


sce_corrected <- sce_main_group_corrected

# plot the sample composition of each cluster in a bar plot
dat <- as.data.frame(table(sce_corrected$cluster, sce_corrected$Sample))
colnames(dat) <- c('cluster', 'Sample_id', 'n_cells')
dat_2 <- data.frame(Sample_id = sce_corrected$Sample,
                    cull_date = sce_corrected$cull_date,
                    condition = sce_corrected$condition)
dat_2 <- unique(dat_2)  
dat  <- base:::merge(dat, dat_2, all.x = TRUE)




library(ggpubr)
library(ggrepel)
p <- ggplot(dat, aes(fill=Sample_id, y=n_cells, x=cluster)) +
  geom_bar(position='stack', stat='identity') +
  facet_wrap(cull_date ~ condition,  scales="free_x") +
  labs(title=paste0("cluster compostion ",sampSet,"_k",clustK))

print(p)

png(filename = paste0('Results/clustering_', sampSet, '_', selected_main_group, '_cluster_composition_barplot.png') , 
    height=1800, width=2600, res = 300)
print(p)
dev.off()  

# draw a bar plot to visulize the composition of each virus within each cluster
# colour cluster plot by condition
ggplot(dat, aes(fill=condition, y=n_cells, x=cluster)) +
  geom_bar(position='stack', stat='identity') +
  facet_wrap( ~ cull_date,  scales="free_x") +
  labs(title=paste0("cluster compostion ",sampSet,"_k",clustK))


```

```{r randomly pick one point among dots whose UMAP location is identical and draw UMAP, fig.width=10}
library(dplyr)

library(scater)
selected_sce <- sce_corrected

draw_UMAP_after_sampling <- function(selected_sce, 
                                     title_flag = "batch effect corrected")
{
  
  # Get UMAP coordinates and sample information
  
  umap_coords <- as.data.frame(reducedDim(selected_sce, "UMAP"))
  
  umap_coords$Sample <- selected_sce$Sample
  umap_coords$condition <- selected_sce$condition
  umap_coords$sum_counts_level <- selected_sce$sum_counts_level
  umap_coords$run <- selected_sce$run
  umap_coords$cull_date <- selected_sce$cull_date
  head(selected_sce$run)
  
  set.seed(100)
  # Group by UMAP coordinates and sample one point from each group
  umap_sampled <- umap_coords %>%
    group_by(UMAP1, UMAP2) %>%
    sample_n(1)
  
  # Plot the UMAP with sampled points
  umap_plot <- ggplot(umap_sampled, aes(x = UMAP1, y = UMAP2, color = Sample)) +
    geom_point(size = 0.1) +
    ggtitle(paste(sampSet, title_flag)) +
    facet_wrap(cull_date ~ condition) +
    guides(colour = guide_legend(override.aes = list(size = 6)))
  
  # Print the plot
  print(umap_plot)
  
  
  # Plot the UMAP with sampled points
  umap_plot <- ggplot(umap_sampled, aes(x = UMAP1, y = UMAP2, color = condition)) +
    geom_point(size = 0.1) +
    facet_wrap( ~ cull_date) +
    ggtitle(paste(sampSet, title_flag)) +
    guides(colour = guide_legend(override.aes = list(size = 6)))
  
  # Print the plot
  print(umap_plot)
  
  umap_plot <- ggplot(umap_sampled, aes(x = UMAP1, y = UMAP2, color = sum_counts_level)) +
    geom_point(size = 0.1) +
    ggtitle(paste(sampSet, title_flag)) +
    facet_wrap(cull_date ~ condition) +
    guides(colour = guide_legend(override.aes = list(size = 6)))
  
  # Print the plot
  print(umap_plot)
  
  umap_plot <- ggplot(umap_sampled, aes(x = UMAP1, y = UMAP2, color = run)) +
    geom_point(size = 0.1) +
    ggtitle(paste(sampSet, title_flag)) +
    guides(colour = guide_legend(override.aes = list(size = 6)))
  
  # Print the plot
  print(umap_plot)  
}
draw_UMAP_after_sampling(sce_corrected, "batch effect corrected" )


```



## Evaluating the clustering quality

We will now run several metrics to evaluate how well our cells are clustered, thinking about whether they have been either over- or under-clustered.

### check silhouette widths:
First we'll look at the silhouette widths, which incorporates both the distances to other cells in the same cluster as well as the distances to cells in other clusters. The plot we make will colour cells according to their nearest cluster if they have a negative silhouette width.

```{r calculate silhouette widths and visilize it}
library("ggbeeswarm")

sil.approx <- approxSilhouette(reducedDim(sce_corrected, "corrected"), clusters=sce_corrected$cluster)
sil.approx

sil.data <- as.data.frame(sil.approx)
sil.data$closest <- factor(ifelse(sil.data$width > 0, sce_corrected$cluster, sil.data$other))
sil.data$cluster <- sce_corrected$cluster
scTkn <- paste0(sampSet,"_k",clustK)
ggplot(sil.data, aes(x=cluster, y=width, colour=closest)) +
  ggbeeswarm::geom_quasirandom(method="smiley") +
  labs(title=scTkn)

```

### check cluster purity:
To better separate the factors that contribute to this metric, we'll divide out purity and heterogeneity, first looking at cluster purity. Here again cells are coloured according to their nearest cluster if their purity is particularly low.

clusters which have low purity, indicate some of cell within these clusters have gene expression profile poorly aligns with the centroid or the majority of cells in a cluster.
```{r calculate cells purity and visualization in UMAP}


########################################################################### 

pure <- neighborPurity(reducedDim(sce_corrected, "corrected"), sce_corrected$cluster)
pure
pure.data <- as.data.frame(pure)
pure.data$maximum <- factor(pure.data$maximum)
pure.data$cluster <- sce_corrected$cluster

ggplot(pure.data, aes(x=cluster, y=purity, colour=maximum)) +
  ggbeeswarm::geom_quasirandom(method="smiley") +
  labs(title=scTkn)
########################################################################### 

# check the UMAP location of cells have low purity values
colData(selected_sce)$purity <- pure.data[colnames(selected_sce), 'purity']
location_purity_UMAP <- function(selected_sce,cluster_n = 8)
{
  p <-  plotUMAP(selected_sce[,selected_sce$cluster==cluster_n], colour_by="purity", point_size=1) + ggtitle(paste('cluster',cluster_n))
  # plotUMAP(selected_sce, colour_by="cluster", point_size=1)
  print(p)
}

# visulize selected unpure clusters, color their cells by purity values to see the distribution features of low purity cells

# location_purity_UMAP(selected_sce,2)
# table(Cluster=sce_corrected$cluster, pure.data$maximum)

df <- table(Cluster=sce_corrected$cluster, pure.data$maximum)

# Extract diagonal elements
diagonal_values <- diag(as.matrix(df))
# Calculate row sums
row_sums <- rowSums(df)

# Subtract diagonal values from row sums
adjusted_sums <- (row_sums - diagonal_values)
summary_purity_df <- data.frame(misclassifed_counts = adjusted_sums,
                                misclassifed_proportion = (adjusted_sums/diagonal_values))



# df$misclassifed_counts <- adjusted_sums
# df$misclassifed_proportion <- adjusted_sums/diagonal_values
print(df)
print(summary_purity_df)

print(paste(sum(summary_purity_df$misclassifed_counts),sum(summary_purity_df$misclassifed_proportion)))




```


from the table and UMAP, we can see the cells with low purity values always locate at the interface area with adjacent clusters

in UMAP plot, we can find some clusters locate quiet close to other clusters, some cells classified as one cluster actually locate closer to adjacent clusters. which indicate some cells might get mis-classified.


### check heterogeneity within each cluster:
And finally we'll look at the heterogeneity within each cluster by computing the within cluster sum of squares.

```{r check heterogeneity within each cluster and visualization}
########################################################################### 
rmsd <- clusterRMSD(reducedDim(sce_corrected, "corrected"), sce_corrected$cluster)

df <- as.data.frame(rmsd)
df$cluster <- as.factor(1:length(levels(sce_corrected$cluster)))


# Basic barplot
ggplot(data=df, aes(x=cluster, y=rmsd)) +
  geom_bar(stat="identity")+
  geom_text(aes(label=cluster), vjust=-0.3, size=3.5) +
  ggtitle("examine heterogeneity within each cluster")

```



## check suspicious clusters
### check QC metrics all clusters
check QC metrics for all clusters (sum_counts, genes_detected, mito_proportion).
check sample contribution of suspicious cluster, sample distribution in UMAP of suspicious cluster



```{r check QC metrics for all clusters and sample contribution - sample-condition distribution in UMAP of suspicious cluster, fig.width=10}
QC.tab <- data.frame(colData(sce_corrected)[,c('cluster', 'sum_counts', 'mito_proportion', 'genes_detected')])

p <-  ggplot(QC.tab, aes(x=cluster, y=log2(sum_counts) )) +
  ggbeeswarm::geom_quasirandom(method="smiley")
print(p)

p <-  ggplot(QC.tab, aes(x=cluster, y=log2(genes_detected) )) +
  ggbeeswarm::geom_quasirandom(method="smiley")
print(p)

p <-  ggplot(QC.tab, aes(x=cluster, y=mito_proportion)) +
  ggbeeswarm::geom_quasirandom(method="smiley")
print(p)


```

## Doublet inspection

It may be the case that our suspected doublets, which fell in a particular region of the UMAP, have formed their own cluster. This is a convenient means of removing them as we can get rid of the whole cluster and don't have to rationalist a threshold for suspected doublet value.

```{r check Doublet distribution across clusters, fig.width=10}
# print the proportion of estimated Doublet in each cluster
prop.table(table(sce_corrected$cluster, sce_corrected$query_doublet), margin=1)

# visualize cxds_score and estimated Doublet in each cluster

doublet.tab <- data.frame(colData(sce_corrected)[,c('cxds_score', 'cluster', 'query_doublet', 'sum_counts')])

ggplot(doublet.tab, aes(x=cluster, y=cxds_score, colour=query_doublet)) +
    ggbeeswarm::geom_quasirandom(method="smiley")

ggplot(doublet.tab, aes(x=cluster, y=sum_counts, colour=query_doublet)) +
    ggbeeswarm::geom_quasirandom(method="smiley")
```


# save workspace image

# output DE genes for all clusters
Finally, we need to re-run our marker scoring because we have added some clusters. For this, we will only write to file the scores for the clusters we have added.


```{r save top DE genes in excel file for new produced sub cluster}

selected_main_group <- gsub(' ', '_', selected_main_group)
assays(sce_main_group_corrected)$counts <- assays(sce_main_group)$counts[rownames(sce_main_group_corrected),]
assays(sce_main_group_corrected)$logcounts <- assays(sce_main_group)$logcounts[rownames(sce_main_group_corrected),]

sce_corrected <- sce_main_group_corrected


cell_n_cluster <- table(sce_corrected$cluster)


marker.info.corrected <- scoreMarkers(sce_corrected,
                                      groups=sce_corrected$cluster,
                                      block=sce_corrected$run,
                                      assay.type='logcounts',
                                      full.stats=TRUE)



library(openxlsx)
# Create a new workbook
wb <- createWorkbook()

for(i in unique(names(marker.info.corrected)) )
{
  # print(i)
  ord <- marker.info.corrected[[i]][order(marker.info.corrected[[i]]$mean.logFC.cohen, decreasing=TRUE),]
  
  if(bsp_num == "bsp_248")
  {
    if(reference_genome_version == "108")
    {
      ord$symbol <- rowData(sce_corrected)[rownames(ord), 'ENTREZ_symbol']      
    }else if(reference_genome_version == "113")
    {
      ord$symbol <- rowData(sce_corrected)[rownames(ord), 'Symbol']      
    }      
  }
  if(bsp_num == "bsp_247")
  {
    ord$symbol <- rowData(sce_corrected)[rownames(ord), 'Symbol']       
  }
  
  
  pfx <- paste0(work_dir,"markers_",sampSet,'_k',clustK)
  dir.create(pfx, showWarnings = FALSE)
  
  
  
  DE_genes <- cbind(ord$symbol,ord)
  DE_genes <- as.data.frame(ord)
  DE_genes <- cbind(rownames(DE_genes), ord$symbol, DE_genes)
  DE_genes_mid <- DE_genes[ ,c(7,1:6)]
  DE_genes_mid$'mean.AUC' <- DE_genes[ ,'mean.AUC']
  DE_genes <- DE_genes_mid
  colnames(DE_genes)[2:3] <- c('ensemble_id', 'symbols')
  
  sheet_name <- paste("cluster", i, '#',cell_n_cluster[i])
  
  # Add a worksheet to the workbook with the created name
  addWorksheet(wb, sheetName = sheet_name)
  # Write the data frame to the worksheet
  writeData(wb, sheet = sheet_name, DE_genes)
  
}
# Save the workbook to a file
saveWorkbook(wb, file = paste0('Results/clustering_', sampSet, '_',
                               selected_main_group, "_marker_per_cluster.xlsx"), overwrite = TRUE)




```




<!-- ```{r draw heatmap for selected markers across clusters} -->

<!-- if(bsp_num == "bsp_248") -->
<!-- { -->
<!--     defining_genes_185 <- c("CD163", "FLT3", "CD14", "CD4", "CADM1", "CSF1R", "FCGR1A", "BCL6", "CCL17", "IRF8", "ADAM8", "PRDM1", "IRF4", "JCHAIN" , "TNFRSF17", "TNFRSF13B", "CD19", "PAX5", "EBF1", "MS4A1", "CD79B", "IRF8", "CD28", "CD5", "CD3D", "CD3E", "CD8A", "CD8B", "FOXP3", "E2F1", "E2F2", "KLRK1", "KLRB1", "KIT", "NCR1", "TBX21", "EOMES", "GATA3", "IL1RL1", "IL7R", "ID3", "ACTG1", "GPR37", "NAPSA", "SFTPA1", "RPS3A", "RPS26", "RPS11", "RPL37") -->

<!--     defining_genes_external <- c( -->
<!--       "CD2", "CD3E", "CD4", "CD8A", "CCR7", "GZMH", "NCR1", "NKG7", "SOX13", "CD69", -->
<!--       "CXCR3", "CXCR6", "KCNQ5", "LRRC43", "RUNX2", "ST6GALNAC3", "CD19", "CD40", "CXCR5", -->
<!--       "PAX5", "SLA-DQB1", "TBX21", "DUT", "IRF4", "JCHAIN", "PRDM1", "XBP1", "BATF3", -->
<!--       "CXCR4", "FLT3", "ID2", "IDO1", "IL4I1", "IRAK2", "LY75", "NAV1", "C2", "C3", "CD93", -->
<!--       "IRF8", "NRP1", "TNF", "APOE", "ARG1", "CCL2", "CCL8", "CCR2", "CD14", "CD163", -->
<!--       "CX3CR1", "FCN1", "IL1B", "IL4R", "LAMP1", "TNFRSF1B", "MSR1", "PPARG", "SIGLEC1", -->
<!--       "CD24", "CXCL8", "IL18", "S100A8", "FCER1A", "GATA2", "HDC", "KIT", "LTC4S", "MS4A2", -->
<!--       "ACTA2", "ACTG2", "CAV1", "CAVIN2", "CCDC80", "CCL11", "CD34", "CD9", "CNN1", -->
<!--       "COL3A1", "MYH11", "PECAM1", "TIMP3") -->
<!--     added_markers  <- !(defining_genes_185 %in% defining_genes_external) -->
<!--     defining_genes_final <- c(defining_genes_external, defining_genes_185[added_markers]) -->
<!--     length(defining_genes_final) -->


<!--     n_cluster <-  length(levels(sce_run_x_corrected_clust)) -->
<!--     selected_markers_flag <- (rowData(sce)$Symbol %in% defining_genes_final) -->
<!--     # table(selected_markers_flag) -->
<!--     selected_markers <- as.data.frame(rowData(sce)[selected_markers_flag , ]) -->


<!--     print('maker genes not found in sce:') -->
<!--     print(defining_genes_final[!(defining_genes_final %in% selected_markers$Symbol)]) -->
<!--     # length(unique(selected_markers$Symbol)) -->

<!--     defining_genes_final <-  defining_genes_final[(defining_genes_final %in% selected_markers$Symbol)] -->


<!--     sce_selected_markers <- sce -->
<!--     rownames(sce_selected_markers) <- rowData(sce)$Symbol -->
<!--     rowData(sce_selected_markers) <- rowData(sce_selected_markers)[, c("ID", "Symbol")] -->
<!--     #subset to defining genes and calculate mean logcounts -->
<!--     sce_selected_markers <- sce_selected_markers[rownames(sce_selected_markers) %in% defining_genes_final, ] -->
<!--     selected_markers_matrix_average_expression <- as.data.frame(logcounts(aggregateAcrossCells(sce_selected_markers, ids=sce_selected_markers$cluster, statistics='mean', use.assay.type='logcounts'))) -->
<!--     selected_markers_matrix_average_expression$gene_symbol <- rownames(selected_markers_matrix_average_expression) -->

<!--     library(tidyr) -->
<!--     average_expression_df <- gather(selected_markers_matrix_average_expression, Cluster,  AverageExpression,  -->
<!--                         1:n_cluster ) -->


<!--     selected_markers_matrix_percent_expressed <- as.data.frame(logcounts(aggregateAcrossCells(sce_selected_markers, ids=sce_selected_markers$cluster, statistics="prop.detected", use.assay.type='logcounts'))) -->
<!--     selected_markers_matrix_percent_expressed$gene_symbol <- rownames(selected_markers_matrix_percent_expressed) -->

<!--     percent_expressed_df <- gather(selected_markers_matrix_percent_expressed, Cluster,  PercentExpressed,  -->
<!--                                     1:n_cluster ) -->

<!--     # Merge both data frames by Gene and Cluster -->
<!--     plot_data <- merge(average_expression_df, percent_expressed_df, by = c("gene_symbol", "Cluster")) -->
<!--     plot_data$Cluster <- factor(plot_data$Cluster, levels = sort(as.numeric(unique(plot_data$Cluster)))) -->

<!--     plot_data$gene_symbol <- factor(plot_data$gene_symbol, levels = rev(defining_genes_final)) -->


<!--     png(filename='markers_clusters_heatmap.png', height=1500, width=1000) -->
<!--     # Draw the dot plot using ggplot2 -->
<!--     ggplot(plot_data, aes(x = Cluster, y = gene_symbol)) + -->
<!--       geom_point(aes(size = PercentExpressed, color = AverageExpression)) + -->
<!--       scale_size_continuous(range = c(0.5, 6)) + # Adjust the size range as needed -->
<!--       scale_color_gradient(low = "white", high = "red") + # Set gradient colors for average expression -->
<!--       labs( -->
<!--         x = "Cluster", -->
<!--         y = "gene_symbol", -->
<!--         color = "Average Expression", -->
<!--         size = "Percent Expressed" -->
<!--       ) + -->
<!--       theme_minimal() + -->
<!--       theme( -->
<!--         axis.text.x = element_text(angle = 45, hjust = 1), -->
<!--         panel.grid.major = element_blank(),  # Remove major grid lines -->
<!--         panel.grid.minor = element_blank()   # Remove minor grid lines -->
<!--       ) -->
<!--     dev.off() -->

<!-- } -->




<!-- ``` -->



# save workspace image

```{r save workspace image after first time clustering}

# save the R variables


# save.image(file = paste0(sampSet, "_k", clustK, selected_main_group, "_initial_cluster.RData"))

saveRDS(sce_main_group_corrected, file = paste0('sce_',sampSet, '_', selected_main_group,'_corrected_initial_cluster.rds'))
saveRDS(sce_main_group, file = paste0('sce_',sampSet, '_', selected_main_group,'_initial_cluster.rds'))


```


```{r save workspace image and create file for r shiny visulization}
# save the R variables
setwd(work_dir)

# save.image(file = paste0(sampSet, "_", "k", clustK, "_with_subcluster.RData"))

# saveRDS(sce_corrected, file = paste0('sce_',sampSet,'_corrected_with_subcluster.rds'))
# saveRDS(sce, file = paste0('sce_',sampSet,'_with_subcluster.rds'))
sessionInfo()

```
