---
title: "clustering and sub-clustering within each subset"
author: "lily yang"
date: "`r Sys.Date()`"
output: html_document
---
# set up
```{r, setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

work_dir <- Sys.getenv("work_dir")
bsp_num <- Sys.getenv("bsp_num")
sampSet <- Sys.getenv("sampSet")
clustK <- Sys.getenv("clustK")
reference_genome_version <- Sys.getenv("reference_genome_version")

knitr::opts_knit$set(root.dir = work_dir)

if (clustK == "") {
   clustK <- 20          
   # default
}


print(paste('work_dir', work_dir, 'sampSet', sampSet, 'bsp_num', bsp_num, 'clustK', clustK))              
pfx <- paste0(work_dir,"Results/")
dir.create(pfx, showWarnings = FALSE)

library(SingleCellExperiment)
library(DropletUtils)
library(scran)
library(scater)
library(uwot)
library(biomaRt)
library(scds)
library(batchelor)
library(ggplot2)
library(ggbeeswarm)
library(bluster)
library(scran)
library(scater)


```

This script picks up where initial clustering step left off.

We will load the data, divide cells into different subset, re-cluster, remove doublet clusters, evaluate cluster quality, etc, in order to come up with the best dataset for annotation.

This script is usually run on the hpc, see "code/3_html/muci_html_cluster_subset.sh".


# initial Clustering and evaluation

## Reading in the data


Load the data.

```{r load processed data produced at end of initial clustering script, subset selected cells, overview cell number per sample, fig.width=10}

sce_all <- readRDS("sce_all_with_named_subcluster.rds")

# ensure sample 20 has been deleted 
# read in a table record which initial clusters belong to which subset, extract the clusters id for one subset
library(readxl)
population_divide <- read_excel("Population devision.xlsx")
# selected_main_group <-  as.character(population_divide[1, 1])
selected_main_group <-  sampSet
population_divide <- as.data.frame(population_divide)
rownames(population_divide) <-  population_divide[,1]
selected_cluster <- as.character(population_divide[selected_main_group, 3])
selected_cluster <- strsplit(selected_cluster, split = ",")[[1]]

library(stringr)
selected_cluster <- str_replace_all(selected_cluster, fixed(" "), "")
# save.image('cluster_subset_uniform_debug.RData')
# load('cluster_subset_uniform_debug.RData')
print(selected_cluster)

# extract cells belong to selected initial clusters
sce_main_group <- sce_all[ , (sce_all$cluster %in% selected_cluster) ]
table(sce_main_group$Sample)
overview_cell_number_per_sample <- function(sce, flag)
{
  library(ggplot2)
  df <- as.data.frame(table(sce$Sample))
  colnames(df) <- c("Sample_id","n_cells")


  dat_2 <- data.frame(Sample_id = sce$Sample,
                      cull_date = sce$cull_date,
                      condition = sce$condition)
  dat_2 <- unique(dat_2)
  dat  <- base:::merge(df, dat_2, all.x = TRUE)

  ggplot(data=dat, aes(x=Sample_id, y=n_cells,fill = Sample_id)) +
    geom_bar(stat="identity") +
    geom_text(aes(label=n_cells), vjust=1, size=3.5) +
    ggtitle(paste0(flag, ", cell number summary")) +
    facet_wrap(cull_date ~ condition,  scales="free_x") +
    theme(axis.text.x = element_text(angle = 90))

}

# my_bar_plot(sce_run_x,"before batch effect correction")
overview_cell_number_per_sample(sce_main_group, flag = paste('main group:', selected_main_group))

```

```{r re-normarlize sce subset, batch effect correction, calculate PCA and UMAP, fig.width=10}


clust <- quickCluster(sce_main_group)
size_factors <- calculateSumFactors(sce_main_group, cluster=clust)
# log transfrom gene counts
sce_main_group <- logNormCounts(sce_main_group, size_factors=size_factors)
varmod <- modelGeneVar(sce_main_group, block=sce_main_group$run)
# Model the variance of the log-expression profiles for each gene, decomposing it into technical and biological components based on a fitted mean-variance trend.
vargenes <- getTopHVGs(varmod, n=5000)

sce_main_group <- fixedPCA(sce_main_group, subset.row=vargenes)

set.seed(100)
sce_main_group <- runUMAP(sce_main_group, dimred='PCA')
sce_main_group <- runTSNE(sce_main_group, dimred='PCA')
library(batchelor)
set.seed(100)
sce_main_group_corrected <- fastMNN(sce_main_group, batch=sce_main_group$run, subset.row = NULL, get.variance = TRUE)
sce_main_group_corrected$Sample <- sce_main_group$Sample
sce_main_group_corrected$condition <- sce_main_group$condition
sce_main_group_corrected$run <- sce_main_group$run
sce_main_group_corrected$cull_date <- sce_main_group$cull_date
sce_main_group_corrected$animal_id <- sce_main_group$animal_id
sce_main_group_corrected$cxds_score <- sce_main_group$cxds_score
sce_main_group_corrected$query_doublet <- sce_main_group$query_doublet
sce_main_group_corrected$sum_counts <- sce_main_group$sum_counts
sce_main_group_corrected$genes_detected <- sce_main_group$genes_detected
sce_main_group_corrected$mito_proportion <- sce_main_group$mito_proportion
sce_main_group_corrected$sum_counts_level <- sce_main_group$sum_counts_level

rowdat <- rowData(sce_main_group_corrected)
rowdat <- cbind(rowData(sce_main_group)[rownames(sce_main_group_corrected),], rowdat )
rowData(sce_main_group_corrected) <- rowdat
set.seed(100)
sce_main_group_corrected <- runUMAP(sce_main_group_corrected, dimred="corrected")
sce_main_group_corrected <- runTSNE(sce_main_group_corrected, dimred="corrected")


```


## Clustering the data

Now we re cluster the data, look at sample contribution to each cluster, and find marker genes for each cluster, which might be used in cell type annotation step.

We'll use the graph-based clustering implemented in bluster, specifically using jaccard index to weight edges and the louvain method for community detection, with a k value of `r clustK`. This is largely because this combination gave us a reasonable number of clusters with decent quality metrics after trying a variety of different options. Thus we will run this on the batch-corrected data and look at distribution of samples across clusters and top marker genes for each cluster.


### Cluster, check and save DE genes


```{r re-cluster subset}
library(bluster)
library(scran)
library(scater)
# scran library need to load after some packages, otherwise could not find function "clusterCells"

sce_corrected_clust <- clusterCells(sce_main_group_corrected,
                                          use.dimred="corrected",
                                          BLUSPARAM=SNNGraphParam(k=clustK, type="jaccard",cluster.fun="louvain"))

# k An integer scalar specifying the number of nearest neighbors to consider during graph construction.



table(sce_corrected_clust)
sce_main_group_corrected$cluster <- sce_corrected_clust
sce_main_group$cluster <- sce_corrected_clust


```


### Basic dim reduction plots
```{r create basic UMAP plots, fig.width=10, fig.height=7 }
UMAP_color_by_cluster <- function(selected_sce, title_flag = "batch effect corrected" )
{
  set.seed(55698)
  umap_plot <- plotUMAP(selected_sce, colour_by="cluster", point_size=0.3,
                        text_by = "cluster") +
    guides(colour = guide_legend(override.aes = list(size=6))) +
    ggtitle(paste(sampSet, title_flag))
  # Adjust the legend text and title size


  # Print the plot
  print(umap_plot)


  TSNE_plot <- plotTSNE(selected_sce, colour_by="cluster", point_size=0.3,
                        text_by = "cluster") + guides(colour = guide_legend(override.aes = list(size=6))) + ggtitle(paste(sampSet, title_flag))
  print(TSNE_plot)
}

UMAP_color_by_cluster(sce_main_group_corrected, "batch effect corrected" )
UMAP_color_by_cluster(sce_main_group, "original, no batch effect correction")

ReducedDim_plot <-  plotReducedDim(sce_main_group_corrected, dimred='corrected',
                                   colour_by="cluster", point_size=0.1, ncomponents=5) +
  guides(colour = guide_legend(override.aes = list(size=6)))
print(ReducedDim_plot)


```

```{r draw bar plot to check sample/condition contribution within each cluster, fig.width=10, fig.height=7}
assays(sce_main_group_corrected)$counts <- assays(sce_main_group)$counts[rownames(sce_main_group_corrected),]
assays(sce_main_group_corrected)$logcounts <- assays(sce_main_group)$logcounts[rownames(sce_main_group_corrected),]


sce_corrected <- sce_main_group_corrected

# plot the sample composition of each cluster in a bar plot
dat <- as.data.frame(table(sce_corrected$cluster, sce_corrected$Sample))
colnames(dat) <- c('cluster', 'Sample_id', 'n_cells')
dat_2 <- data.frame(Sample_id = sce_corrected$Sample,
                    cull_date = sce_corrected$cull_date,
                    condition = sce_corrected$condition)
dat_2 <- unique(dat_2)
dat  <- base:::merge(dat, dat_2, all.x = TRUE)




library(ggpubr)
library(ggrepel)
p <- ggplot(dat, aes(fill=Sample_id, y=n_cells, x=cluster)) +
  geom_bar(position='stack', stat='identity') +
  facet_wrap(cull_date ~ condition,  scales="free_x") +
  labs(title=paste0("cluster compostion ",sampSet,"_k",clustK))

print(p)

png(filename = paste0('Results/clustering_', sampSet, '_', selected_main_group, '_cluster_composition_barplot.png') ,
    height=1800, width=2600, res = 300)
print(p)
dev.off()

# draw a bar plot to visulize the composition of each virus within each cluster
# colour cluster plot by condition
ggplot(dat, aes(fill=condition, y=n_cells, x=cluster)) +
  geom_bar(position='stack', stat='identity') +
  facet_wrap( ~ cull_date,  scales="free_x") +
  labs(title=paste0("cluster compostion ",sampSet,"_k",clustK))


```

```{r randomly pick one point among dots whose UMAP location is identical and draw UMAP, fig.width=10, fig.height=7}
library(dplyr)

library(scater)
selected_sce <- sce_corrected

draw_UMAP_after_sampling <- function(selected_sce,
                                     title_flag = "batch effect corrected")
{

  # Get UMAP coordinates and sample information

  umap_coords <- as.data.frame(reducedDim(selected_sce, "UMAP"))

  umap_coords$Sample <- selected_sce$Sample
  umap_coords$condition <- selected_sce$condition
  umap_coords$sum_counts_level <- selected_sce$sum_counts_level
  umap_coords$run <- selected_sce$run
  umap_coords$cull_date <- selected_sce$cull_date
  head(selected_sce$run)

  set.seed(100)
  # Group by UMAP coordinates and sample one point from each group
  umap_sampled <- umap_coords %>%
    group_by(UMAP1, UMAP2) %>%
    sample_n(1)

  # Plot the UMAP with sampled points
  umap_plot <- ggplot(umap_sampled, aes(x = UMAP1, y = UMAP2, color = Sample)) +
    geom_point(size = 0.1) +
    ggtitle(paste(sampSet, title_flag)) +
    facet_wrap(cull_date ~ condition) +
    guides(colour = guide_legend(override.aes = list(size = 6)))

  # Print the plot
  print(umap_plot)


  # Plot the UMAP with sampled points
  umap_plot <- ggplot(umap_sampled, aes(x = UMAP1, y = UMAP2, color = condition)) +
    geom_point(size = 0.1) +
    facet_wrap( ~ cull_date) +
    ggtitle(paste(sampSet, title_flag)) +
    guides(colour = guide_legend(override.aes = list(size = 6)))

  # Print the plot
  print(umap_plot)

  umap_plot <- ggplot(umap_sampled, aes(x = UMAP1, y = UMAP2, color = sum_counts_level)) +
    geom_point(size = 0.1) +
    ggtitle(paste(sampSet, title_flag)) +
    facet_wrap(cull_date ~ condition) +
    guides(colour = guide_legend(override.aes = list(size = 6)))

  # Print the plot
  print(umap_plot)

  umap_plot <- ggplot(umap_sampled, aes(x = UMAP1, y = UMAP2, color = run)) +
    geom_point(size = 0.1) +
    ggtitle(paste(sampSet, title_flag)) +
    guides(colour = guide_legend(override.aes = list(size = 6)))

  # Print the plot
  print(umap_plot)
}
draw_UMAP_after_sampling(sce_corrected, "batch effect corrected" )


```



## Evaluating the clustering quality

We will now run several metrics to evaluate how well our cells are clustered, thinking about whether they have been either over- or under-clustered.

### check silhouette widths:
First we'll look at the silhouette widths, which incorporates both the distances to other cells in the same cluster as well as the distances to cells in other clusters. The plot we make will colour cells according to their nearest cluster if they have a negative silhouette width.

```{r calculate silhouette widths and visilize it}
library("ggbeeswarm")

sil.approx <- approxSilhouette(reducedDim(sce_corrected, "corrected"), clusters=sce_corrected$cluster)
sil.approx

sil.data <- as.data.frame(sil.approx)
sil.data$closest <- factor(ifelse(sil.data$width > 0, sce_corrected$cluster, sil.data$other))
sil.data$cluster <- sce_corrected$cluster
scTkn <- paste0(sampSet,"_k",clustK)
ggplot(sil.data, aes(x=cluster, y=width, colour=closest)) +
  ggbeeswarm::geom_quasirandom(method="smiley") +
  labs(title=scTkn)

```

### check cluster purity:
To better separate the factors that contribute to this metric, we'll divide out purity and heterogeneity, first looking at cluster purity. Here again cells are coloured according to their nearest cluster if their purity is particularly low.

clusters which have low purity, indicate some of cell within these clusters have gene expression profile poorly aligns with the centroid or the majority of cells in a cluster.
```{r calculate cells purity and visualization in UMAP}


###########################################################################

pure <- neighborPurity(reducedDim(sce_corrected, "corrected"), sce_corrected$cluster)
pure
pure.data <- as.data.frame(pure)
pure.data$maximum <- factor(pure.data$maximum)
pure.data$cluster <- sce_corrected$cluster

ggplot(pure.data, aes(x=cluster, y=purity, colour=maximum)) +
  ggbeeswarm::geom_quasirandom(method="smiley") +
  labs(title=scTkn)
###########################################################################

# check the UMAP location of cells have low purity values
colData(selected_sce)$purity <- pure.data[colnames(selected_sce), 'purity']
location_purity_UMAP <- function(selected_sce,cluster_n = 8)
{
  p <-  plotUMAP(selected_sce[,selected_sce$cluster==cluster_n], colour_by="purity", point_size=1) + ggtitle(paste('cluster',cluster_n))
  # plotUMAP(selected_sce, colour_by="cluster", point_size=1)
  print(p)
}

# visulize selected unpure clusters, color their cells by purity values to see the distribution features of low purity cells

# location_purity_UMAP(selected_sce,2)
# table(Cluster=sce_corrected$cluster, pure.data$maximum)

df <- table(Cluster=sce_corrected$cluster, pure.data$maximum)

# Extract diagonal elements
diagonal_values <- diag(as.matrix(df))
# Calculate row sums
row_sums <- rowSums(df)

# Subtract diagonal values from row sums
adjusted_sums <- (row_sums - diagonal_values)
summary_purity_df <- data.frame(misclassifed_counts = adjusted_sums,
                                misclassifed_proportion = (adjusted_sums/diagonal_values))



# df$misclassifed_counts <- adjusted_sums
# df$misclassifed_proportion <- adjusted_sums/diagonal_values
print(df)
print(summary_purity_df)

num_cells_misclassifed <- sum(summary_purity_df$misclassifed_counts)
n_cells <-  dim(selected_sce)[2]
print(paste(num_cells_misclassifed, num_cells_misclassifed/n_cells))





```


from the table and UMAP, we can see the cells with low purity values always locate at the interface area with adjacent clusters

in UMAP plot, we can find some clusters locate quiet close to other clusters, some cells classified as one cluster actually locate closer to adjacent clusters. which indicate some cells might get mis-classified.


### check heterogeneity within each cluster:
And finally we'll look at the heterogeneity within each cluster by computing the within cluster sum of squares.

```{r check heterogeneity within each cluster and visualization}
###########################################################################
rmsd <- clusterRMSD(reducedDim(sce_corrected, "corrected"), sce_corrected$cluster)

df <- as.data.frame(rmsd)
df$cluster <- as.factor(1:length(levels(sce_corrected$cluster)))


# Basic barplot
ggplot(data=df, aes(x=cluster, y=rmsd)) +
  geom_bar(stat="identity")+
  geom_text(aes(label=cluster), vjust=-0.3, size=3.5) +
  ggtitle("examine heterogeneity within each cluster")

```



### check QC metrics of all clusters
check QC metrics for all clusters (sum_counts, genes_detected, mito_proportion).
check sample contribution of suspicious cluster, sample distribution in UMAP of suspicious cluster



```{r check QC metrics for all clusters and sample contribution - sample-condition distribution in UMAP of suspicious cluster, fig.width=10}
QC.tab <- data.frame(colData(sce_corrected)[,c('cluster', 'sum_counts', 'mito_proportion', 'genes_detected')])

p <-  ggplot(QC.tab, aes(x=cluster, y=log2(sum_counts) )) +
  ggbeeswarm::geom_quasirandom(method="smiley")
print(p)

p <-  ggplot(QC.tab, aes(x=cluster, y=log2(genes_detected) )) +
  ggbeeswarm::geom_quasirandom(method="smiley")
print(p)

p <-  ggplot(QC.tab, aes(x=cluster, y=mito_proportion)) +
  ggbeeswarm::geom_quasirandom(method="smiley")
print(p)


```

## Doublet inspection of all clusters

It may be the case that our suspected doublets, which fell in a particular region of the UMAP, have formed their own cluster. This is a convenient means of removing them as we can get rid of the whole cluster and don't have to rationalist a threshold for suspected doublet value.

```{r check Doublet distribution across clusters, fig.width=10}
# print the proportion of estimated Doublet in each cluster
prop.table(table(sce_corrected$cluster, sce_corrected$query_doublet), margin=1)

violin_plot_doublet_inspection <- function(sce_corrected,
                                      title_flag = "before filtering estimated Doublet")
{
  # visualize cxds_score and estimated Doublet in each cluster
  doublet.tab <- data.frame(colData(sce_corrected)[,c('cxds_score', 'cluster', 'query_doublet', 'sum_counts')])

  p1 <- ggplot(doublet.tab, aes(x=cluster, y=cxds_score, colour=query_doublet)) +
      ggbeeswarm::geom_quasirandom(method="smiley") +
    ggtitle(paste("doublet inspection",selected_main_group,  title_flag))
  print(p1)

  p1 <- ggplot(doublet.tab, aes(x=cluster, y=sum_counts, colour=query_doublet)) +
      ggbeeswarm::geom_quasirandom(method="smiley")  +
    ggtitle(paste("doublet inspection", selected_main_group, title_flag))
  print(p1)
}

violin_plot_doublet_inspection(sce_corrected)

sce_corrected <- sce_corrected[ , !sce_corrected$query_doublet]

violin_plot_doublet_inspection(sce_corrected)
```


# save workspace image after initial clustering

```{r save workspace image after initial clustering}

# save the R variables


# save.image(file = paste0('subset', selected_main_group, "_k", clustK, "_initial_cluster.RData"))

saveRDS(sce_main_group_corrected, file = paste0('sce_', 'subset_', selected_main_group, '_corrected_initial_cluster.rds'))
saveRDS(sce_main_group, file = paste0('sce_', 'subset_', selected_main_group, '_initial_cluster.rds'))


```

# re-clustering T cells and NK cells if necessary 
```{r create a function to visulize new generated clusters in this step}

# check the location of selected new generated clusters in UMAP

check_sample_contribution_location_for_selected_subcluster <- function(selected_sce,cluster_n = c('8a','8b','8c'))  
{
# check the sample contribution for cluster 
the_cluster <- prop.table(table(selected_sce$cluster, selected_sce$Sample), margin=1)[cluster_n,]


the_cluster <- as.data.frame(the_cluster)
colnames(the_cluster) <- c('sub_cluster','sample','contribution' )


p <-   ggplot(data=the_cluster, aes(x = sub_cluster , y=contribution, fill=sample)) +
  geom_bar(stat="identity")  + ggtitle('sample contribution in sub_cluster')

print(p) 

p <-  plotUMAP(selected_sce[,selected_sce$cluster %in% cluster_n], colour_by="Sample", point_size=1) + ggtitle('sub cluster')
print(p)

p <-  plotUMAP(selected_sce[,selected_sce$cluster %in% cluster_n], colour_by="condition", point_size=1) + ggtitle('sub cluster')
print(p)

p <- plotUMAP(selected_sce[,selected_sce$cluster %in% cluster_n], colour_by="cluster", point_size=1) + ggtitle('sub cluster')
p <- p + guides(colour = guide_legend(override.aes = list(size=6)))
print(p)  
  
}

```


```{r unguided unsupervised re-cluster, fig.width=10, fig.height=7 }


unguided_unsupervised_recluster <- function(sce, cluster_n = '12',n_HVG = 2000, k_values = 20)
{
  # extract all cells in selected clusters, and get the top 1000 (n_HVG) HVG in these cells
  selected_cells <- (sce$cluster %in% cluster_n)
  print(table(selected_cells))
  
  varmod8 <- modelGeneVar(sce[,selected_cells], block=sce[,selected_cells]$run)
  vargenes8 <- getTopHVGs(varmod8, n=n_HVG)
  vargenes8 <- union(vargenes8, c('ENSSSCG00000000687', 'ENSSSCG00000008213',
                                  "ENSSSCG00000039708", 'ENSSSCG00000040465' , 'ENSSSCG00000040140'))
  library(batchelor)
  set.seed(100)
  sce_8_fastMNN <- fastMNN(sce[,selected_cells], batch=sce[,selected_cells]$run,
                           subset.row = vargenes8, get.variance = TRUE)
  sce_8_fastMNN <- runUMAP(sce_8_fastMNN , dimred="corrected")  
  
  library(bluster)
  library(scran)
  library(scater)    
  selected_k_value <-  k_values
  ########################################################################### 
  
  
  
  set.seed(200)
  sce_8_fastMNN_clust <- scran::clusterCells(sce_8_fastMNN, use.dimred="corrected",
                                             BLUSPARAM=SNNGraphParam(k = selected_k_value, 
                                                                     type="jaccard",
                                                                     cluster.fun="fast_greedy"))
  # change the clustering algorithm to KmeansParam
  # sce_8_fastMNN_clust <- scran::clusterCells(sce_8_fastMNN, use.dimred="corrected",
  #                                     BLUSPARAM=KmeansParam(centers = 6))
  
  print(paste('n_HVG:', n_HVG, 'selected_k_value:', selected_k_value ))
  
  print(table(sce_8_fastMNN_clust))
  sce_8_fastMNN$cluster <- sce_8_fastMNN_clust
  p <- plotUMAP(sce_8_fastMNN, colour_by="cluster", point_size=1) + 
    ggtitle(paste(n_HVG, 'HVG',  'k', selected_k_value ))
  p <- p + guides(colour = guide_legend(override.aes = list(size=6)))
  print(p)  
  
  # add necessary log-counts and gene id, symbol metadata into sce_8_fastMNN
  assays(sce_8_fastMNN)$logcounts <- assays(sce)$logcounts[rownames(sce_8_fastMNN), selected_cells]
  
  rowdat <- rowData(sce_8_fastMNN)
  head(rowData(sce))
  rowdat <- cbind( rowData(sce)[rownames(sce_8_fastMNN), ], rowdat )
  rowData(sce_8_fastMNN) <- rowdat      
  
  # draw the UMAP plots to check the selected genes expression in UMAP 
  for(selected_gene in c("CD4", "CD8B",  "KLRB1", "KLRK1", "CD3E"))
  {
    
    selected_gene_id <- rownames(sce)[rowData(sce)$Symbol == selected_gene]
    print(paste(selected_gene, selected_gene_id))
    p <- plotUMAP(sce_8_fastMNN, colour_by=selected_gene_id, by_exprs_values='logcounts') + 
      ggtitle(paste(selected_gene, n_HVG, 'HVG',  'k', selected_k_value )) 
    print(p)
  }
  
  # draw the violin plots to check the selected genes expression across new clusters 
  # selected_genes <- rownames(sce)[rowData(sce)$Symbol %in% c("CD4", "CD8B",  "KLRB1", "KLRK1", "CD3E")]
  selected_genes <- c("CD4", "CD8B",  "KLRB1", "KLRK1", "CD3E")
  library(scater)
  p <- plotExpression(sce_8_fastMNN, features=selected_genes, x='cluster',
                      exprs_values='logcounts', swap_rownames = 'Symbol' )   
  print(p)
  
  
  ########################################################################### 
  
  # Ensure sce_8_fastMNN_clust is a character vector
  sce_8_fastMNN_clust <- as.character(sce_8_fastMNN_clust)
  
  # Loop through the numbers and assign new labels
  subcluster_vector <- NULL
  for (i in unique(sce_8_fastMNN_clust) ) {
    new_cluster_lable <- paste0(i, "re")
    sce_8_fastMNN_clust[sce_8_fastMNN_clust == as.character(i)] <- new_cluster_lable
    subcluster_vector <- c(subcluster_vector, new_cluster_lable)
  }
  # This will assign '1' with '1re', '2' with '2re', ..., '5' with '5re', ...
  

  
  c12 <- (sce$cluster %in% cluster_n)
  cs <- as.character(sce$cluster)
  cs[c12] <- sce_8_fastMNN_clust
  sce$cluster <- cs
  check_sample_contribution_location_for_selected_subcluster(sce, subcluster_vector)
  
  
  return(sce)
}


if(bsp_num == 'bsp_248' & sampSet == "adaptive_immune_cells")
{
  cluster_n <- c("1", "3", "4", "5", "6", "7", "8")
  selected_k_values <- 10

  sce_main_group <- unguided_unsupervised_recluster(sce_main_group, cluster_n, n_HVG = 4000, k_value = selected_k_values)
}
sce_main_group_corrected$cluster <- sce_main_group$cluster

UMAP_color_by_cluster(sce_main_group_corrected, "batch effect corrected after re-clustering T cells, NK cells" )

```

# output DE genes within each clusters and save image
Finally, we need to re-run our marker scoring because we have added some clusters. 


```{r save top DE genes in excel file for new produced sub cluster}

selected_main_group <- gsub(' ', '_', selected_main_group)
assays(sce_main_group_corrected)$counts <- assays(sce_main_group)$counts[rownames(sce_main_group_corrected),]
assays(sce_main_group_corrected)$logcounts <- assays(sce_main_group)$logcounts[rownames(sce_main_group_corrected),]

sce_corrected <- sce_main_group_corrected

########################################################################### 
cell_n_cluster <- table(sce_corrected$cluster)


marker.info.corrected <- scoreMarkers(sce_corrected,
                                      groups=sce_corrected$cluster,
                                      block=sce_corrected$run,
                                      assay.type='logcounts',
                                      full.stats=TRUE)



library(openxlsx)
# Create a new workbook
wb <- createWorkbook()

for(i in unique(names(marker.info.corrected)) )
{
  # print(i)
  ord <- marker.info.corrected[[i]][order(marker.info.corrected[[i]]$mean.logFC.cohen, decreasing=TRUE),]

  if(bsp_num == "bsp_248")
  {
    if(reference_genome_version == "pig_108")
    {
      ord$symbol <- rowData(sce_corrected)[rownames(ord), 'ENTREZ_symbol']
    }else if(reference_genome_version == "pig_113")
    {
      ord$symbol <- rowData(sce_corrected)[rownames(ord), 'Symbol']
    }
  }
  if(bsp_num == "bsp_247")
  {
    ord$symbol <- rowData(sce_corrected)[rownames(ord), 'Symbol']
  }


  pfx <- paste0(work_dir,"markers_",sampSet,'_k',clustK)
  dir.create(pfx, showWarnings = FALSE)



  DE_genes <- cbind(ord$symbol,ord)
  DE_genes <- as.data.frame(ord)
  DE_genes <- cbind(rownames(DE_genes), ord$symbol, DE_genes)
  DE_genes_mid <- DE_genes[ ,c(7,1:6)]
  DE_genes_mid$'mean.AUC' <- DE_genes[ ,'mean.AUC']
  DE_genes <- DE_genes_mid
  colnames(DE_genes)[2:3] <- c('ensemble_id', 'symbols')

  sheet_name <- paste("cluster", i, '#',cell_n_cluster[i])

  # Add a worksheet to the workbook with the created name
  addWorksheet(wb, sheetName = sheet_name)
  # Write the data frame to the worksheet
  writeData(wb, sheet = sheet_name, DE_genes)

}
# Save the workbook to a file
saveWorkbook(wb, file = paste0('Results/clustering_', sampSet, '_',
                               selected_main_group, "_marker_per_cluster.xlsx"), overwrite = TRUE)




```




```{r draw marker genes expression heatmap for selected genes across clusters, fig.width= 12, fig.height=8}
# sce_main_group <- readRDS("sce_subset_adaptive_immune_cells_with_named_refined_cluster.rds")

if(bsp_num == "bsp_248" )
{

  library(readxl)
  if(sampSet == "innate_immune_cells")
  {
    df_cell_type_annotation <- read_excel("references/heatmap porcine_innate_clusters_genes_ordered.xlsx")    
   
  }else if(sampSet == "adaptive_immune_cells")
  {
    df_cell_type_annotation <- read_excel("references/heatmap procine adaptive gene ID.xlsx")    
  }else if(sampSet == "non_immune_cells")
  {
    df_cell_type_annotation <- read_excel("references/heatmap porcine Non immune cell gene list.xlsx")    
   
  }
  df_cell_type_annotation <- as.data.frame(df_cell_type_annotation)
  # head(df_cell_type_annotation)

  selected_marker_genes <- df_cell_type_annotation[,1]
  defining_genes_final <- unique(selected_marker_genes) 



# n_cluster <- length(levels(sce_run_x_corrected_clust))
n_cluster <- length(unique(sce_main_group$cluster))  
sce <- sce_main_group

selected_markers_flag <- (rowData(sce)$Symbol %in% defining_genes_final)
# table(selected_markers_flag)
# obtain the gene metadata information for selected genes， and save in data frame
selected_markers <- as.data.frame(rowData(sce)[selected_markers_flag , ])


print('marker genes not found in our data:')
print(defining_genes_final[!(defining_genes_final %in% selected_markers$Symbol)])
# length(unique(selected_markers$Symbol))
# filter the gene list to make sure all genes are measured and can be found in our seq data 
defining_genes_final <-  defining_genes_final[(defining_genes_final %in% selected_markers$Symbol)]

# generate a table contain the average gene expression values per gene per cluster
sce_selected_markers <- sce
rownames(sce_selected_markers) <- rowData(sce)$Symbol
rowData(sce_selected_markers) <- rowData(sce_selected_markers)[, c("ID", "Symbol")]
#subset sce_selected_markers to only store measurements of defining genes and calculate mean logcounts
sce_selected_markers <- sce_selected_markers[rownames(sce_selected_markers) %in% defining_genes_final, ]
# calculate the mean expression value for each gene within each cluster
selected_markers_matrix_average_expression <- as.data.frame(logcounts(aggregateAcrossCells(sce_selected_markers, ids=sce_selected_markers$cluster, statistics='mean', use.assay.type='logcounts')))
selected_markers_matrix_average_expression$gene_symbol <- rownames(selected_markers_matrix_average_expression)
# head(selected_markers_matrix_average_expression)

library(tidyr)
# change the df from short format to long format to facilitate plotting 
average_expression_df <- pivot_longer(data = selected_markers_matrix_average_expression,
                                      cols = !gene_symbol,
                                      names_to = "Cluster",  values_to = "AverageExpression" )

  if(sampSet == "innate_immune_cells")
  {
    df_cell_type_annotation <- read.csv("munual_annotation_innate_immune_cells.csv", header = FALSE)    
  }else if(sampSet == "adaptive_immune_cells")
  {
    df_cell_type_annotation <- read.csv("munual_annotation_adaptive_immune_cells.csv", header = FALSE)  
  }else if(sampSet == "non_immune_cells")
  {
    df_cell_type_annotation <- read.csv("munual_annotation_non_immune_cells.csv", header = FALSE)    
  }




colnames(df_cell_type_annotation) <- c("Cluster", "clean_cell_type")


df_cell_type_annotation$Cluster <- as.character(df_cell_type_annotation$Cluster)
average_expression_df <- merge(average_expression_df, df_cell_type_annotation, all.x = TRUE)
# head(df_cell_type_annotation)
# head(average_expression_df)

# calculate the proportion of cells expressed certain gene within certain cluster
# and calculate such proportion for each gene in each cluster
selected_markers_matrix_percent_expressed <- as.data.frame(logcounts(aggregateAcrossCells(sce_selected_markers, ids=sce_selected_markers$cluster, statistics="prop.detected", use.assay.type='logcounts')))
selected_markers_matrix_percent_expressed$gene_symbol <- rownames(selected_markers_matrix_percent_expressed)
percent_expressed_df <- pivot_longer(data = selected_markers_matrix_percent_expressed,
                                     cols = !gene_symbol,
                                     names_to = "Cluster",  values_to = "PercentExpressed" )

# percent_expressed_df <- gather(selected_markers_matrix_percent_expressed, Cluster,  PercentExpressed, 
#                                all_of(n_cluster) )

# Merge 2 data frames store the average_expression and percent detected information by Gene and Cluster
plot_data <- merge(average_expression_df, percent_expressed_df, by = c("gene_symbol", "Cluster"))
# head(percent_expressed_df)
# order the gene in y-axis, order cluster in x-axis
# plot_data$Cluster <- factor(plot_data$Cluster, levels = sort(as.numeric(unique(plot_data$Cluster))))
plot_data$gene_symbol <- factor(plot_data$gene_symbol, levels = rev(defining_genes_final))
table(defining_genes_final)


library(stringr)
library(plyr)
library(dplyr)

var_width = 10
plot_data$clean_cell_type = str_wrap(plot_data$clean_cell_type, width = var_width)


# generate a vector store color of each cell type used for plot
library(dplyr)

celltype_map <- plot_data %>%
  distinct(Cluster, clean_cell_type)

n_cell_types <- length(unique(celltype_map$clean_cell_type))

celltype_colors <- setNames(
  colorRampPalette(RColorBrewer::brewer.pal(12, "Set3"))(n_cell_types),
  unique(celltype_map$clean_cell_type))

# Add cell type annotation bar under x-axis
# Use geom_tile with y positioned just below genes (e.g., y = -1 or a new 'CellType' row)
celltype_map$y <- -1  # Place the annotation below the lowest gene row

################################################# 
# desired_cell_type_order <- read_excel("references/desired_cell_type_order.xlsx", col_names = FALSE)

desired_cell_type_order <- unique(sce_main_group$named_cluster)
desired_gene_order <- defining_genes_final


# Specify desired order as a character or integer vector
# desired_cell_type_order <- as.data.frame(desired_cell_type_order)[,1] 
# desired_cell_type_order <- str_wrap(desired_cell_type_order, width = var_width)
# 
# plot_data$clean_cell_type <- factor(plot_data$clean_cell_type, levels = desired_cell_type_order)
# celltype_map$clean_cell_type <- factor(celltype_map$clean_cell_type, levels = desired_cell_type_order)
# 
# desired_gene_order <- read_excel("references/desired_gene_order_bsp293 WG 25.07.19.xlsx", col_names = FALSE)
# # Specify desired order as a character or integer vector
# desired_gene_order <- as.data.frame(desired_gene_order)[,1] 

# length(unique(plot_data$gene_symbol))


plot_data = plot_data[ !is.na(plot_data$gene_symbol), ]

if(sampSet == "subset_adaptive_immune_cells")
{
  library(dplyr)
  plot_data <- plot_data %>%
     dplyr::filter(!grepl("remove", clean_cell_type))
  desired_cell_type_order <- desired_cell_type_order[!grepl("remove", desired_cell_type_order)]  
  celltype_map <- celltype_map[ !grepl("remove", celltype_map$clean_cell_type), ]
  
}
plot_data$gene_symbol <- factor(plot_data$gene_symbol, levels = desired_gene_order)

head(plot_data)
# desired_gene_order[! desired_gene_order %in% unique(plot_data$gene_symbol)]
########################################################################### 
# Draw the dot plot using ggplot2
p  <- ggplot(plot_data, aes(x = Cluster, y = gene_symbol)) +
  geom_point(aes(size = PercentExpressed, color = AverageExpression)) +
  scale_size_continuous(range = c(0.5, 6)) + # Adjust the size range as needed
  scale_color_gradient(low = "white", high = "red") + # Set gradient colors for average expression
  labs(
    x = "Cluster",
    y = "gene_symbol",
    color = "Average Expression",
    size = "Percent Expressed"
  ) + 
  facet_grid(. ~ clean_cell_type,  scales="free_x") +
  theme_bw()  +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank()   # Remove minor grid lines
  )

  p1 <- p + 
    geom_tile(
      data = celltype_map, 
      aes(x = factor(Cluster), y = y, fill = clean_cell_type), 
      height = 0.5
    ) +
    scale_fill_manual(values = celltype_colors, name = "Cell type") +
    # Expand y axis so the new row is visible
    scale_y_discrete(expand = expansion(add = c(0.2, 1.5))) +
    facet_grid(. ~ clean_cell_type,  scales="free_x") +
    guides(fill = guide_legend(override.aes = list(size = 5)))+
    theme(strip.text.x = element_text(size = 8))


# scale_y_discrete(expand = expansion(add = c(0.2, 1.5))) 
# 0.5 extra units at the bottom (for your annotation bar below genes)
# 1.5 extra units at the top (above your highest gene)
    n_clusters <- length(unique(plot_data$Cluster))
    width_of_plot <- n_clusters * 0.6 + 5

       
  print(p1)
  
  filename1 <- paste0('Results/clustering_markers_per_clusters_', sampSet, '_heatmap.png')
  # png(filename = filename1,
  #     height=1500, width=1500, res = 130)
  # print(p1)
  # dev.off()
  ggsave( filename = filename1,
    plot = p1,
    height = 1500/130, # tranfer pixels into inches
    width  = width_of_plot,
    units  = "in",
    dpi    = 300)


# save.image("cluster_develop.RData")

  
}



```




```{r save workspace image and create file for r shiny visulization}
# save the R variables
setwd(work_dir)

# save.image(file = paste0(sampSet, "_", "k", clustK, "_with_refined_cluster.RData"))


saveRDS(sce_main_group_corrected, file = paste0('sce_subset_',sampSet,'_corrected_with_refined_cluster.rds'))
saveRDS(sce_main_group, file = paste0('sce_subset_',sampSet,'_with_refined_cluster.rds'))
sessionInfo()

```
