---
title: "Differential adundance & expression edgeR Analyses"
author: "lily yang"
date: "`r Sys.Date()`"
output: html_document
---
# set up enrivonment and load data
```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

work_dir <- Sys.getenv("work_dir")
bsp_num <- Sys.getenv("bsp_num")
clustK <- Sys.getenv("clustK")
sampSet <- Sys.getenv("sampSet")
current_run <- sampSet


knitr::opts_knit$set(root.dir = work_dir)


if (clustK == "") {
   clustK <- 20          # defaulty
}
print(clustK)

print(paste('work_dir', work_dir, 'sampSet', sampSet, bsp_num))              
if (sampSet == "") {
   sampSet <- "error, sampSet not provided"

   }
print(sampSet)
pfx <- paste0(work_dir,"Results/")
dir.create(pfx, showWarnings = FALSE)

```


This script is usually run on the hpc, see "lily code/html/html_differential.sh".

```{r load input data }


# sce_all <- readRDS(paste0('sce_',sampSet,'_initial_cluster.rds')) 
# sce_all_corrected <- readRDS(paste0('sce_',sampSet,'_corrected_initial_cluster.rds'))   


# if we refine initial cluster results by conduct sub-clustering or re-clustering
sce_all <- readRDS(paste0('sce_',sampSet,'_with_refined_cluster.rds'))
sce_all_corrected <- readRDS(paste0('sce_',sampSet,'_corrected_with_refined_cluster.rds'))
  
unique(sce_all_corrected$cluster)
unique(sce_all$cluster)
# sce_all_corrected$cluster <- sce_all$cluster

```


This script picks up where "cluster_lily.Rmd" left off. 

We will start by labelling the clusters 
We will then run differential analyses between conditions. 

Specifically, we will be looking for differential abundance of 'cells within each cluster', as well as differential expression of 'cells within each cluster', according to treatment condition. 

There are 2 conditions: virus1 vs virus1 vs control, 2 time point: DPI 1 vs DPI 12



# Loading the packages

```{r load packages}

library(SingleCellExperiment)
library(DropletUtils)
library(edgeR)
library(bluster)
library(scran)
library(scater)
library(uwot)
library(biomaRt)
library(scds)
library(batchelor)
library(gridExtra)
library(ggplot2)
library(ggrepel)

```



# Cluster annotation, remove clusters from downstream analysis 

Based on manual annotation from the clustered data using the generated marker gene tables and our shiny app, we'll give the clusters more intuitive names that can be transposed into figures. As a quick check in this script, we look at particular marker genes first.

```{r double check by look at marker genes expression used in cell sorting}
## first look at marker genes for sorting



# plotExpression(sce_all_corrected, features=rownames(sce_all)[rowData(sce_all)$hgnc_symbol == 'CD4'], x='cluster', exprs_values='logcounts')
plotExpression(sce_all_corrected, features=rownames(sce_all_corrected)[rowData(sce_all_corrected)$Symbol == 'CD4'], x='cluster')

# plotExpression(sce_all_corrected, features=rownames(sce_all)[rowData(sce_all)$Symbol == 'CD8B'], x='cluster', exprs_values='logcounts')
plotExpression(sce_all_corrected, features=rownames(sce_all_corrected)[rowData(sce_all_corrected)$Symbol == 'CD8B'], x='cluster')

# plotExpression(sce_all_corrected, features=rownames(sce_all)[rowData(sce_all)$Symbol == 'CD163'], x='cluster', exprs_values='logcounts')
plotExpression(sce_all_corrected, features=rownames(sce_all_corrected)[rowData(sce_all_corrected)$Symbol == 'CD163'], x='cluster')

# plotExpression(sce_all_corrected, features=rownames(sce_all)[rowData(sce_all)$Symbol == 'CD14'], x='cluster', exprs_values='logcounts')
plotExpression(sce_all_corrected, features=rownames(sce_all_corrected)[rowData(sce_all_corrected)$Symbol == 'CD14'], x='cluster')
 
# plotExpression(sce_all_corrected, features=rownames(sce_all)[rowData(sce_all)$Symbol == 'CD3D'], x='cluster', exprs_values='logcounts')
plotExpression(sce_all_corrected, features=rownames(sce_all_corrected)[rowData(sce_all_corrected)$Symbol == 'CD3D'], x='cluster')

# plotExpression(sce_all_corrected, features=rownames(sce_all)[rowData(sce_all)$Symbol == 'CD3E'], x='cluster', exprs_values='logcounts')
plotExpression(sce_all_corrected, features=rownames(sce_all_corrected)[rowData(sce_all_corrected)$Symbol == 'CD3E'], x='cluster')

```



Note that we are maintaining 2 separate datasets here: 
sce_filt (from sce_all_corrected), which contains all detected genes and reconstructed gene expression values after sequencing run batch correction

sce_all_filt (from sce_all), which contains all detected genes with normalised (but not batch corrected) log-count values and original counts. 

sce_all_filt is important for count-based differential analyses and seeing expression values that aren't interpolated after batch correction.


Next we name the good quality clusters. 

```{r name cell types for each cluster }
sce_filt <- sce_all_corrected
sce_all_filt <- sce_all
unique(sce_all$cluster)

print(sce_all)
# rm(sce_all, sce_all_corrected)

## add in cluster names as sce_all_filt$named_cluster
#Long-form annotated version including cell subtypes
if(sampSet == "subset_innate_immune_cells")
{
  munual_annotation <- read.csv("munual_annotation_innate_immune_cells.csv", header = FALSE)  
}else if(sampSet == "subset_adaptive_immune_cells")
{
  munual_annotation <- read.csv("munual_annotation_adaptive_immune_cells.csv", header = FALSE)
}else if(sampSet == "subset_non_immune_cells")
{
  munual_annotation <- read.csv("munual_annotation_non_immune_cells.csv", header = FALSE)
}else
{
  munual_annotation <- read.csv("munual_annotation.csv", header = FALSE) 
}


colnames(munual_annotation) <- c("cluster", "name")
# if(sampSet == "subset_adaptive_immune_cells")
# { 
#   # remove the content in parentheses as well as the parentheses
#   munual_annotation$name <- sub(" \\(.*\\)", "", munual_annotation$name)
#   
# }

# add an column show all clusters belong to same cell types
library(dplyr)
df_grouped <- munual_annotation %>%
  dplyr::group_by(name) %>%
  dplyr::summarise(cluster_contained = paste(cluster, collapse = ", ")) %>%
  dplyr::ungroup()
df_grouped$celltype_cluster <- paste(df_grouped$name, df_grouped$cluster_contained)

unique(sce_all_filt$cluster)
sce_all_filt$cluster <- as.character(sce_all_filt$cluster)


if(bsp_num == "bsp_248")
{
  for (i in unique(sce_all_filt$cluster))
  {
    cell_type <- munual_annotation[ munual_annotation$cluster == i, "name"]
    long_celltype <- as.character(df_grouped[ df_grouped$name == cell_type, "celltype_cluster"])         
    # print(paste(i, cell_type, long_celltype))
    # print(typeof(long_celltype))
    sce_all_filt$named_cluster[sce_all_filt$cluster==i] <- cell_type
    sce_all_filt$long_named_cluster[sce_all_filt$cluster==i] <- long_celltype
  }
}
table(sce_all_filt$long_named_cluster)
# head(sce_all_filt$long_named_cluster)
# sce_all_filt$cluster <- as.integer(sce_all_filt$cluster)

unique(sce_all_filt$long_named_cluster)


```


```{r manually assign color to the good quality clusters}
# Create label and assign colour for each cluster for convenient annotation later

# label every cell with its tissue type and the cluster it belong to
sce_all_filt$verbose_cluster <- paste0('lung, ', sce_all_filt$named_cluster, ', ', sce_all_filt$cluster)
# label every cell with its tissue type 
sce_all_filt$EXP_celltype <- paste0('lung, ', sce_all_filt$named_cluster)

sce_all_filt

# BAL_colour_list <- vector("list", length(sce_all_filt$EXP_celltype))
# BAL_colour_list[grep('Macrophages', sce_all_filt$EXP_celltype)] <- '#E31A1C' #red
# BAL_colour_list[grep('Monocytes', sce_all_filt$EXP_celltype)] <- '#C9C32F' #pink
# BAL_colour_list[grep('conventional', sce_all_filt$EXP_celltype)] <- '#FB9A99' 
# BAL_colour_list[grep('pDCs', sce_all_filt$EXP_celltype)] <- '#FDDBC7' 
# BAL_colour_list[grep('cDCs', sce_all_filt$EXP_celltype)] <- '#6A3D9A' #purple
# BAL_colour_list[grep('Plasma Cells', sce_all_filt$EXP_celltype)] <- '#1F78B4' #dark blue
# BAL_colour_list[grep('B Cells', sce_all_filt$EXP_celltype)] <- '#A6CEE3' #pale blue
# BAL_colour_list[grep('Tregs', sce_all_filt$EXP_celltype)] <- '#CAB2D6' #light purple
# BAL_colour_list[grep('CD4', sce_all_filt$EXP_celltype)] <- '#B2DF8A' #light green
# BAL_colour_list[grep('Mitotic CD4 T Cells', sce_all_filt$EXP_celltype)] <- '#FFFF00' 
# BAL_colour_list[grep('CD8', sce_all_filt$EXP_celltype)] <- '#33A02C' # green
# BAL_colour_list[grep('Mitotic CD8 T Cells', sce_all_filt$EXP_celltype)] <- '#FF7F00' #orange
# BAL_colour_list[grep('NK', sce_all_filt$EXP_celltype)] <- '#FDBF6F' 
# # unknown clusters are colored with brown
# BAL_colour_list[sapply(BAL_colour_list, is.null)] <- '#B15928' #brown
# sce_all_filt$colour  <- BAL_colour_list

colData(sce_filt)[,c('named_cluster', 'long_named_cluster', 'verbose_cluster', 'EXP_celltype')] <- colData(sce_all_filt)[, c('named_cluster', 'long_named_cluster', 'verbose_cluster', 'EXP_celltype')]

```

And make some UMAPs for visulization.

```{r plot UMAPs to show annotated cell types of clusters, fig.width= 12, fig.height=8}
output_color_code_in_UMAP <- function(x = p1,
                                      clusterID_or_celltype = "named_cluster") 
{
  # Build the ggplot object
  p1_build <- ggplot_build(p1)
  
  # Extract color and label (named_cluster) from the data
  df_colors <- p1_build$data[[1]][, c("colour", "group")]
  
  # Pull named_cluster info from the input data used in plotUMAP
  # Assuming `sce_filt` is a SingleCellExperiment object
  if(clusterID_or_celltype == "named_cluster")
  {
    named_cluster_vec <- sce_filt$named_cluster 
  }else if(clusterID_or_celltype == "cluster")
  {
    named_cluster_vec <- sce_filt$cluster
  }else if(clusterID_or_celltype == "long_named_cluster")
  {
    named_cluster_vec <- sce_filt$long_named_cluster
  }
  
  
  # Create a data frame with unique mapping between factor value and actual value
  df_mapping <- data.frame(
    named_cluster = named_cluster_vec,
    group = as.numeric(factor(named_cluster_vec))
  )
  
  # Remove duplicates to get one row per cluster
  df_mapping_unique <- unique(df_mapping)
  
  # Merge with extracted color info

  color_cluster_map <- df_colors %>%
    distinct(group, colour) %>%
    inner_join(df_mapping_unique, by = "group") %>%
    select(named_cluster, colour)
  # distinct(group, colour) Keeps only unique combinations of group and colour
  # View result
  # print(color_cluster_map)
  # Define your custom color palette (names must match your 'cell_type' exactly)
  
  # color_cluster_map <- base:::merge(color_cluster_map, unique(dat_annot[ , c("cluster_id", "cell_type")]) ,
  #                    by.x = "named_cluster", by.y = "cluster_id")
  # 
  # color_cluster_map$named_cluster <- paste(color_cluster_map$named_cluster,
  #                                          color_cluster_map$cell_type,
  #                                          sep = "_")
  # dat_annot$named_cluster <- paste(dat_annot$cluster_id, dat_annot$cell_type,
  #                                  sep = "_")
  
  cell_type_colors <- setNames(
    as.character(color_cluster_map[["colour"]]),  # values
    as.character(color_cluster_map[["named_cluster"]])   # names
  )
  return(cell_type_colors)
  
}

selected_point_size = 0.2
# plot automatic assigned color for each cell type and cluster in UMAP
sce_filt
p1 <- plotUMAP(sce_filt, colour_by="cluster", text_by = "cluster",  point_size = selected_point_size) + 
  guides(colour = guide_legend(override.aes = list(size=15, alpha = 1))) + 
  theme(text = element_text(size = 20)) + ggtitle(paste('UMAP', sampSet))
print(p1)
clusterid_colors <- output_color_code_in_UMAP (p1, "cluster")
print(clusterid_colors)  

p1 <- plotUMAP(sce_filt, colour_by = "named_cluster", text_by = "cluster", point_size = selected_point_size) +
  guides(colour = guide_legend(override.aes = list(size = 15, alpha = 1))) +
  ggtitle(paste('UMAP', sampSet))

print(p1)
cell_type_colors <- output_color_code_in_UMAP (p1, "named_cluster")
print(cell_type_colors)


p1 <- plotUMAP(sce_filt, colour_by = "long_named_cluster", text_by = "cluster", point_size = selected_point_size) +
  guides(colour = guide_legend(override.aes = list(size = 15, alpha = 1))) +
  ggtitle(paste('UMAP', sampSet))

print(p1)



p1 <- plotUMAP(sce_filt, colour_by = "named_cluster", text_by = "named_cluster", point_size = selected_point_size) + 
  guides(colour = guide_legend(override.aes = list(size = 15, alpha = 1), title = 'Cell type')) + 
  ggtitle(paste('UMAP', sampSet)) 
print(p1)  

# draw additional UMAP with cluster 11, 12 not been displayed for subset_adaptive_immune_cells requested by collaborators
if(bsp_num == "bsp_248" )
{
  if(sampSet == "subset_adaptive_immune_cells")
  {
    sce_tem  <- sce_filt[, sce_filt$cluster != "11"]  
    sce_tem  <- sce_tem[, sce_tem$cluster != "12"]    
  }else if(sampSet == "subset_non_immune_cells")
  {
    sce_tem  <- sce_filt[, sce_filt$cluster != "22"]
  }else
  {
    sce_tem  <- sce_filt
  }


  p1 <- plotUMAP(sce_tem, colour_by = "named_cluster", text_by = "cluster", point_size = selected_point_size) +
  guides(colour = guide_legend(override.aes = list(size = 15, alpha = 1))) +
  ggtitle(paste('UMAP', sampSet))
  print(p1)  
  p1 <- plotUMAP(sce_tem, colour_by = "long_named_cluster", text_by = "cluster", point_size = selected_point_size) +
  guides(colour = guide_legend(override.aes = list(size = 15, alpha = 1))) +
  ggtitle(paste('UMAP', sampSet))
  print(p1)    
}



```


```{r plot a UMAP with manually assigned color and legend}
# legend_colours <- BAL_colour_list
# names(legend_colours) <- sce_all_filt$long_named_cluster
# legend_colours <- legend_colours[!duplicated(names(legend_colours))]
# 
# # assign the elements' order in list legend_colours to keep them comsistent
# 
# # legend_colours <- legend_colours[c("Plasma Cells, 1, 4, 15", "CD8 T Cells, 2, 8c", "CD4 T Cells, 8b, 3, 17", "Macrophages, 5, 10", "Unknown, 6, 9, 11a, 14, 16", "B Cells, 7, 13", "Monocytes, 12c", "cDCs, 12b", "NK cells, 11b, 19")]
# 
# # # create a data frame to record color and cell types
# df_legend_colours <- data.frame(unlist(names(legend_colours)), unlist(legend_colours))
# names(df_legend_colours) <- c("Cell_Type", "Colour")
# df_legend_colours$Cell_Type <- factor(df_legend_colours$Cell_Type, levels=c(df_legend_colours$Cell_Type))
# df_legend_colours$dummy_data <- seq(1, length(df_legend_colours[,1]))
# 
# 
# # plot a UMAP with manually assigned color and size, lable by cluster, aspect.ratio
# 
# p1 <- plotUMAP(sce_all_filt, point_size=0, text_by="cluster", text_size=5) +
#   aes(colour = sce_all_filt$colour) +
#   geom_point(shape='.', size=1, alpha=0.5) +
#   geom_text(x=-10, y=14, size=30, label="C", color="black") +
#   theme(aspect.ratio = 0.5, plot.margin = unit(c(0,0,0,0), "cm"), text = element_text(size = 40))
# print(p1)
# png(filename='Results/UMAP_manually_assigned_color.png', height=1080, width=2160)
# print(p1)
# dev.off()
# 
# 
# # plot a UMAP with manually assigned color without cluster number label 
# plotUMAP(sce_all_filt, point_size=0) +
#   aes(colour = sce_all_filt$colour) +
#   geom_point(shape='.', size=1, alpha=0.5) +
#   geom_text(x=-10, y=14, size=30, label="C", color="black") +
#   theme(aspect.ratio = 0.5, plot.margin = unit(c(0,0,0,0), "cm"), text = element_text(size = 40))

```

```{r plot cell type legend of UMAPs, fig.width= 12, fig.height=8}

# # plot a legend with manually assigned color for each cell type and cluster 
# ggplot(data=df_legend_colours, aes(x=dummy_data)) +
# geom_point(aes(y=dummy_data, color=Cell_Type), size=10) +
# labs(x=NULL, y=NULL, color=NULL) +
# scale_color_manual(values = legend_colours) +
# guides(colour = guide_legend(ncol = 1)) + 
# theme(legend.position.inside=c(0.5,0), legend.direction="vertical", aspect.ratio=0, text = element_text(size = 30),
#       axis.text.x=element_blank(), axis.ticks.x=element_blank(),
#       axis.text.y=element_blank(), axis.ticks.y=element_blank())

```


add a heatmap of all defining gene markers for our cell types
```{r add a heatmap of all defining gene markers for our cell types}
# # install.packages('gplots')
# library(gplots)
# library(viridis)
# 
# defining_genes <- c("CD163", "FLT3", "CD14", "CD4", "CADM1", "CSF1R", "FCGR1A", "BCL6", "CCL17", "IRF8", "ADAM8", "PRDM1", "IRF4", "JCHAIN" , "TNFRSF17", "TNFRSF13B", "CD19", "PAX5", "EBF1", "MS4A1", "CD79B", "IRF8", "CD28", "CD5", "CD3D", "CD3E", "CD8A", "CD8B", "FOXP3", "E2F1", "E2F2", "KLRK1", "KLRB1", "KIT", "NCR1", "TBX21", "EOMES", "GATA3", "IL1RL1", "IL7R", "ID3", "ACTG1", "GPR37", "NAPSA", "SFTPA1", "RPS3A", "RPS26", "RPS11", "RPL37")
# 
# defining_sce <- sce_all_filt
# rownames(defining_sce) <- rowData(sce_all_filt)$hgnc_symbol
# # Remove forbidden rowData names (for generation of CompressedGRangesList)
# names(rowData(defining_sce))[which(names(rowData(defining_sce))=="start")]="gene_start"
# names(rowData(defining_sce))[which(names(rowData(defining_sce))=="end")]="gene_end"
# #subset to defining genes and calculate mean logcounts
# defining_sce <- defining_sce[rownames(defining_sce) %in% defining_genes, ]
# defining_matrix <- as.matrix(logcounts(aggregateAcrossCells(defining_sce, ids=defining_sce$long_named_cluster, statistics='mean', use.assay.type='logcounts')))
# 
# xAxis <- colnames(defining_matrix)
# #xColours <- ifelse(grepl("BAL", xAxis, fixed=TRUE), "blue", "orange")
# colour_df <- unique(as.data.frame(colData(defining_sce))[, c("long_named_cluster", "colour")])
# colour_df <- colour_df[match(xAxis, colour_df$long_named_cluster),]
# identical(colour_df$long_named_cluster, xAxis)
# Col_Colours <- as.character(colour_df$colour)
# # Col_Colours has the length of cluster number 
# head(Col_Colours)
# length(Col_Colours)
# png(filename=paste0("Results/", sampSet,'_heatmap_defining_genes.png'), height=2500, width=1080)
# heatmap.2(defining_matrix, margins=c(40,20), scale="row", col = bluered(100),
#           cexRow=4, cexCol=4,  trace="none", key=TRUE, keysize=2, key.title="", key.xlab="row Z-score",
#           lhei=c(1,20), lwid=c(1,8), ColSideColors=c(Col_Colours), 
#           main=paste(sampSet,'heatmap of cell type defining genes') )
# dev.off()
# 
# png(filename=paste0("Results/", sampSet,'_heatmap_defining_genes_scale.png'), height=1080, width=1080)
# heatmap.2(defining_matrix, scale="row", col = bluered(100), 
#           cexRow=4, cexCol=4,  trace="none", key=TRUE, keysize=2, key.title="", key.xlab="row Z-score",
#           lhei=c(1,9), lwid=c(1,2), ColSideColors=c(Col_Colours))
# dev.off()  
```


add a bubbleplot of defining genes
```{r add a bubbleplot of defining genes, fig.height=8}
# #reorder axis's using heatmap dendrogram
# pp <- heatmap.2(defining_matrix, scale="row")
# rowIndex <- pp$rowInd
# colIndex <- pp$colInd
# 

# 
# 
# png(filename=paste0('Results/', sampSet, '_bubbleplot_defining_genes.png'), height=3840, width=3840)
# plotDots(defining_sce, features=rownames(defining_sce), group="named_cluster", color=turbo(256)) +
#   theme_minimal(base_size = 50) +
#   theme(aspect.ratio = 6, axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x=element_blank(), axis.title.y=element_blank(), legend.key.size = unit(2, 'cm')) +
#   scale_y_discrete(limits=rownames(defining_matrix)[rowIndex]) +
#   scale_x_discrete(limits=colnames(defining_matrix)[colIndex])
# dev.off()
```




```{r exclude suspicious clusters and bad quality samples}

# when run this script on bsp248 data before subset cells, we remove sample_20 here
if(bsp_num == "bsp_248" & !grepl("subset", sampSet))
{
  sce_all_filt  <- sce_all_filt[, sce_all_filt$Sample != "sample_20"]
  sce_filt  <- sce_filt[, sce_filt$Sample != "sample_20"]  
}


# # Update shiny app with filtered data for easier visualization and identification. For easier visualization in app, pre-convert counts metrics to log scale. 
saveRDS(sce_filt, file = paste0('sce_',sampSet,'_corrected', '_with_named_refined_cluster.rds'))
saveRDS(sce_all_filt, file = paste0('sce_',sampSet,'_with_named_refined_cluster.rds'))

# save.image(file = paste0('differential_annotated.RData'))
```


# Differential abundance analysis

For running DA analysis, we looking for differential abundance/ cell type proportion within these groups.


```{r divide cells based on sorted population}



# load annotation table as meta information need to use later
if(bsp_num == "bsp_248")
{
  annot <- read.csv('references/sample_information.csv')
  head(annot)
  samps <- paste("sample", formatC(1:24, width=2, flag="0"), sep="_")
  annot$sample_id <- samps
  rownames(annot) <- annot$sample_id
  head(sce_all_filt$cull_date)
  
  # DPI_1 <- sce_all_filt[,sce_all_filt$cull_date == 1]
  # DPI_12 <- sce_all_filt[,sce_all_filt$cull_date == 12]
  
  
  DPI_1 <- sce_all_filt[, (sce_all_filt$cull_date == 1) | (sce_all_filt$cull_date == 12 & sce_all_filt$condition == "control")]
  DPI_12 <- sce_all_filt[, (sce_all_filt$cull_date == 12) | (sce_all_filt$cull_date == 1 & sce_all_filt$condition == "control")]    
  # table(sce_all_filt$cull_date == 1)
  # table(sce_all_filt$cull_date == 12 & sce_all_filt$condition == "control")
  # table((sce_all_filt$cull_date == 1) | (sce_all_filt$cull_date == 12 & sce_all_filt$condition == "control"))
  # colData(sce_all_filt)
  
  
  subsets <- list(DPI_1 = DPI_1, DPI_12 = DPI_12)
}

print(table(sce_all_filt$cluster))
print(table(sce_all_filt$Sample))
print(table(sce_all_filt$condition))
head(colData(sce_all_filt))
head(annot)
# table(annot$condition)

```

Our DA analysis will be run using a negative binomial test.


We'll consider a change in abundance (proportion of one cluster) with condition virus1 vs virus2 vs control.

```{r conduct Differential abundance analysis for each sorted population}


df_cluster_number_name  <- as.data.frame(unique(colData(sce_all_filt)[,c('named_cluster', 'cluster')]))
rownames(df_cluster_number_name) <- df_cluster_number_name$cluster
# head(df_cluster_number_name)

# conduct_DA conduct DA for DPI1, DPI12 for selected comparison
conduct_DA <- function(subsets, selected_comparison = 3)
{
  DA <- NULL
  
  for (i in 1:length(subsets) ) 
  {
    # subsets is list with 2 elements contain the data of DPI1, DPI12 each 
    DPI <- names(subsets[i])
    x <- subsets[[i]]
    
    abundances <- table(x$cluster, x$Sample)
    abundances <- unclass(abundances)
    # typeof(abundances)
    # write.csv(abundances, file = 'abundances_check.csv')
    # head(abundances)
    # head(colnames(abundances))
    
    # extract rows of samples used there, s1-8 
    meta_data <- subset(colData(x), select=c("condition", "run", "cull_date", "animal_id"))[match(colnames(abundances), x$Sample),] 
    
    y.ab <- DGEList(abundances, samples= meta_data)
    
    # filter some clusters which only exist in a few samples 
    keep <- filterByExpr(y.ab, group=y.ab$samples$condition, 
                         min.count = 10, min.total.count = 60)
    # default min.count = 10, min.total.count = 15, large.n = 10, min.prop = 0.7
    # min.count: numeric. Minimum count required for at least some samples.
    # min.total.count: numeric. Minimum total count required across all samples.
    # large.n: integer. Number of samples per group that is considered to be “large”.
    # min.prop: numeric. In large sample situations, the minimum proportion of samples in a group that a gene needs to be expressed in. See Details below for the exact formula.
    # here counts represent cell number 

    print(paste('FALSE show clusters need to be filtered as they only exist in a few samples', DPI))
    print(keep)    
    keep 
    # filters out some clusters 
    y.ab <- y.ab[keep,]
    
    
    y.ab$samples$condition <- factor(y.ab$samples$condition, levels=c('control', 'H1N1', 'PRCV'))
    y.ab$samples$run <- factor(y.ab$samples$run, levels=c('1', '2', '3'))
    
    
    design <- model.matrix(~ 0 + condition + run, data = y.ab$samples)
    print(design)
    contrasts <- makeContrasts(control_H1N1 = conditioncontrol-conditionH1N1, 
                               control_PRCV =  conditioncontrol-conditionPRCV, 
                               H1N1_PRCV = conditionH1N1-conditionPRCV, levels = design)
    
    print(contrasts)
    
    # nrow(abundances) represent the number of clusters been take account.
    if(nrow(abundances) > 2)
    {
      y.ab <- estimateDisp(y.ab, design, trend="none")
      # Setting trend="none" disables the fitting of a trend to the dispersion estimates. This means each gene's dispersion is estimated independently of its mean expression level. This can be useful if you believe the dispersion does not depend on the abundance, but it might lead to less stable estimates, especially for lowly expressed genes.
      # Setting robust=TRUE enables robust estimation methods, which help mitigate the influence of outliers on the model fitting. This is useful if your data contain outliers or if you want to ensure that the results are not unduly affected by a few extreme values.
      
      summary(y.ab$common.dispersion)
      # plotBCV(y.ab, cex=1)
      fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
      # Setting abundance.trend=FALSE disables the consideration of a trend in gene abundance during model fitting. This assumes that the quasi-likelihood dispersion does not depend on the average abundance of the genes. , this might be suitable if you believe there is no relationship between dispersion and mean expression.
      
    }else
    {
      y.ab <- estimateDisp(y.ab, design, trend="none")
      summary(y.ab$common.dispersion)
      fit.ab <- glmQLFit(y.ab, design, abundance.trend=FALSE)
      # print(fit.ab)
    }
    summary(fit.ab$var.prior)
    # summary(fit.ab$df.prior)
    # plotQLDisp(fit.ab, cex=1)
    res <- glmQLFTest(fit.ab, contrast=contrasts[, selected_comparison])
    
    restab <- res$table
    
    # print(head(restab$PValue))
    restab$FDR <- p.adjust(restab$PValue, method='fdr')
    restab$comparison <- colnames(contrasts)[selected_comparison]  
    # print(head(restab, 10))
    
    print(paste("cluster proportion has significantly changed (p.adjust) across condition: ", DPI))
    print(restab[ (restab$FDR < 0.05), ])
    
    # print(paste('significant P values ', DPI))
    # print(restab[ (restab$PValue < 0.05), ])    
    
    DA[[DPI]] <- restab
    
  }
  return(DA)  
}

# output DA results to excel files
output_DA_excel <- function(DA, file_name_tag = '')
{
  
  library(openxlsx)
  
  for(i in names(DA))
  {
    wb <- createWorkbook() 
    # print(i)
    DA_list_per_element <- DA[[i]]
    # colnames(DA_list_per_element)
    DA_list_per_element$cluster <-  paste0('cluster_', rownames(DA_list_per_element))
    # unique(DA_list_per_element$comparison)
    # Add a worksheet to the workbook with the created name
    addWorksheet(wb, sheetName = i)
    # Write the data frame to the worksheet
    writeData(wb, sheet = i, DA_list_per_element, rowNames = TRUE)
    # Save the workbook to a file
    saveWorkbook(wb, file = paste0('Results/DA_edgeR_', sampSet, file_name_tag, i, ".xlsx"), overwrite = TRUE)     
  }
  
  # run_2 table contain the results of run 1+2
}

DA_t1_t2 <- conduct_DA(subsets, selected_comparison = 3)
output_DA_excel(DA_t1_t2, file_name_tag='_t1_t2_')

# head(DA[["DPI_1"]])
# typeof(DA)
# names(DA)
DA_ctl_t1 <- conduct_DA(subsets, selected_comparison = 1)
output_DA_excel(DA_ctl_t1, file_name_tag='_ctl_t1_')

DA_ctl_t2 <- conduct_DA(subsets, selected_comparison = 2)
output_DA_excel(DA_ctl_t2, file_name_tag='_ctl_t2_')

```

```{r draw dot plots to direct visulize and compare cluster proportion between condition, fig.width= 8, fig.height= 6}
# draw dot plots to direct visualize and compare cluster proportion between condition
direct_t_test <- function(x = DPI_1, selected_DPI = "DPI_1")
{
  abundances <- table(x$cluster, x$Sample)
  abundances <- unclass(abundances)
  
  df_proportion <- prop.table(as.matrix(abundances), margin = 2)
  df_proportion <-as.data.frame(t(df_proportion)) 
  n_cluster <- unique(sce_all_filt$cluster)
  df_proportion$sample_id <- rownames(df_proportion)
  
  library(ggpubr)
  library(ggrepel)

  selected_anno <- annot[colnames(abundances), ]
  
  df_condition <- selected_anno[ , c("condition", "sample_id")]  
  df_pro_condition <- merge(df_condition, df_proportion)
  
  for (selected_cluster in n_cluster) 
  {
    # print(selected_cluster)
    df_pro_one <- df_pro_condition[ , c("sample_id", "condition", as.character(selected_cluster))]
    colnames(df_pro_one)[3] <- "proportion"
    compare_means(proportion ~ condition,  data = df_pro_one, 
                  method = "t.test")
    df_pro_one$condition = factor(df_pro_one$condition, levels=c('control', 'H1N1', 'PRCV'))
    df_pro_one$S_id = gsub("sample_", "S", df_pro_one$sample_id )     
    library(ggrepel)
    my_comparisons <- list( c("PRCV", "H1N1"), c("PRCV", "control"), c("control", "H1N1") )
    cell_type <- df_cluster_number_name[ selected_cluster, "named_cluster" ]
    library(ggpubr)    
    # View(df_pro_one)
    # colnames(df_pro_one)
    p <-  ggdotplot(df_pro_one, x = "condition", y = "proportion",
                    fill = "sample_id", color = "sample_id", 
                     size = 3) +
      ggtitle(paste("proportion of cluster", selected_cluster, 
                    cell_type, "across condition", selected_DPI))  +
      guides(color = guide_legend(override.aes = list(size = 10))) +
      facet_grid(. ~ condition,  scales="free_x" )
    
    # +
    #   stat_compare_means(comparisons = my_comparisons,
    #                      method = "t.test") 
        # geom_text(aes(label = S_id) , hjust = 1, nudge_x = 0.05,
        #        size = 2) +
    
    print(p)

    
    # p <- ggdotplot(
    #   df_pro_one, 
    #   x = "condition", 
    #   y = "proportion",
    #   color = "sample_id",  # Keeps the color by sample_id
    #   fill = "sample_id",   # Fills the dots with sample_id color
    #   size = 5              # Increases dot size
    # ) +
    #   ggtitle(paste("proportion of cluster", selected_cluster, cell_type, "across condition", selected_DPI)) +
    #   
    #   # Add statistical comparisons
    #   stat_compare_means(comparisons = my_comparisons, method = "t.test") 
    # 
    # print(p)
    
    # p <-  ggboxplot(df_pro_one, x = "condition", y = "proportion",
    #           color = "condition")  + ggtitle( paste("proportion of cluster", selected_cluster, "across condition")) +
    #   geom_jitter(aes(color = condition), width = 0.2, size = 3, alpha = 0.7) +
    #   stat_compare_means(comparisons = my_comparisons,
    #                      method = "t.test") 
    #   print(p)
    
    
    
    
    
  }
  
  
  
  
}

direct_t_test(DPI_1, selected_DPI = "DPI_1")
direct_t_test(DPI_12, selected_DPI = "DPI_12")
```


```{r draw composition bar plot to visulize and compare cluster proportion between condition, fig.width= 10, fig.height= 8}

visulize_abundance <- function(sce_all_filt, annot)
{

  dat <- as.data.frame(table(sce_all_filt$cluster, sce_all_filt$Sample))
  colnames(dat) <- c('cluster_id', 'sample_id', 'n_cells')
  library(dplyr)
  # Calculate proportion per sample
  dat <- dat %>%
    group_by(sample_id) %>%                         # Group by sample
    mutate(proportion = n_cells / sum(n_cells)) %>% # Compute proportion
    ungroup()                                       # Remove grouping
  sum(dat[dat$sample_id == "sample_01", "proportion"])

  
  dat$S_id <- gsub("sample_", "S", dat$sample_id ) 
  dat_annot <- merge(dat, annot[ , c("DPI", "condition", "sample_id")], by = "sample_id")
  dat_2 = data.frame(cluster_id = colData(sce_all_filt)$cluster,
                     long_named_cluster = colData(sce_all_filt)$long_named_cluster,
                     cell_type = colData(sce_all_filt)$named_cluster)
  dat_2 = unique(dat_2)  
  
  # add cell type column in dat_annot
  dat_annot <- merge(dat_annot, dat_2, by = "cluster_id", all.x = TRUE)
  dat_annot$cluster_type <-  paste(dat_annot$cluster_id, dat_annot$cell_type, sep = "_")
  dat_annot$DPI <- paste0("DPI_", dat_annot$DPI)
  colnames(dat_annot)[8] <- "type_cluster"
  head(dat_annot)
  p <- ggplot(data=dat_annot, aes(x=S_id, y=proportion, fill=cell_type)) +
    geom_bar(stat="identity") + 
    facet_wrap(DPI ~ condition,  scales="free_x")+
    scale_fill_manual(values = cell_type_colors) +
    ggtitle("cell_type_composition_barplot")
  print(p)
  
  png(filename = paste0('Results/DA_edgeR_', sampSet, '_composition_barplot_cell_type.png') , height=2600, width=1800, res = 300)
  print(p)
  dev.off()  
  
  
  # head(dat_annot)
  dat_annot$cluster_id <- as.factor(dat_annot$cluster_id)
  # names(clusterid_colors) <- as.integer(names(clusterid_colors))
  p <- ggplot(data = dat_annot, aes(x = S_id, y = proportion, fill = cluster_id)) +
    geom_bar(stat = "identity") + 
    facet_wrap(DPI ~ condition, scales = "free_x") +
    scale_fill_manual(values = clusterid_colors) +
    ggtitle("clusterid_composition_barplot")
  
  print(p)
  
  png(filename = paste0('Results/DA_edgeR_', sampSet, '_composition_barplot_cluster.png') , height=3000, width=2600, res = 300)
  print(p)
  dev.off() 
  
  
  for (selected_cluster in levels(dat_annot$cluster_id)  ) 
  {
    cluster_dat_annot <- dat_annot[dat_annot$cluster_id == selected_cluster , ]
    selected_cluster <- unique(cluster_dat_annot$cluster_type ) 
    cluster_dat_annot$S_id <- factor(cluster_dat_annot$S_id, levels= cluster_dat_annot$S_id[order(cluster_dat_annot$condition)])
    
    # draw facet bar plots to display sample composition for each cluster
    p <- ggplot(cluster_dat_annot, aes(x=S_id, y=n_cells, fill=condition))+
      geom_bar(stat='identity') +
      facet_wrap(~ DPI,  scales="free_x")+
      ggtitle("cell number of selected cluster", paste("cluster", selected_cluster)) +
      geom_text(aes(label=n_cells))+
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
    print(p)
    
    p <- ggplot(cluster_dat_annot, aes(x=S_id, y=proportion, fill=condition))+
      geom_bar(stat='identity') +
      facet_wrap(~DPI,  scales="free_x") +
      ggtitle("proportion of selected cluster", paste("cluster", selected_cluster)) +
      geom_text(aes(label=round(proportion, digits = 3) ))+
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
    print(p)
  }
  
  for (selected_DPI in unique(dat_annot$DPI) ) 
  {
    DPI_dat_annot <- dat_annot[dat_annot$DPI == selected_DPI , ]

    
#################################################    
    # code below is used to generate DPI_ordered_S_id, which order sample id in x-axis
    # need to choose a cluster that every sample has cells in this cluster, we first find a cluster meet this requirement 
    df_cluster_freq_in_sample <- 
      dat_annot %>%
      group_by(cluster_id) %>%
      summarise(n = n())
    
    qualified_clusters  <- df_cluster_freq_in_sample$cluster_id[df_cluster_freq_in_sample$n == length(unique(dat_annot$sample_id) )]
    if(length(qualified_clusters) > 0)
    {
    print(paste('select cluster', qualified_clusters[1], 'that every sample has cells in this cluster'))
      
    }else
    {
      print("warning: none cluster meet requirement that every sample has cells in this cluster")
    }
 #################################################   

    # order the S_id to make S_id of same condition stay together in plot below
    cluster_dat_annot <- dat_annot[dat_annot$cluster_id == qualified_clusters[1] , ]
    cluster_dat_annot$S_id <- factor(cluster_dat_annot$S_id, levels= cluster_dat_annot$S_id[order(cluster_dat_annot$condition)])
    DPI_ordered_S_id <-  levels(cluster_dat_annot$S_id)[(levels(cluster_dat_annot$S_id) %in% unique(DPI_dat_annot$S_id))]
 #################################################       
       head(DPI_dat_annot) 
    p <- ggplot(DPI_dat_annot, aes(x=S_id, y=proportion, fill=condition))+
      geom_bar(stat='identity') +
      facet_wrap(~ cluster_type) + 
      scale_x_discrete(limits=DPI_ordered_S_id) +
      ggtitle("proportion of each cluster", selected_DPI) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))  
    print(p)
    
    png(filename = paste0('Results/DA_edgeR_', sampSet, '_proportion_barplot_allcluster',selected_DPI, '.png') , height=5000, width=5000, res = 500)
    print(p)
    dev.off() 
    
    
    p <- ggplot(DPI_dat_annot, aes(x=S_id, y=n_cells, fill=condition))+
      geom_bar(stat='identity') +
      facet_wrap(~ cluster_type) + 
      scale_x_discrete(limits=DPI_ordered_S_id) +
      ggtitle("n_cells of each cluster", selected_DPI) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))  
    print(p)
    
    png(filename = paste0('Results/DA_edgeR_', sampSet, '_ncells_barplot_allcluster',selected_DPI, '.png') , height=5000, width=5000, res = 500)
    print(p)
    dev.off()
    
  }
  
  
}
# save.image('differential_lily_develop.RData')
visulize_abundance(sce_all_filt, annot)


```

Let's plot box plots of proportions between 2 conditions to have a look.

```{r plot proportion box plots to visulize cell proportion between conditions}


# in bsp185, we only run code below to plot those clusters whose composition changed significantly in certain main populations

boxplot_cluster_proportion <- function(selected_dataset, selected_cluster = '2',
                                       flag = 'total', composition = FALSE)
{
  if(composition)
  {
    # plot the stacked bar chart to show the percentage of each cluster in a sample among sorted CD4 population 
    # calculate proportion of each cluster within each sample among sorted CD4 population
    # calculate proportion of cluster 4a within sample 9 among sorted CD4 population
    
    props <- prop.table(table(selected_dataset$cluster, selected_dataset$Sample), margin=2)
    
    props <- data.frame(t(props))
    colnames(props) <- c('sample', 'cluster', 'prop')
    props <- merge(props, annot[, c("Condition", "sample")], by= "sample")

    props$sample <- as.factor(props$sample)        
    # props$sample <- factor(props$sample, levels=c("sample_09", "sample_14", "sample_15", 
    #                                               "sample_16", "sample_10", "sample_11", 
    #                                               "sample_12", "sample_13"))
    as.character(props$cluster)
    props <- merge(props, df_cluster_number_name, by= "cluster")
    

    
    # props$named_cluster[grep('non-conventional T cells', props$cell_type)] <- 'NCT'
    props$named_cluster <- paste(props$cluster, props$named_cluster, sep = ", ")
    
    p <- ggplot(props, aes(fill=named_cluster, y=prop, x=sample)) +
      geom_bar(position='stack', stat='identity') +
      facet_grid(~Condition, scales = "free_x", space = "free_x")+
      theme_classic() +  # Apply classic theme with white background
      theme(axis.text.x = element_text(angle = 90)) + 
      scale_fill_brewer(palette = "Pastel1")   # better color palette      
    print(p)
    # The scales = "free_x" and space = "free_x" options ensure that the x-axis is shared but allows different sample groupings for each condition.
    
    ## and print it
    print(paste0('Results/', sampSet, '_proportion_each_cluster_', flag, '.png'))
    png(filename=paste0('Results/', sampSet, '_proportion_each_cluster_', flag, '.png'))
    print(p)
    dev.off()  
    # ggplot(props, aes(fill=cluster, y=prop, x=sample)) +
    #   geom_bar(position='stack', stat='identity') +
    #   labs(x=NULL, y='proportion') +
    #   theme_classic() +
    #   theme(axis.text.x = element_text(angle = 45, hjust=1), text=element_text(size=21))
  }
  
  
  
  
  
  ## calculate the proportion of cluster 2 in selected_dataset cells in each sample, 
  # eg: cells of cluster 2 composed 14% of cells in sample 1 in sorted CD4 population
  props <- prop.table(table(selected_dataset$cluster, selected_dataset$Sample), margin=2)[selected_cluster,]

  # print(props)
  proptab <- annot[ ,c('sample_analysis_id', 'condition')]
  proptab$proportions <- props[rownames(proptab)]
  proptab <-  proptab[!is.na(proptab$proportions) , ]

  # proptab$Condition <- factor(proptab$Condition)
  proptab$condition <- factor(proptab$condition, levels = c('control', 'PRCV', 'H1N1'))
  cell_type = df_cluster_number_name[df_cluster_number_name$cluster == selected_cluster, 'named_cluster']
  group_colors <- c('PRCV' = 'red', 'H1N1' = 'blue', 'control' = 'black')
  
  p <- ggplot(data=proptab, aes(x=condition, y=proportions)) +
    geom_boxplot(outlier.shape = NA) +
    geom_point(aes(col=condition), position='jitter', size = 5) + 
    ggtitle(paste(sampSet, flag , '\n', 'proportion of cluster', selected_cluster, cell_type)) +
    scale_color_manual(values = group_colors)
  print(p)
  
  png(filename=paste0('Results/', sampSet, '_DA_cluster', selected_cluster, '_prop_', flag, '.png'), height=900, width=1080)  
  print(p)
  dev.off()
  
}

myList  <- list("H1N1_PRCV" = DA_t1_t2, "control_H1N1" = DA_ctl_t1, "control_PRCV"= DA_ctl_t2)
for (name in names(myList)) 
{
  print(name)
  for(element_name in names(myList[[name]]))
  {
    print(element_name)
    sig_cluster_df <- myList[[name]][[element_name]]
    sig_cluster_df <- sig_cluster_df[sig_cluster_df$FDR < 0.05, ]
    if(nrow(sig_cluster_df) > 0)
    {
      print(sig_cluster_df)
      sig_clusters <- rownames(sig_cluster_df)
      for(cluster1 in sig_clusters)
      {
        boxplot_cluster_proportion(selected_dataset = subsets[[element_name]], selected_cluster = cluster1, 
                                   flag = paste(name, element_name, "DA"))
      }
    }
  }
  
  

}


# # if(sampSet == 'run_1' )
# # {
  # boxplot_cluster_proportion(selected_dataset = subsets$DPI_12, selected_cluster = '2re', flag = 'DPI_12 DA')
#   # boxplot_cluster_proportion(selected_dataset = CD8s, selected_cluster = '2', flag = 'CD8')
#   # boxplot_cluster_proportion(selected_dataset = all, selected_cluster = '8c', flag = 'total')
#   # boxplot_cluster_proportion(selected_dataset = CD8s, selected_cluster = '8c', flag = 'CD8')
#   # 
#   # ## calculate the proportion of cluster 17 in cells sorted as CD4s in each sample
#   # ## proportion within those cells sorted as CD4s
#   # boxplot_cluster_proportion(selected_dataset = all, selected_cluster = '17', flag = 'total')
#   # boxplot_cluster_proportion(selected_dataset = CD4s, selected_cluster = '17', flag = 'CD4')
#   
# }
# if(sampSet == 'run_2' )
# {
#   boxplot_cluster_proportion(selected_dataset = CD4s, selected_cluster = '4a', flag = 'CD4', composition = TRUE, annot=annot)  
#   boxplot_cluster_proportion(selected_dataset = CD4s, selected_cluster = '4b', flag = 'CD4', annot=annot) 
#   boxplot_cluster_proportion(selected_dataset = CD4s, selected_cluster = '8', flag = 'CD4', annot=annot) 
#   
#   boxplot_cluster_proportion(selected_dataset = MMDs, selected_cluster = '20', flag = 'MMDs', composition = TRUE, annot=annot)  
#   
#   # boxplot_cluster_proportion(selected_dataset = Bs, selected_cluster = '17', flag = 'Bs')  
# }

```

# Differential expression analysis

Now we move on to differential expression analyses. 
all samples are from female pigs, we only compare samples in same sequencing run at this stage. compare two conditions 135 (s5-8) vs ISU (s1-4)


Now we test for differential expression. This analysis will be run as a pseudo-bulk analysis, pooling counts from all of the cells within each cluster for each sample before treating like bulk RNAseq data. 

Differential expression will then be analysed for each cluster. For this first pass, 

```{r create functions and variables needed later for DE analysis}

plot_volcano <- function(df_re, title) 
{
  # here we use p values rather than p adjusted value to try
  # df_re$p_adjust <- df_re$PValue
  set.seed(1)
################################################# 
  # v1.14.6
  logp_to_power <- function(y) (10^(-y))^(-0.5)  # inverse for breaks

  ggplot(data=df_re, aes(x = logFC, y = p_adjust**-0.5 , col=diffexpressed, label=delabel)) +
    geom_hline(yintercept=0.05**-0.5, col="red") +
    geom_point( alpha = 0.3) +
    geom_text_repel() +
    ggtitle(title, paste("DE genes")) +
    labs( x = "logFC", y = "-log10(adjusted p-value)") +
    scale_color_manual(values=c("blue", "black", "red"), breaks=c("DOWN", "NO", "UP"), guide="none")+
    theme(axis.title.x = element_text(size = 30),
          axis.title.y = element_text(size = 30),
          plot.margin = margin(5.5, 60, 5.5, 5.5),
          axis.text=element_text(size=10)) +
    coord_cartesian(clip = "off")  +
    scale_y_continuous(
    breaks = logp_to_power(c(1, 2, 3, 4, 5, 6)),   # positions in power scale
    labels = c(1, 2, 3, 4, 5, 6)  ) 
      # scale_y_continuous(breaks = c(0, 0.05**-0.5, 0.01**-0.5, 0.001**-0.5, 0.0001**-0.5) )+

  ################################################# 
  # # v1.14.5
  # ggplot(data=df_re, aes(x = logFC, y = -log10(p_adjust), col=diffexpressed, label=delabel)) +
  #   geom_point( alpha = 0.3) +
  #   geom_text_repel() +
  #   ggtitle(title, paste("DE genes")) +
  #   labs( x = "logFC", y = "-log10(p value)") +
  #   geom_hline(yintercept=-log10(0.05), col="red") +
  #   scale_color_manual(values=c("blue", "black", "red"), breaks=c("DOWN", "NO", "UP"), guide="none")+
  #   theme(axis.title.x = element_text(size = 15),
  #         axis.title.y = element_text(size = 15),
  #         plot.margin = margin(5.5, 60, 5.5, 5.5),
  #         axis.text=element_text(size=10)) +
  #   coord_cartesian(clip = "off")  +
  #   scale_y_continuous(trans = scales::modulus_trans(p = 5),  breaks = c(0, 1, 2, 3, 4) )
  # 
  
      
  #   ggplot(data=df_re, aes(x = logFC, y = -log10(p_adjust), col=diffexpressed, label=delabel)) +

  #     geom_point( alpha = 0.3, position = "jitter") +
 # Customize axis labels here if needed
                   # X-axis label font size
# extra right margin for labels
    # 4) compress buttom region, expand upper 
  # trim anything below 0
  # xlim(-3, 3) + ylim(0, -log10(y_limit)) +     
  # plot.title = element_text(size = 20),        # Title font size and style
      # geom_vline(xintercept=c(-2, 2), col="green") +
      # geom_vline(xintercept=c(-0.6, 0.6), col="red") +
## Warning: Removed 7866 rows containing missing values or values outside the scale range
## (`geom_text_repel()`).
## Warning: ggrepel: 191 unlabeled data points (too many overlaps). Consider
## increasing max.overlaps
  # as there is insufficient space in the plot to display gene symbol of each dot, warning above appear. 
}  

mod_data <- function(df_re, i, cell_type, flag) 
{
  y_limit <- 1.0e-20
  # add a column of NAs
  df_re$diffexpressed <- "NO"
  # if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
  df_re$diffexpressed[df_re$logFC > 0.6 & df_re$p_adjust < 0.05] <- "UP"
  # if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
  df_re$diffexpressed[df_re$logFC < -0.6 & df_re$p_adjust < 0.05] <- "DOWN"
  
  
  # df_re$diffexpressed[df_re$logFC > 0.6 & df_re$PValue < 0.05] <- "UP"
  # df_re$diffexpressed[df_re$logFC < -0.6 & df_re$PValue < 0.05] <- "DOWN"
  table(df_re$diffexpressed)
  
  write.table(df_re, 
              file=paste0('Results/DE_edgeR_', sampSet, flag, '_cluster_', i, '.csv'),
              sep=',', quote=FALSE, row.names = FALSE)  
  
  df_re_clean <- df_re[ df_re$diffexpressed != "NO", ]
  df_re$delabel <- NA
  df_re$delabel[df_re$diffexpressed != "NO"] <- df_re$symbol[df_re$diffexpressed != "NO"]  
  df_re$delabel[startsWith(df_re$delabel, "ENSSSCG")] <- NA 
  
  
  if(nrow(df_re_clean) > 0)
  {
    # dir.create('Results')
    write.table(df_re_clean, 
                file=paste0('Results/DE_edgeR_', sampSet, flag, '_clean_cluster_', i, '.csv'),
                sep=',', quote=FALSE, row.names = FALSE)     
  }
 
  
  
  if(sum(df_re$diffexpressed != "NO") == 0)
  {
    
    # print(paste('no qualified DE genes observed in this cluster', i, ', no volcano plot'))
    return(df_re)
  }else
  {
    
    # print(paste('qualified DE genes observed in this cluster', i, ', show volcano plot'))

    # for row/gene who up/down, its delabel column will store the gene symbol again
    # for row/gene who is not DE, its delabel column is NA
    # Set p-value floor
    # df_re$adj.P.Val[df_re$adj.P.Val <= y_limit] <- y_limit 
    
    p1 <- plot_volcano(df_re, title=paste(cell_type, "cluster", i, "comparison", flag))
    # print(p1)
    
    png(filename = paste0('Results/DE_edgeR_', sampSet, flag, '_volcano_cluster', i, '.png') , height=1600, width=1600, res = 220)
    # when we set height=1080, width= higher, more gene symbols can shown on the plot since there are more space 
    print(p1)
    dev.off()    
  }
  
  return(df_re)
}

# integrate and output DE of each cluster into one single excel file
output_DE_to_files <- function(de.results.any, flag= 't1_t2_DPI1_')
{
  # print(flag)
  cls <- lapply(de.results.any, class)
  cls <- as.character(cls)
  # print(cls)
  # check which cluster fail to have DE results and remove these clusters in de.results.any
  
  if(length(which(cls == 'character')) > 0)
  {
    pos_fail_DE <- which(cls == 'character')
    print('delete DE results of clusters below as it has invalid DE results')
    print(de.results.any[pos_fail_DE])
    de.results.any <- de.results.any[-pos_fail_DE]
    
  }
  n_cluster <- length(de.results.any)
  summary_df <- data.frame('DOWN' = rep(0, n_cluster),'NO' = rep(0, n_cluster),
                           'UP' = rep(0, n_cluster))
  rownames(summary_df) = names(de.results.any)
  
  library(openxlsx)
  # Create a new workbook
  wb <- createWorkbook()
  
  for(i in names(de.results.any))
  {
    # print(i)
    output_df <- de.results.any[[i]]
    
    
    
    output_df$p_adjust <- p.adjust(output_df$PValue, method='fdr')
    
    output_df$symbol <- rowData(sce_all_filt)[rownames(output_df), 'Symbol']
    output_df$ensemble_id <- rownames(output_df)  
    
    output_df <- output_df[order(output_df$p_adjust),]
    # head(output_df$p_adjust,50)    
    de.results.any[[i]] <- output_df
    # change format GHSR, TNFSF10 into GHSR;TNFSF10 for those duplicates in our DE (edgeR) csv files. 
    ###################################
    my_vector <- output_df$symbol
    # Function to replace ', ' with ';' for those Ensembl gene ID who has duplicates symbols
    replace_comma <- function(x) {
      gsub(", ", ";", x)
    }
    # Identify elements that match the pattern ', '
    matching_indices <- grep(", ", my_vector)
    
    # Apply the function to identified elements and print each changed element
    for (index in matching_indices) {
      original_element <- my_vector[index]
      changed_element <- replace_comma(original_element)
      # print(paste(original_element, 'after modify', changed_element))
      
    }   
    changed_vector <- sapply(my_vector, replace_comma)
    length(changed_vector)
    dim(output_df)
    output_df$symbol <- changed_vector
    #############################################
    # write.table(output_df, 
    #             file=paste0('Results/', sampSet, '_', flag,'_DE_cluster_', i, '.csv'), 
    #             sep=',', quote=FALSE, row.names = FALSE)
    
    
    sheet_name <- paste("cluster", i)
    # Add a worksheet to the workbook with the created name
    addWorksheet(wb, sheetName = sheet_name)
    # Write the data frame to the worksheet
    writeData(wb, sheet = sheet_name, output_df)
    
    cell_type = df_cluster_number_name[df_cluster_number_name$cluster == i, 'named_cluster']
    # if(flag == 'm2')
    # {
    # only run mod_data function for method 2, it will output final DE files per cluster that would be used for downstream enrichment analysis
    output_df <- mod_data(output_df, i, cell_type, flag)      
    # }
    
    
    
    if(length(table(output_df$diffexpressed)) == 1)
    {
      
      # print(paste('no significant DE in cluster', i))
      # print(table(output_df$diffexpressed))      
    }else
    {
      vector1 <- table(output_df$diffexpressed)
      summary_df[i,'DOWN'] <- vector1['DOWN']
      summary_df[i,'UP'] <- vector1['UP']
      # print(paste('remain gene number is', nrow(output_df), 'exist significant DE in cluster', i))
      # print(table(output_df$diffexpressed))
      
    }
    
    # add column diffexpressed to denote a DE is up or down expressed
  }
  
  # Save the workbook to a file
  file_name = paste0('Results/DE_edgeR_', sampSet, flag, "_all_cluster.xlsx")
  saveWorkbook(wb, file = file_name, overwrite = TRUE)
  print(paste('output excel file store the DE results in', file_name ))
  
  for(i in names(de.results.any))
  {
    # print(paste('TRUE means DE genes whose FDR below 0.05 in cluster', i))
    # print(table(de.results.any[[i]]$p_adjust < 0.05))
    
  }
  rownames(summary_df) = paste0('cluster_', names(de.results.any))
  return(summary_df)
}  

# function to conduct core edgeR DE analysis
pseudoBulkDGEmod_m2 <- function(clust, sce, filtexprsmin=20, filtexprstotal=40,
                                selected_comparison = 3, DPI = 1)
{
  sc <- sce[,sce$cluster==clust]
  
  colData(sc) <- colData(sc)[ , c("Sample", "condition", "run", "cull_date", "animal_id", "cluster",
                                  "named_cluster", "ncells" )]
  # head(colData(sc))
  # colnames(colData(sc))
  rownames(colData(sc)) <- colData(sc)$Sample   
  
  # retrieve 8 samples in cluster 1
  y <- DGEList(counts=counts(sc), samples=colData(sc))
  

  if(length(which(table(y$samples$condition) < 3)) > 0)
  {
    # if repelicates under any condition less than 3, we don't obtain enough repelicates for valid statistic test
    print(paste('check condition replicates in DE edgeR in cluster', clust, 'DPI', DPI))
    print(table(y$samples$condition))    
    print("one or more condition contain less than 3 replicates after fitering sample whose cell number < 10")
    # return()
  }
  
  keep <- filterByExpr(y, group=sc$condition, min.count=filtexprsmin,
                       min.total.count=filtexprstotal)
  # Determine which genes have sufficiently large counts to be retained in a statistical analysis.
  # min.count: a gene needs to have at least 20 counts in at least some samples to be retained
  # filter genes that has total counts below 40 across all samples
  
  y <- y[keep,]
  
  remain_genes <- rownames(y)
  
  print(paste("after fitering low-expressed genes, genes retained in cluster", clust, "is", 
              length(remain_genes)) )
  # those retained genes will be statistically tested to see whether differentially expressed 
  
  y <- calcNormFactors(y)
  # Calculate scaling factors to convert the raw sample/library sizes into normalized effective sample/library sizes. since raw counts are requested to enter 
  y$samples$condition <- factor(y$samples$condition, levels = c('control', 'H1N1', 'PRCV'))
  y$samples$run <- factor(y$samples$run, levels=c('1', '2', '3'))
  
  
  # par(mfrow=c(2,3))

  # if(length(which(table(y$samples$condition) < 3)) > 0)
  # {
  #   print(' fail to plot MDS, one or more condition has less than 3 replicates')
  # }else
  # {
  # limma:::plotMDS(cpm(y, log=TRUE), labels=y$samples$condition ,
  #                 col = c('red', 'blue', 'turquoise')[as.numeric(factor(y$samples$condition))],
  #                 main = paste('cluster:', clust, 'in DPI', DPI))    
  # }


  design <- model.matrix(~ 0 + condition + run, y$samples)
  # print(design)
  # we don't compare baseline with treatment there, we compare 2 treatment there, so i use 0 + factor 
  contrasts <- makeContrasts(control_H1N1 = conditionH1N1-conditioncontrol, 
                             control_PRCV =  conditionPRCV-conditioncontrol, 
                             H1N1_PRCV = conditionPRCV-conditionH1N1, levels = design)
  
  
  # print(contrasts)
  
  tc <- tryCatch(estimateDisp(y, design), error=function(e) e, warning=function(w) w)
  if(is(tc, 'error') | is(tc, 'warning'))
  {

    print(paste("estimateDisp error or warning in cluster", clust) )
    return(paste(tc))
  }else
  {
    y <- estimateDisp(y, design)
    # The NB dispersion is estimated using the estimateDisp function and visualized with plotBCV.
    # plotBCV(y, main=clust)
    fit <- glmQLFit(y, design, robust=TRUE)
    # estimate quasi-likelihood (QL) dispersions, visualized with plotQLDisp.
    # plotQLDisp(fit)
    res <- glmQLFTest(fit, contrast=contrasts[, selected_comparison])
    
    # We then perform quasi-likelihood F-test for contrast 2 conditions.
    # topTags(res)
    restab <- res$table
    restab$comparison <- colnames(contrasts)[selected_comparison]  
    # print(head(restab))        
    return(restab)
    
  }

}

# plot up and down DE genes number per cluster bar plot
DRAW_up_down_summary_barplot <- function(summary_df, prefix = "_t1_t2_DPI1")
{
  # plot up and down DE genes number per cluster bar plot
  # Melt the data frame for ggplot2
  library(reshape2)
  summary_df$cluster <-  gsub("cluster_", "", rownames(summary_df))
  summary_df$NO <- NULL


  df_cell_type_cluster = data.frame(cluster = colData(sce_all_filt)$cluster,
                                    cell_type = colData(sce_all_filt)$named_cluster)
  df_cell_type_cluster = unique(df_cell_type_cluster)

  # merge two data frames by ID and Country
  summary_df <- merge(df_cell_type_cluster ,summary_df ,by=c("cluster"))
  summary_df$named_cluster <- paste(summary_df$cluster, summary_df$cell_type, sep = ', ' )

  # reduce the length of some cell type names
  # summary_df$cell_type[grep('Monocytes', summary_df$cell_type)] <- 'Monocyte'
  # summary_df$cell_type[grep('conventional', summary_df$cell_type)] <- 'NCT'
  # summary_df$cell_type[grep('T regs', summary_df$cell_type)] <- 'Tregs'

  # summary_df$cell_type[grep('Mitotic CD4 T Cells', summary_df$cell_type)] <- 'mCD4T'
  # summary_df$cell_type[grep('Mitotic CD8 T Cells', summary_df$cell_type)] <- 'mCD8T'
  #
  # summary_df$cell_type[grep('CD8 T Cells', summary_df$cell_type)] <- 'CD8T'
  # summary_df$cell_type[grep('CD4 T Cells', summary_df$cell_type)] <- 'CD4T'


  summary_df_sim <- summary_df[,c("DOWN", "UP", "named_cluster")]
  melted_df <- melt(summary_df_sim, id.vars = 'named_cluster')
  melted_df <- merge(melted_df ,summary_df[,c('named_cluster', 'cell_type')] ,by=c("named_cluster"))



  # Plotting overview
  p1 <- ggplot(melted_df, aes(x = named_cluster, y = ifelse(variable == "DOWN", -value, value),
                              fill = variable)) +
    geom_bar(stat = 'identity', position = 'identity') +
    labs(x = 'named_cluster', y = 'Number of Genes') +
    ggtitle(label = sampSet,
          subtitle = paste(prefix, 'edgeR summary of DE genes per cluster') ) +
    scale_fill_manual(values = c('UP' = 'red', 'DOWN' = 'blue')) +
    theme_minimal() +
    geom_text(aes(label = value,
                  vjust = ifelse(variable == "DOWN", 1.2, -0.5)),  # Position text inside or outside the bar
              color = "black",
              size = 3) +
    theme(axis.text.x = element_text(angle = 90)) +
    facet_grid(~ cell_type, scales = "free_x", space = "free_x") +
    theme(
      strip.background = element_rect(color = "black", fill = "white", linewidth = 1))  # Add black rectangle border

  print(p1)




  # # monocytes, macrophages, cDCs, pDCs, NCT, CD8T, mCD8T, CD4T, mCD4T, Tregs (please without a blank between "T" and "regs") B cells, plasma cells.
  # # assign order of clusters in x-axis
  # melted_df$cell_type <- factor(melted_df$cell_type, levels = c('Monocyte', 'Macrophages',
  #                                                               'cDCs', 'pDCs', "NCT", "CD8T","mCD8T","CD4T","mCD4T",
  #                                                               'Tregs', 'B Cells', "Plasma Cells" ))
  # unique(melted_df$cell_type)

  # Save the plot with larger dimensions for better clarity
  png(filename = paste0('Results/DE_edgeR_', sampSet, prefix, '_DEG_num_overview.png'),
      height = 1000, width = 2000, res = 150)  # Adjust height, width, and resolution for better clarity
  p1 <- ggplot(melted_df, aes(x = named_cluster, y = ifelse(variable == "DOWN", -value, value),
                              fill = variable)) +
    geom_bar(stat = 'identity', position = 'identity') +
    labs(x = 'annotated cluster',  # Custom x-axis title
         y = 'Number of DE Genes',  # Custom y-axis title
         title = paste(prefix, 'edgeR summary of DE genes per cluster'),  # Custom plot title
         fill = 'change') +  # Custom legend title
    scale_fill_manual(values = c('UP' = 'red', 'DOWN' = 'blue')) +
    theme_minimal() +
    geom_text(aes(label = value,
                  vjust = ifelse(variable == "DOWN", 1.2, -0.5)),  # Position text inside or outside the bar
              color = "black",
              size = 4) +  # Increase text size inside bars
    theme(axis.text.x = element_text(angle = 90, size = 10),  # Increase x-axis text size
          axis.text.y = element_text(size = 10),  # Increase y-axis text size
          axis.title.x = element_text(size = 10, face = "bold"),  # Increase x-axis label size
          axis.title.y = element_text(size = 10, face = "bold"),  # Increase y-axis label size
          plot.title = element_text(size = 10, face = "bold", hjust = 0.5),  # Increase title size and center
          strip.text = element_text(size = 9),  # Increase facet label size
          legend.text = element_text(size = 10),  # Increase legend text size
          legend.title = element_text(size = 10, face = "bold")) +  # Increase legend title size
    facet_grid(~ cell_type, scales = "free_x", space = "free_x") +
    theme(strip.background = element_rect(color = "black", fill = "white"))  # Add black rectangle border
  print(p1)
  dev.off()
  return(melted_df)

}


```




```{r apply method 2 conduct DE analysis between 2 virus condition for each cluster and output formal DE results,  fig.width= 12, fig.height=8}

DE_analysis <- function(sce_all_filt, selected_DPI, Select_comparison = 3, comparison_prefix = "t1_t2")
{
  # at this stage, sce_all_filt contain 30077 genes, 37511 cells, 8 samples, 21 clusters
  
  
  length(unique(sce_all_filt$cluster))
  table(sce_all_filt$cluster)
  # head(colData(sce_all_filt)) 
  
  # merge 4 control samples as one group
  sce_all_filt_DPI_selected <- sce_all_filt[, (sce_all_filt$cull_date == selected_DPI) 
                                            | (sce_all_filt$cull_date != selected_DPI & sce_all_filt$condition == "control")]
  
  # not merge 4 control samples as one group  
  # sce_all_filt_DPI_selected <- sce_all_filt[ , sce_all_filt$cull_date == selected_DPI]
  
  
  # pooling/sum counts from all cells within each cluster for each sample 
  # Differential expression will then be analysed based on each sample-cluster pair
  summed <- aggregateAcrossCells(sce_all_filt_DPI_selected, 
                                 id=colData(sce_all_filt_DPI_selected)[ ,c("cluster", "Sample")])
  
  
  # inside summed@assays@data@listData[["counts"]], the resulting columns should be 21*8=168 (168 sample-cluster pairs) rather than 167, is there a mistake? no, it's because a few samples have 0 cell in certain clusters.
  # table(summed$cluster, summed$Sample) 
  # View(summed@assays@data@listData[["counts"]])
  head(colData(summed)) 
  
  summed.filt <- summed[,summed$ncells >= 10]
  # remove those sample-cluster pairs when one sample has cell number below 10 in certain cluster

  
  
  # write.csv(table(summed.filt$cluster, summed.filt$Sample), file = "summed.filt_check.csv")
  # dim(summed.filt)
  # head(summed.filt@colData)
  
  #######################################################################
  # plot MDS for all valid samples-cluster pair
  sc <- summed.filt
  
  colData(sc) <- colData(sc)[ , c("Sample", "condition", "run", "cull_date", "animal_id", "cluster",
                                  "named_cluster", "ncells" )]
  # head(colData(sc))
  # colnames(colData(sc))
  rownames(colData(sc)) <- paste(colData(sc)$Sample, colData(sc)$cluster, sep = "_cluster")
  
  # retrieve 8 samples in cluster 1
  y <- DGEList(counts=counts(sc), samples=colData(sc))
  
  keep <- filterByExpr(y, group=sc$condition, min.count=20,
                       min.total.count=40)
  # Determine which genes have sufficiently large counts to be retained in a statistical analysis.
  # min.count: a gene needs to have at least 20 counts in at least some samples to be retained
  # filter genes that has total counts below 40 across all samples
  
  y <- y[keep,]
  
  remain_genes <- rownames(y)
  
  print(paste("after fitering low-expressed genes, genes retained in all cluster is", 
              length(remain_genes)) )
  # those retained genes will be statistically tested to see whether differentially expressed 
  
  y <- calcNormFactors(y)
  # Calculate scaling factors to convert the raw sample/library sizes into normalized effective sample/library sizes. since raw counts are requested to enter 
  y$samples$condition <- factor(y$samples$condition, levels = c('control', 'H1N1', 'PRCV'))
  


  cluster <- as.factor(y$samples$cluster)

  limma:::plotMDS(cpm(y, log=TRUE), labels=y$samples$cluster ,
                  col = c(1:length(unique(cluster)))[cluster],
                  main = paste("DPI", selected_DPI, "MDS"))
  # print(p)
  
  #######################################################################
  
  
  clusts <- as.character(levels(factor(summed.filt$cluster)))
  de.results.any <- vector('list', length(clusts))
  
  names(de.results.any) <- clusts
  
  for(i in clusts)
  {
    de.results.any[[i]] <- pseudoBulkDGEmod_m2(clust = i, sce = summed.filt, 
                                               selected_comparison = Select_comparison,
                                               DPI = selected_DPI)
  }
  length(de.results.any)
  
  
  
  summary_df <- output_DE_to_files(de.results.any, flag = paste0(comparison_prefix, '_DPI', selected_DPI))
  melted_df <- DRAW_up_down_summary_barplot(summary_df, prefix = paste0(comparison_prefix, '_DPI', selected_DPI))
  return(summary_df)
}

# compare between 2 virus in same DPI
summary_df_DPI1_H1N1_PRCV <- DE_analysis(sce_all_filt, selected_DPI = 1,  Select_comparison= 3, comparison_prefix = "_H1N1_PRCV")
summary_df_DPI12_H1N1_PRCV <- DE_analysis(sce_all_filt, selected_DPI = 12, Select_comparison = 3, comparison_prefix = "_H1N1_PRCV")


summary_df_DPI1_ctl_PRCV <- DE_analysis(sce_all_filt, selected_DPI = 1, Select_comparison = 2, comparison_prefix = "_ctl_PRCV")
# DE_analysis(sce_all_filt, selected_DPI = 12, Select_comparison = 2, comparison_prefix = "_ctl_t2")


summary_df_DPI1_ctl_H1N1 <- DE_analysis(sce_all_filt, selected_DPI = 1, Select_comparison = 1, comparison_prefix = "_ctl_H1N1")
# DE_analysis(sce_all_filt, selected_DPI = 12, Select_comparison = 1, comparison_prefix = "_ctl_t1")

```



```{r use holm to correct for multiple test due to overlapping comparisons}
# three overlapping comparisons (control-H1N1, control-PRCV, H1N1-PRCV at 1 dpi)
# sampSet = "subset_innate_immune_cells"
# Create the table in R

mod_data_holm <- function(df_re, i, comparison_flag, write_table_flag = TRUE, power_flag = "power80",
                          DPI_flag = 1) 
{
  y_limit <- 1.0e-20
  # add a column of NAs
  df_re$diffexpressed <- "NO"
  # if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
  df_re$diffexpressed[df_re$logFC > 0.6 & df_re$p_holm_adjust < 0.05] <- "UP"
  # if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
  df_re$diffexpressed[df_re$logFC < -0.6 & df_re$p_holm_adjust < 0.05] <- "DOWN"
  
  table(df_re$diffexpressed)

  if(write_table_flag)
  {
    write.table(df_re, file=paste0('Results/DE_edgeR_', sampSet, '_', comparison_flag,'_DPI',DPI_flag, '_cluster_', i, '_holm.csv'),
              sep=',', quote=FALSE, row.names = FALSE)      
  }
   
  
  df_re_clean <- df_re[ df_re$diffexpressed != "NO", ]
  if(nrow(df_re_clean) > 0)
  {
    # dir.create('Results')
    if(write_table_flag)
    {
      write.table(df_re_clean, 
                  file=paste0('Results/DE_edgeR_', sampSet, '_', comparison_flag, '_DPI',DPI_flag, '_clean_cluster_', i, '_holm.csv'),
                  sep=',', quote=FALSE, row.names = FALSE)           
    }
  }
 
  
  df_re$delabel <- NA
  if(sum(df_re$diffexpressed != "NO") == 0)
  {
    
    # print(paste('no qualified DE genes observed in this cluster', i, ', no volcano plot'))
    return(df_re)
  }else
  {
    
    # print(paste('qualified DE genes observed in this cluster', i, ', show volcano plot'))
    df_re$delabel[df_re$diffexpressed != "NO"] <- df_re$symbol[df_re$diffexpressed != "NO"]
    # for row/gene who up/down, its delabel column will store the gene symbol again
    # for row/gene who is not DE, its delabel column is NA
    # Set p-value floor
    # df_re$adj.P.Val[df_re$adj.P.Val <= y_limit] <- y_limit 



    selected_logFC_threshold = df_logFC_threshold$log2FC_threshold[(df_logFC_threshold$DPI == DPI_flag & df_logFC_threshold$comparison == comparison_flag)] 

    p1 <- plot_volcano_holm(df_re, title=paste( "cluster", i, "comparison", comparison_flag, DPI_flag, power_flag),
                             selected_logFC_threshold)
        
    # p1 <- plot_volcano_holm(df_re, title=paste(cell_type, "cluster", i, "comparison", flag))
    # print(p1)
    
    png(filename = paste0('Results/DE_edgeR_', sampSet, '_', comparison_flag, DPI_flag, '_volcano_cluster', i, '_',  power_flag, '_holm.png') , height=1600, width=1600, res = 220)
    # when we set height=1080, width= higher, more gene symbols can shown on the plot since there are more space 
    print(p1)
    dev.off()    
  }
  
  return(df_re)
}


plot_volcano_holm <- function(df_re, title, logFC_threshold=2) 
{
  # code below use p values rather than p adjusted value to try
  # df_re$p_adjust <- df_re$PValue
    print(logFC_threshold)
  logFC_threshold <- as.numeric(logFC_threshold)

  ggplot(data=df_re, aes(x = logFC, y = -log10(p_holm_adjust), col=diffexpressed, label=delabel)) +
    geom_point() +
    geom_text_repel() +
    ggtitle(title, paste("DE genes holm, bespoke logFC_threshold")) +
    labs( x = "logFC", y = "-log10(p value)") +  # Customize axis labels here if needed
    geom_vline(xintercept=c(-0.6, 0.6), col="red") +
    geom_vline(xintercept=c(-logFC_threshold, logFC_threshold), col="green") +
    geom_hline(yintercept=-log10(0.05), col="red") +
    scale_color_manual(values=c("blue", "black", "red"), breaks=c("DOWN", "NO", "UP"), guide="none")+
    theme(axis.title.x = element_text(size = 15),                     # X-axis label font size
          axis.title.y = element_text(size = 15)   )

}  

holm_logFC_thresh_main <- function(sce_all_filt, df_logFC_threshold, power_flag = "power80") 
{
  # only DPI1 comparison need holm correct
  for(cluster_id in unique(sce_all_filt$cluster))
  {
    # correct for multiple test using holm, only DPI1 need adjust in bsp248
    # Files to read and write
    in_files <- c(
      paste0("Results/DE_edgeR_", sampSet,  "_ctl_PRCV_DPI1_cluster_", cluster_id, ".csv"),
      paste0("Results/DE_edgeR_", sampSet,  "_ctl_H1N1_DPI1_cluster_", cluster_id, ".csv"),
      paste0("Results/DE_edgeR_", sampSet,  "_H1N1_PRCV_DPI1_cluster_", cluster_id, ".csv") )
    
    
    if (any(!file.exists(in_files)) ) 
    {
      print(paste("File not found, skip: cluster", cluster_id))
      print(in_files[!file.exists(in_files)] )
      next 
    }
    
    comparison1 <- read.csv(in_files[1])
    comparison2 <- read.csv(in_files[2])
    comparison3 <- read.csv(in_files[3])
    
    comparison1$p_holm <- NA
    comparison2$p_holm <- NA
    comparison3$p_holm <- NA
    
    total_ens_id <- union(comparison1$ensemble_id, comparison2$ensemble_id)
    total_ens_id <- union(total_ens_id, comparison3$ensemble_id)
    
    
    for (selected_ens_id in total_ens_id) 
    {
      comparison1_p <- comparison1[comparison1$ensemble_id == selected_ens_id , 'PValue']
      comparison2_p <- comparison2[comparison2$ensemble_id == selected_ens_id , 'PValue']  
      comparison3_p <- comparison3[comparison3$ensemble_id == selected_ens_id , 'PValue']  
      
      p_holm <- p.adjust(c(comparison1_p, comparison2_p, comparison3_p), method = "holm")
      comparison1[comparison1$ensemble_id == selected_ens_id , 'p_holm'] <- p_holm[1]
      comparison2[comparison2$ensemble_id == selected_ens_id , 'p_holm']  <- p_holm[2]
      comparison3[comparison3$ensemble_id == selected_ens_id , 'p_holm'] <- p_holm[3]
      
    }

    
    comparison1$p_holm_adjust <- p.adjust(comparison1$p_holm, method='BH')
    comparison2$p_holm_adjust <- p.adjust(comparison2$p_holm, method='BH')    
    comparison3$p_holm_adjust <- p.adjust(comparison3$p_holm, method='BH')    
 
    # output table after holm correction, draw volcano plot and add bespoke logFC threshold, need to do those for DPI12 data too

    mod_data_holm(comparison1, cluster_id, "ctl_PRCV", write_table_flag = FALSE, power_flag, DPI_flag = 1)  
    
    mod_data_holm(comparison2, cluster_id, "ctl_H1N1", write_table_flag = FALSE, power_flag, DPI_flag = 1)  
    
    mod_data_holm(comparison3, cluster_id, "H1N1_PRCV", write_table_flag = FALSE, power_flag, DPI_flag = 1)  
    
  }
  
  
  #  DPI12 only add bespoke logFC threshold
  for(cluster_id in unique(sce_all_filt$cluster))
  {
    # correct for multiple test using holm, only DPI1 need adjust in bsp248
    # Files to read and write
    in_files <- paste0("Results/DE_edgeR_", sampSet,  "_H1N1_PRCV_DPI12_cluster_", cluster_id, ".csv")
    
    
    if (!file.exists(in_files)) 
    {
      print(paste("File not found, skip: cluster", cluster_id))
      print(in_files )
      next 
    }
    
    comparison1 <- read.csv(in_files)
    comparison1$p_holm_adjust <- comparison1$p_adjust
  
    #  draw volcano plot and add bespoke logFC threshold, need to do those for DPI12 data too
    mod_data_holm(comparison1, cluster_id, "H1N1_PRCV", write_table_flag = FALSE, power_flag, DPI_flag = 12)  
    
  }
  
}


# 
# # when desired power of 80%
# df_logFC_threshold <- data.frame(
#   DPI = c(12, 1, 1, 1),
#   log2FC_threshold = c(1.83, 2.75, 3.18, 3.62),
#   comparison = c("PRCV_H1N1", "PRCV_H1N1", "ctl_PRCV", "ctl_H1N1"))
# 
# holm_logFC_thresh_main(sce_all_filt, df_logFC_threshold, power_flag = "power80")
# 
# # when desired power of 50%
# df_logFC_threshold <- data.frame(
#   DPI = c(12, 1, 1, 1),
#   log2FC_threshold = c(1.195, 1.98, 2.35, 2.65),
#   comparison = c("PRCV_H1N1", "PRCV_H1N1", "ctl_PRCV", "ctl_H1N1")
# )
# 
# holm_logFC_thresh_main(sce_all_filt, df_logFC_threshold, power_flag = "power50")
# 
# # when desired power of 20%
# df_logFC_threshold <- data.frame(
#   DPI = c(12, 1, 1, 1),
#   log2FC_threshold = c(1.07, 1.43, 1.92, 2.23),
#   comparison = c("PRCV_H1N1", "PRCV_H1N1", "ctl_PRCV", "ctl_H1N1")
# )
# 
# holm_logFC_thresh_main(sce_all_filt, df_logFC_threshold, power_flag = "power20")

```




# check if up & down direction consistent VennDiagrams and shared-gene-tables
plot VennDiagrams of shared DE genes between main cell population (Macrophages, CD8 T cells, CD4 T cells, B cells)

```{r VennDiagrams functions }
# library(dplyr)
# 
# mod2_data <- function(df_re, i, cell_type)
# {
#   y_limit <- 1.0e-20
#   # add a column of NAs
#   df_re$diffexpressed <- "NO"
#   # if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
#   df_re$diffexpressed[df_re$logFC > 0.6 & df_re$p_adjust < 0.05] <- "UP"
#   # if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
#   df_re$diffexpressed[df_re$logFC < -0.6 & df_re$p_adjust < 0.05] <- "DOWN"
# 
#   df_re$delabel <- NA
#   if(sum(df_re$diffexpressed != "NO") == 0)
#   {
# 
#     print(paste('no qualified DE genes observed in this cluster', i, ', no volcano plot'))
#     return(df_re)
#   }else
#   {
# 
#     print(paste('qualified DE genes observed in this cluster', i, ', '))
#   # df_re$delabel[df_re$diffexpressed != "NO"] <- df_re$symbol[df_re$diffexpressed != "NO"]
#   # for row/gene who up/down, its delabel column will store the gene symbol again
#   # for row/gene who is not DE, its delabel column is NA
#   # Set p-value floor
#   # df_re$adj.P.Val[df_re$adj.P.Val <= y_limit] <- y_limit
# 
#   }
# 
#   return(df_re)
# }
# 
# 
# 
# modify_data_for_Venn <- function(de.results.any, flag= 'm2', selected_clusters )
# {
#   cls <- lapply(de.results.any, class)
# 
#   # check which cluster fail to have DE results and remove these clusters in de.results.any
#   # when a cluster have DE results, class of de.results.any[cluster] is data.frame
#   # when a cluster fail to have DE results, class of de.results.any[cluster] is character
#   if(length(which(cls == 'character')) > 0)
#   {
#     delete_cluster <- names(de.results.any)[which(cls %in% 'character')]
#     print(paste('delete_cluster',delete_cluster))
#     selected_clusters <- selected_clusters[! selected_clusters %in% delete_cluster]
#     de.results.any <- de.results.any[-which(cls %in% 'character')]
#   }
#   output_df_main_population <- NULL
#   print(selected_clusters)
#   for(i in selected_clusters)
#   {
#     print(i)
#     output_df <- NULL
# 
#     output_df$p_adjust <- p.adjust(de.results.any[[i]]$PValue, method='fdr')
#     output_df$symbol <- rowData(sce_all_filt)[rownames(de.results.any[[i]]), 'hgnc_symbol']
#     output_df$ensemble_id <- rownames(de.results.any[[i]])
#     output_df$logFC <- de.results.any[[i]]$logFC
#     output_df <- as.data.frame(output_df)
# 
#     output_df = output_df[order(output_df$p_adjust) , ]
# 
#     # change format GHSR, TNFSF10 into GHSR;TNFSF10 for those duplicates in our DE (edgeR) csv files.
#     my_vector <- output_df$symbol
# 
#     # Function to replace ', ' with ';' for those Ensembl gene ID who has duplicates IDs
#     replace_comma <- function(x) {
#       gsub(", ", ";", x)
#     }
# 
#     # Identify elements that match the pattern ', '
#     matching_indices <- grep(", ", my_vector)
# 
#     # Apply the function to identified elements and print each changed element
#     for (index in matching_indices) {
#       original_element <- my_vector[index]
#       changed_element <- replace_comma(original_element)
#       # print(paste(original_element, 'after modify', changed_element))
#     }
#     changed_vector <- sapply(my_vector, replace_comma)
#     output_df$symbol <- changed_vector
# 
# 
#     cell_type = df_cluster_number_name[df_cluster_number_name$cluster == i, 'named_cluster']
#     output_df <- mod2_data(output_df, i, cell_type)
#     output_df$cluster <- i
#     output_df_main_population <- rbind(output_df_main_population, output_df)
#     # add column diffexpressed to denote a DE is up or down expressed
#   }
# 
# 
#   output_df_main_population$symbol[output_df_main_population$symbol == ""] <- output_df_main_population$ensemble_id[output_df_main_population$symbol == ""]
#   return(output_df_main_population)
# 
# }
# 
# # Define the aggregation function
# aggregate_rows <- function(df) {
#   #   # this version is less strict
#   # 1.	If a gene is upregulated in one cluster and downregulated in another, I randomly assign it as either up or down in the main population.
#   # 2.	If a gene is upregulated in one of the clusters, it is considered upregulated in the main population.
#   # 2.	If a gene is downregulated in one of the clusters, it is considered downregulated in the main population.
#   # 3.	If a gene is not DE in all clusters, it is considered non DE in the main population.
# 
#   # Check if there are any rows with 'UP' or 'DOWN' in 'diffexpressed'
#   up_down_rows <- df %>% filter(diffexpressed %in% c("UP", "DOWN"))
# 
#   if (nrow(up_down_rows) > 0) {
#     if(nrow(up_down_rows) >= 2 && length(unique(up_down_rows$diffexpressed)) == 1)
#     {
#       # print('same direction')
#     }
# 
#     if(length(unique(up_down_rows$diffexpressed)) == 2)
#     {
#       print('opposite direction')
#       print(up_down_rows)
#     }
#     # Keep only one of the 'UP' or 'DOWN' rows
# 
#     result <- up_down_rows %>% sample_n(1)
#   } else {
#     # Randomly select one row if all rows have 'NO' in 'diffexpressed'
#     result <- df %>% sample_n(1)
#   }
# 
#   return(result)
# }
# 
# aggregate_up_rows <- function(df) {
#   #   # this version is less strict
#   # 1.	If a gene is upregulated in one cluster and downregulated in another, I randomly assign it as either up or down in the main population.
#   # 2.	If a gene is upregulated in one of the clusters, it is considered upregulated in the main population.
#   # 2.	If a gene is downregulated in one of the clusters, it is considered downregulated in the main population.
#   # 3.	If a gene is not DE in all clusters, it is considered non DE in the main population.
# 
#   # Check if there are any rows with 'UP' or 'DOWN' in 'diffexpressed'
#   up_down_rows <- df %>% filter(diffexpressed %in% c("UP"))
# 
#   if (nrow(up_down_rows) > 0) {
#     # Keep only one of the 'UP' or 'DOWN' rows
#     print(up_down_rows)
#     result <- up_down_rows %>% sample_n(1)
#   } else {
#     # Randomly select one row if all rows have 'NO' in 'diffexpressed'
#     result <- df %>% sample_n(1)
#   }
# 
#   return(result)
# }
# 
# aggregate_rows_strict <- function(df) {
#   # this version is more strict
#   # If a gene is upregulated in one cluster and downregulated in another, I now assign it as "not DE" in the main population.
# # 2.	If a gene is upregulated in all clusters, it is considered upregulated in the main population.
# # 3.	If a gene is downregulated in all clusters, it is considered downregulated in the main population.
# #
#   # Get unique values of 'diffexpressed'
#   unique_values <- unique(df$diffexpressed)
# 
#   # Case 1: All rows have 'DOWN' value
#   if (all(df$diffexpressed == "DOWN")) {
#     result <- df %>% sample_n(1)
# 
#     # Case 2: All rows have 'UP' value
#   } else if (all(df$diffexpressed == "UP")) {
#     result <- df %>% sample_n(1)
# 
#     # Case 3: All rows have 'NO' value
#   } else if (all(df$diffexpressed == "NO")) {
#     result <- df %>% sample_n(1)
# 
#     # Case 4: Rows have mixed 'UP', 'DOWN', 'NO' values
#   } else {
#     # Randomly choose one row but set 'diffexpressed' to 'NO'
#     result <- df %>% sample_n(1)
#     result$diffexpressed <- "NO"
#   }
# 
#   return(result)
# }
# 
# # to obtain the required Macrophages_Data input for VennDiagrams plot
# integrete_DE_main_population <- function(de.results.any, selected_clusters = c("5", "10", "12c"),
#                                          sampSet, main_population )
# {
#   output_df_main_population <- modify_data_for_Venn(de.results.any, flag= 'm2', selected_clusters = selected_clusters)
#   # a large dataframe that merges the DE results for several clusters
# 
#   # Apply the function to the data frame
#   aggregated_data <- output_df_main_population %>%
#     group_by(ensemble_id) %>%
#     do(aggregate_rows(.)) %>%
#     ungroup()
# 
#   write.csv(aggregated_data, file = paste0("Results/", sampSet, "_DE_edgeR_", main_population, ".csv"),
#             row.names = FALSE)
# 
#   return(aggregated_data)
# }
# 
# 
# # add UP DOWN NO lable to DE genes' results
# Create_Set <- function(Data){
#   DE_subset <- Data[Data$diffexpressed != "NO",]
#   # keep the DE genes are up-regulated or down-regulateds
# 
#   # remove any DE genes whose symbol column is NA or empty
#   DE_subset <- DE_subset[!is.na(DE_subset$symbol),]
#   DE_subset <- DE_subset[DE_subset$symbol != "",]
# 
#   # check if any DE genes whose symbol column start with ENSSSCG, and remove those
#   DE_subset <- DE_subset[!grepl("ENSSSCG", DE_subset$symbol, fixed = TRUE),]
#   # fixed logical. If TRUE, pattern is a string to be matched as is. Overrides all conflicting arguments.
#   Set <- DE_subset$diffexpressed == "UP"
#   # give up regulated genes value 1
#   Set[Set == TRUE] <- 1
#    # give non up-regulated genes value -1
#   Set[Set == FALSE] <- -1
#    # add gene name for each value
# 
#   Set <- as.integer(Set)
#   names(Set) <- DE_subset$symbol
#   return(Set)
# }
# # return a named vector of DE genes, with 1 denote up-regulated genes, -1  denote down-regulated genes
# 
# # prepare input data to plot Venn Diagrams
# # save.image(file = paste0('differential_edgeR_',sampSet,'_debug.RData'))

```


the cluster assignment is based on the slides 'Analysis of scRNA seq data on PRCV samples update 24.07.16.pptx'page 25, but delete cluster 11, 19c in run2. and add plasma cell clusters in B cells main population

```{r prepare the input data of VennDiagrams by combine multiple clusters into one main population for edgeR DE results}
# 
# # back_up = de.results.any
# head(as.data.frame(de.results.any[2]))
# #
# # typeof(de.results.any[1])
# 
# if(sampSet == 'run_1' )
# {
# 
#   CD8TCells_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("2", "9a", "14a"),
#                                                  sampSet, main_population = "CD8T")
#   # CD8TCells_Data[CD8TCells_Data$symbol == 'LGALS3',]
#   # table(CD8TCells_Data$diffexpressed)
# 
#   Macrophages_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("5", "10", "12c"),
#                                                  sampSet, main_population = "Macrophages")
#    # table(aggregated_data$diffexpressed)
# 
#   CD4TCells_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("3a", "14c", "17"),
#                                                  sampSet, main_population = "CD4T")
#   BCells_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("7", "13", '1', '4', '15'),
#                                                  sampSet, main_population = "BCells")
#   # BCells_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("7", "13"),
#                                                  # sampSet, main_population = "BCells")
#   # i add plasma cells into B Cells category since there are more DE in plasma cells, worth to have a look at their function
# }
# 
# if(sampSet == 'run_2' )
# {
#   Macrophages_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("3", "6", "20"),
#                                                  sampSet, main_population = "Macrophages")
#    # table(aggregated_data$diffexpressed)
#   CD4TCells_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("4a", "8", "9a"),
#                                                  sampSet, main_population = "CD4T")
#   # BCells_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("7", "17"),
#   #                                                sampSet, main_population = "BCells")
# 
#   CD8TCells_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c("1", "9c"),
#                                                  sampSet, main_population = "CD8T")
#   BCells_Data <- integrete_DE_main_population(de.results.any, selected_clusters = c('5', '10', '12', '13', '14',
#                                                                                     '18', "7", "17"), sampSet,
#                                               main_population = "BCells")
#   }
# 
# 
# 
# 
# 
# 


```


```{r plot VennDiagrams}



# # install.packages("remotes")
# # remotes::install_github("jmw86069/venndir");
# # The development version of venndir
# library(vctrs)
# library(venndir)
# 
# set_colours <- c('#FF999A', '#B2DF8A', '#33A02C', '#A6CEE3')
# theme_set(theme_minimal(base_size = 35))
# 
# 
# 
# 
# DE_genes_Set = list(Macrophages=Create_Set(Macrophages_Data),
#                  CD4_T_Cells=Create_Set(CD4TCells_Data),
#                  CD8_T_Cells=Create_Set(CD8TCells_Data),
#                  B_Cells=Create_Set(BCells_Data))
# 
# print(sapply(DE_genes_Set, length))
# if(sum(sapply(DE_genes_Set, length)) > 0)
# {
#   non_empty_DE_genes_Set_flag = TRUE
#   options("warn"=1)
# 
#   venndir(DE_genes_Set, poly_alpha=0.5, proportional=F, set_colors = set_colours, overlap_type="each", verbose = TRUE)
#   # + ggtitle(paste('edgeR', sampSet, 'Venn Diagrams'))
# 
#   library(grid)
#   png(filename=paste0("Results/", sampSet, "_edgeR_Venn_Diagram.png"), height=1080, width=1600)
#   options("warn"=-1)
#   venndir(DE_genes_Set, poly_alpha=0.5, proportional=F, set_colors = set_colours, font_cex = c(3,2,2), overlap_type="each")
#   # Add a title using grid package
#   grid.text(paste0(sampSet, " edgeR Venn Diagram"), x = 0.5, y = 0.95, gp = gpar(fontsize = 20))
# 
#   dev.off()
# }else
# {
#   print(paste('skip plot venn diagramm because there is no DE genes inside each main population'))
#   non_empty_DE_genes_Set_flag = FALSE
# }
# 
# 
# 

```

create tables of shared genes to accompany venn diagrams
```{r create shared-gene-tables}
# 
# library(dplyr)
# 
# #red for up, blue for down , purple for mix
# #since changed to append up, down, mixed, to use Excel's conditional formatting
# set_gene_colour <- function(n_v){
#   colour_vector <- c()
#   for(value in n_v){
#   if(length(value) == 0){
#     colour <- "NA"
#     next
#   }
#   if(any(1==value)){
#     colour <- "-Up"}
#   if(any(-1==value)){
#     colour <- "-Down"}
#   if(any(1==value) & any(-1==value)){
#     colour <- "-Mixed"}
#     colour_vector <- c(colour_vector, colour)
#   }
#   return(colour_vector)
# }
# 
# #create function to convert sets into lists of genes for each possible combination, colour by up or down regulation (purple for both)
# #convert 4 sets into 15 sets, numerical into vector
# set_to_df <- function(setlist)
#   {
#   #chain intersects to get genes in all four
#   All_4_names <- intersect(intersect(intersect(names(setlist$Macrophages),
#                                                names(setlist$CD4_T_Cells)),
#                                      names(setlist$CD8_T_Cells)),
#                            names(setlist$B_Cells))
#   # intersect(x, y) finds all rows in both x and y.
# 
#   All_4 <- bind_rows(setlist$Macrophages[names(setlist$Macrophages) %in% All_4_names],
#              setlist$CD4_T_Cells[names(setlist$CD4_T_Cells) %in% All_4_names],
#              setlist$CD8_T_Cells[names(setlist$CD8_T_Cells) %in% All_4_names],
#              setlist$B_Cells[names(setlist$B_Cells) %in% All_4_names])
# 
#   # retrieve DE genes that only exist in one of these 4 main cell populations
#   Macro_only_DE_genes  <-  setdiff(names(setlist$Macrophages),
#   c(names(setlist$CD4_T_Cells), names(setlist$CD8_T_Cells), names(setlist$B_Cells)))
#   Macro_only <- bind_rows(setlist$Macrophages[names(setlist$Macrophages) %in% Macro_only_DE_genes])
#   # setdiff(x, y) finds all rows in x that aren't in y.
#   # find DE genes in Macrophages but not in reaming 3 main cell populations then output the rows where such genes locate
# 
# 
#   CD4_only <- bind_rows(setlist$CD4_T_Cells[names(setlist$CD4_T_Cells) %in% setdiff(names(setlist$CD4_T_Cells), c(names(setlist$Macrophages), names(setlist$CD8_T_Cells), names(setlist$B_Cells)))])
# 
#   CD8_only <- bind_rows(setlist$CD8_T_Cells[names(setlist$CD8_T_Cells) %in% setdiff(names(setlist$CD8_T_Cells), c(names(setlist$Macrophages), names(setlist$CD4_T_Cells), names(setlist$B_Cells)))])
# 
#   B_only <- bind_rows(setlist$B_Cells[names(setlist$B_Cells) %in% setdiff(names(setlist$B_Cells), c(names(setlist$Macrophages), names(setlist$CD4_T_Cells), names(setlist$CD8_T_Cells)))])
# 
#   # retrieve DE genes that only exist in 2 of these 4 main cell populations, there are 6 different combinations
#   Macro_CD4_names <- setdiff(intersect(names(setlist$Macrophages), names(setlist$CD4_T_Cells)), c(names(setlist$CD8_T_Cells), names(setlist$B_Cells)))
# 
#   Macro_CD4 <- bind_rows(setlist$Macrophages[names(setlist$Macrophages) %in% Macro_CD4_names], setlist$CD4_T_Cells[names(setlist$CD4_T_Cells) %in% Macro_CD4_names])
# 
#   Macro_CD8_names <- setdiff(intersect(names(setlist$Macrophages), names(setlist$CD8_T_Cells)), c(names(setlist$CD4_T_Cells), names(setlist$B_Cells)))
# 
#   Macro_CD8 <- bind_rows(setlist$Macrophages[names(setlist$Macrophages) %in% Macro_CD8_names], setlist$CD8_T_Cells[names(setlist$CD8_T_Cells) %in% Macro_CD8_names])
#   #
#   Macro_B_names <- setdiff(intersect(names(setlist$Macrophages), names(setlist$B_Cells)), c(names(setlist$CD4_T_Cells), names(setlist$CD8_T_Cells)))
# 
#   Macro_B <- bind_rows(setlist$Macrophages[names(setlist$Macrophages) %in% Macro_B_names], setlist$B_Cells[names(setlist$B_Cells) %in% Macro_B_names])
#   #
#   CD4_CD8_names <- setdiff(intersect(names(setlist$CD4_T_Cells), names(setlist$CD8_T_Cells)), c(names(setlist$Macrophages), names(setlist$B_Cells)))
# 
#   CD4_CD8 <- bind_rows(setlist$CD8_T_Cells[names(setlist$CD8_T_Cells) %in% CD4_CD8_names], setlist$CD4_T_Cells[names(setlist$CD4_T_Cells) %in% CD4_CD8_names])
#   #
#   CD4_B_names <- setdiff(intersect(names(setlist$CD4_T_Cells), names(setlist$B_Cells)), c(names(setlist$Macrophages), names(setlist$CD8_T_Cells)))
#   CD4_B <- bind_rows(setlist$CD4_T_Cells[names(setlist$CD4_T_Cells) %in% CD4_B_names], setlist$B_Cells[names(setlist$B_Cells) %in% CD4_B_names])
#   #
#   CD8_B_names <-setdiff(intersect(names(setlist$B_Cells), names(setlist$CD8_T_Cells)), c(names(setlist$Macrophages), names(setlist$CD4_T_Cells)))
#   CD8_B <- bind_rows(setlist$CD8_T_Cells[names(setlist$CD8_T_Cells) %in% CD8_B_names],setlist$B_Cells[names(setlist$B_Cells) %in% CD8_B_names])
# 
# 
#   # # retrieve DE genes that only exist in 3 of these 4 main cell populations,
#   # there are 3 possible combinations
#   Macro_CD4_CD8_names <- setdiff(intersect(intersect(names(setlist$Macrophages), names(setlist$CD4_T_Cells)), names(setlist$CD8_T_Cells)), names(setlist$B_Cells))
#   Macro_CD4_CD8 <- bind_rows(setlist$Macrophages[names(setlist$Macrophages) %in% Macro_CD4_CD8_names],setlist$CD4_T_Cells[names(setlist$CD4_T_Cells) %in% Macro_CD4_CD8_names],setlist$CD8_T_Cells[names(setlist$CD8_T_Cells) %in% Macro_CD4_CD8_names])
#   #
#   Macro_CD4_B_names <- setdiff(intersect(intersect(names(setlist$Macrophages), names(setlist$CD4_T_Cells)), names(setlist$B_Cells)), names(setlist$CD8_T_Cells))
#   Macro_CD4_B <- bind_rows(setlist$Macrophages[names(setlist$Macrophages) %in% Macro_CD4_B_names],setlist$CD4_T_Cells[names(setlist$CD4_T_Cells) %in% Macro_CD4_B_names],setlist$B_Cells[names(setlist$B_Cells) %in% Macro_CD4_B_names])
#   #
#   Macro_CD8_B_names <- setdiff(intersect(intersect(names(setlist$Macrophages), names(setlist$CD8_T_Cells)), names(setlist$B_Cells)), names(setlist$CD4_T_Cells))
#   Macro_CD8_B <- bind_rows(setlist$Macrophages[names(setlist$Macrophages) %in% Macro_CD8_B_names],setlist$CD8_T_Cells[names(setlist$CD8_T_Cells) %in% Macro_CD8_B_names],setlist$B_Cells[names(setlist$B_Cells) %in% Macro_CD8_B_names])
#   #
#   CD4_CD8_B_names <- setdiff(intersect(intersect(names(setlist$CD4_T_Cells), names(setlist$CD8_T_Cells)), names(setlist$B_Cells)), names(setlist$Macrophages))
#   CD4_CD8_B <- bind_rows(setlist$CD4_T_Cells[names(setlist$CD4_T_Cells) %in% CD4_CD8_B_names],
#                          setlist$CD8_T_Cells[names(setlist$CD8_T_Cells) %in% CD4_CD8_B_names],
#                          setlist$B_Cells[names(setlist$B_Cells) %in% CD4_CD8_B_names])
# 
#   All_Combs <- list(Macro_only, CD4_only, CD8_only, B_only, Macro_CD4, Macro_CD8, Macro_B, CD4_CD8, CD4_B, CD8_B, Macro_CD4_CD8, Macro_CD4_B, Macro_CD8_B, CD4_CD8_B, All_4)
#   All_Names <- c("Macrophages Only", "CD4 T Cells Only", "CD8 T Cells Only", "B Cells Only", "Macrophages/CD4 T Cells",  "Macrophages/CD8 T Cells", "Macrophages/B Cells", "CD4 T Cells/CD8 T Cells", "CD4 T Cells/B Cells", "CD8 T Cells/B Cells", "Macrophages/CD4 T Cells/CD8 T Cells", "Macrophages/CD4 T Cells/B Cells", "Macrophages/CD8 T Cells/B Cells", "CD4 T Cells/CD8 T Cells/B Cells", "All 4")
#   # make vectors same length (fill with NA)
#   max.length <- max(sapply(All_Combs,length))
#   All_Combs2 <- lapply(All_Combs, "length<-", max.length)
#   Export_DF <- tibble(temp = 1:max.length)
#   # Extract names and make lists of colours for export table
#   for(i in 1:length(All_Combs2)){
#     n_v <- All_Combs2[[i]]
#     gene_vector <- names(n_v)
#     colour_vector <- set_gene_colour(n_v)
#     new_vector <- paste0(gene_vector, colour_vector)
#     Title <- All_Names[[i]]
#       if(length(new_vector)==0){
#       new_vector <- rep(NA, max.length)
#       }
#     Export_DF[[Title]] <- new_vector
#   }
#   Export_DF <- Export_DF[,names(Export_DF) != "temp"]
#   return(Export_DF)
# 
#   }
# if(non_empty_DE_genes_Set_flag)
# {
#   setlist = DE_genes_Set
#   write.csv(set_to_df(DE_genes_Set), paste0("Results/", sampSet, "edgeR_shared_genes_table.csv"))
# 
# }else
# {
#   print(paste('skip create shared-gene-tables because there is no DE genes inside each main population'))
# }
# 
# 
# 



```





# Wrapping up

Save everything

```{r save rds}

#
save.image(file = paste0('differential_edgeR_',sampSet,'.RData'))
sessionInfo()
# saveRDS(sce_all_filt, file=paste0('sce_',sampSet,'_cluster_filt.rds'))




```

