---
title: "uniform preprocessing script for scRNA data analysis"
author: "lily yang"
date: "`r Sys.Date()`"
output: html_document
---

# setup Environment, load package and data  
```{r, setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

work_dir <- Sys.getenv("work_dir")
bsp_num <- Sys.getenv("bsp_num")
samples_set <- Sys.getenv("sampSet")


knitr::opts_knit$set(root.dir = work_dir)

count_dir <- paste0('/mnt/lustre/RDS-ephemeral/bioinformatics/analysis/scrna/proj/', bsp_num, '/lily/stages/count')


current_run <- samples_set

reference_genome_version <- Sys.getenv("reference_genome_version")
ENTREZ_symbol_flag <- Sys.getenv("ENTREZ_symbol_flag")
run1_n <- as.numeric(Sys.getenv("run1_n"))
run2_n <- as.numeric(Sys.getenv("run2_n"))  
run3_n <- as.numeric(Sys.getenv("run3_n")) 
seperate_run_value <- as.numeric(Sys.getenv("seperate_run_value"))
n_niters <- as.numeric(Sys.getenv("n_niters"))
limit <- as.numeric(Sys.getenv("limit"))  
UMI_counts_threshold <- as.numeric(Sys.getenv("UMI_counts_threshold"))  
genes_detected_threshold <- as.numeric(Sys.getenv("genes_detected_threshold")) 
sum_counts_up <- as.numeric(Sys.getenv("sum_counts_up")) 

run1 <- 1:run1_n
run2 <- (run1_n+1 ):(run1_n+run2_n)
run3 <- (run1_n+run2_n+1):(run1_n+run2_n+run3_n)
sampNums <- (run1_n+run2_n+run3_n)
print(paste('work_dir', work_dir, 'samples_set', samples_set, bsp_num))              

## set to 100 because this is the default for emptyDrops used above

samps_id <- paste("sample", formatC(1:sampNums, width=2, flag="0"), sep="_")

# load packages
library(SingleCellExperiment)
library(DropletUtils)
library(viridisLite)     
library(ggplot2)
library(scran)
library(scater)
library(uwot)
```


# Intro, import and remove swapped droplets

This code will run an initial processing of the porcine Unsorted lung tissue 10x Genomics scRNAseq data generated at the Pirbright Institute in July and August 2024. bcl files were demultiplexed using `cellranger mkfastq` and aligned and counted using `cellranger count` (cellranger-7.1.0) using default parameters and Sus scrofa genome (genome assembly 11.1, Ensembl release 107).

Note saturation estimated by cellranger as follows:
need to do, write a sh script to extract this values for multiple samples 

We will read in raw counts matrices generated by cellranger to allow us to filter out empty droplets and removed barcode-swapped droplets using tools from DropletUtils. (cellranger filtered pipeline are too loose and can't identify all issues in the data, so it's better to be able to do extra checking and filtering ourselves.)

## 1 Reading in the data

In this data set, we sequenced 18 samples, over 2 sequencing runs. There is some concerns from Graham that run 2 did not run correctly.
experiment design: 
There are 2 time points which will we number as  dpi1 (1 days post-infection) and dpi12 (12 days post-infection).
There are 2 interventions PRCV and H1N1. There are 2 sets of controls: dpi1 (x2) and dpi12 (x2).


```{r import data in cluster and plot overview cell number, fig.width=10}


if(file.exists("preprocess1.RData"))
{
  load("preprocess1.RData")
}else
{
  setwd(count_dir)
  
  dirs <- paste0('muci_', samps_id, '_count' , '/outs/raw_feature_bc_matrix')
  sce_list_raw <- vector('list', length=sampNums)
  names(sce_list_raw)  <- samps_id

  for(i in 1:sampNums) {
      # Use try() to capture potential errors
      sce_result <- try(read10xCounts(samples = dirs[i], sample.names = samps_id[i], col.names = TRUE), silent = TRUE)
      
      # Check if an error occurred
      if (inherits(sce_result, "try-error")) {
        
          print(paste("Error, there is no fq files for sample", samps_id[i], ": demultiplex or count step fail"))
      } else {
          # If no error, store the result in the list
          sce_list_raw[[i]] <- sce_result
          print(paste('import fq files of sample', i, 'successful'))
          
      }
  }
     
  save.image(file = paste0(work_dir,'preprocess1.RData'))
}
# 
# lapply(sce_list_raw, dim)
# head(lapply(sce_list_raw, dim))
# head(sce_list_raw)
# Remove NULL elements from the list
sce_list_raw <- Filter(Negate(is.null), sce_list_raw)
sampNums <- length(sce_list_raw)



multi_list_bar_plot <- function(sce, my_flag, seperate_run = seperate_run_value)
{
  df <- as.data.frame(lapply(sce, dim))
  df <- as.data.frame(t(df))
  colnames(df) <- c("gene_number","droplet_number")
  df$sample <- as.factor(rownames(df))
  library(ggplot2)
  # Basic barplot
  p1 <- ggplot(data=df, aes(x=sample, y=droplet_number)) +
    geom_bar(stat="identity")+
    geom_text(aes(label=droplet_number), vjust=-0.3, size=3.5) +
    geom_vline(xintercept = seperate_run, color = "red", linetype = "dashed", size = 1)+
    ggtitle(paste0("droplet number summary ", my_flag))+
  theme(axis.text.x = element_text(angle = 90)) 
  # geom_vline(xintercept = 16.5, color = "blue", linetype = "dashed", size = 1)+
  print(p1)


}

multi_list_bar_plot(sce_list_raw,"before qc filtering", seperate_run = seperate_run_value )

```
figure 1. droplet number per sample before preprocess 


```{r extract sequencing saturation values and visulize in bar plot, fig.width=10}
setwd(count_dir)

dirs <- paste0('muci_', names(sce_list_raw), '_count' , '/outs/metrics_summary.csv')
seq_saturation <- 1:sampNums
cell_ranger_metrics <- vector('list', length=sampNums)

names(seq_saturation) <- names(sce_list_raw)
names(cell_ranger_metrics) <- names(sce_list_raw)

for (i in 1:sampNums) 
{
  cell_ranger_metrics[[i]] <- read.csv(dirs[i], header = TRUE)
  seq_saturation[i] <- cell_ranger_metrics[[i]]["Sequencing.Saturation"]
}

print(as.data.frame(seq_saturation))

# draw a bar plot 
seq_saturation <- t(as.data.frame(seq_saturation))
seq_saturation = as.data.frame(seq_saturation)
seq_saturation$sample <- rownames(seq_saturation)
colnames(seq_saturation)[1] <- 'value'

# Test if the quotes are part of the string
grepl('"', seq_saturation$value)
seq_saturation$number <- as.numeric(gsub('%', '', seq_saturation$value))/ 100

p <- ggplot(data=seq_saturation, aes(x=sample, y=number)) +
  geom_bar(stat="identity")+
  geom_text(aes(label=value), vjust=-0.2, size=3.5)+
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("sequencing saturation overview")
print(p)

```
the Sequencing.Saturation values are not consistent between run 1 and 2.

```{r check the correlation between sequencing saturation and sequencing metrics}

setwd(count_dir)

dirs <- paste0('muci_', names(sce_list_raw), '_count' , '/outs/metrics_summary.csv')
quality_metrics <- data.frame('Sequencing.Saturation' = 1:sampNums,
                             'Mean.Reads.per.Cell' = 1:sampNums,
                             'Median.Genes.per.Cell' = 1:sampNums,
                             'Median.UMI.Counts.per.Cell' = 1:sampNums)
cell_ranger_metrics <- vector('list', length=sampNums)

rownames(quality_metrics) <- names(sce_list_raw)
names(cell_ranger_metrics) <- names(sce_list_raw)

for (i in 1:sampNums) 
{
  cell_ranger_metrics[[i]] <- read.csv(dirs[i], header = TRUE)
  quality_metrics[i, ] <- cell_ranger_metrics[[i]][c("Sequencing.Saturation", 'Mean.Reads.per.Cell', 'Median.Genes.per.Cell', 'Median.UMI.Counts.per.Cell')]
}


sapply(quality_metrics, class)
quality_metrics$Sequencing.Saturation <- as.numeric(gsub('%', '', quality_metrics$Sequencing.Saturation))/ 100

# Load dplyr library
library(dplyr)

# Convert relevant columns from character to numeric
quality_metrics <- quality_metrics %>%
  mutate(across(everything(), ~ as.numeric(gsub(",", "", .))))

# View the updated dataframe
print(quality_metrics)

correlation_plot <- function(quality_metrics, y_axis = "Mean.Reads.per.Cell")
{
  library("ggpubr")
  p <-  ggscatter(quality_metrics, x = "Sequencing.Saturation", y = y_axis, 
            add = "reg.line", conf.int = TRUE, 
            cor.coef = TRUE, cor.method = "pearson",
            xlab = "Sequencing.Saturation", ylab = y_axis)
  print(p)  
}
correlation_plot(quality_metrics, "Mean.Reads.per.Cell")
correlation_plot(quality_metrics, "Median.Genes.per.Cell")
correlation_plot(quality_metrics, "Median.UMI.Counts.per.Cell")



```


## 2 Removing swapped droplets

Now we want to look for any droplets that have swapped cell barcodes due to use of patterned flow cell sequencing. 8 samples were sequenced in each run, and so swapping could occur anywhere within the sequencing run but not outside of it.

```{r Removing swapped transcript molecules}
sce_list <- sce_list_raw
if(file.exists("preprocess2.RData"))
{
  load("preprocess2.RData")
  }else
{
  getwd()
  setwd(count_dir)

  
  
  ## run 1
  fs <- paste0('muci_', samps_id[run1], '_count/outs/molecule_info.h5')
  run1_filter_swapped <- swappedDrops(samples=fs)

  ## run 2
  
  fs <- paste0('muci_', samps_id[run2], '_count/outs/molecule_info.h5')
  run2_filter_swapped <- swappedDrops(fs)
  
  if(bsp_num == "bsp_248") 
  {
  fs <- paste0('muci_', samps_id[run3], '_count/outs/molecule_info.h5')
  run3_filter_swapped <- swappedDrops(fs)    
  }



  # create a empty list to save the data that have Removed swapped drops
  cleaned <- list()
  
  for (i in 1:run1_n) {
      cleaned[[i]] <- run1_filter_swapped[[1]][[i]]
  }
  for (i in (run1_n+1) : ( run1_n + run2_n) ) {
      cleaned[[i]] <- run2_filter_swapped[[1]][[i-run1_n]]
  }
  if(bsp_num == "bsp_248") 
  {
  for (i in (run1_n + run2_n +1) : ( run1_n + run2_n + run3_n)) {
      cleaned[[i]] <- run3_filter_swapped[[1]][[i-run1_n-run2_n]]
  }    
  }
  
  # add -1 for each droplet barcode 
  for(i in 1:length(cleaned)){
    colnames(cleaned[[i]]) <- paste0(colnames(cleaned[[i]]), '-1')
  }
  
  # lapply(sce_list, dim)
  # save.image(file = paste0(work_dir,'preprocess2.RData'))
  save(cleaned, run1_filter_swapped, run2_filter_swapped, run3_filter_swapped, file = paste0(work_dir,'preprocess2.RData'))
  # evaluate how much droplets are removed in this step

}


for (i in 1:run1_n) 
{
  print(names(sce_list)[i])
  print(sum(colSums(run1_filter_swapped$cleaned[[i]]))/sum(colSums(counts(sce_list[[i]]))))
}

for (i in 1:run2_n) 
{
  print(names(sce_list)[i+run1_n])
  print(sum(colSums(run2_filter_swapped$cleaned[[i]]))/sum(colSums(counts(sce_list[[i+run1_n]]))))
}
if(bsp_num == "bsp_248") 
{
  for (i in 1:run3_n) 
  {
    print(names(sce_list)[i+run1_n+run2_n])
    print(sum(colSums(run3_filter_swapped$cleaned[[i]]))/sum(colSums(counts(sce_list[[i+run1_n+run2_n]]))))

  }

  
}

library(doParallel)
library(foreach)
# num_cores <- parallel::detectCores() - 1  # Use all cores except one
num_cores <- 8
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Replace original data with cleaned data that has removed swapped drops (parallelized)
foreach(i = 1:length(sce_list), .packages = c("SingleCellExperiment")) %dopar% {
  counts(sce_list[[i]])[, colnames(cleaned[[i]])] <- cleaned[[i]]
}


# replace original data with clean data that have Removed swapped drops
# for(i in 1:length(sce_list)){
#   counts(sce_list[[i]])[, colnames(cleaned[[i]])] <- cleaned[[i]]
#     # take long to run this step
# }
# head(counts(sce_list[[i]]))
# head(colnames(cleaned[[i]]))

rm(cleaned, run1_filter_swapped, run2_filter_swapped)  
if(bsp_num == "bsp_248") 
{
  rm(run3_filter_swapped)
}
```

Swapping was extremely minimal (< 0.01). We've removed the swapped reads anyways, but it was hardly consequential.

# QC of droplets
## 3 Filtering out empty droplets
### 3.1 check Count by rank plot
Now we filter empty droplets in each sample separately.
First we'll look at the Count by rank plot for UMI counts within each droplet, just to get a visual overview of whether anything is odd.
Each plot shows the total UMI count of each droplet against their rank. Droplets are ranked based on their total UMI count.

```{r produce Count by rank plot for UMI counts }


Count_by_rank_plot <- function(sce_list, calculate_brs = TRUE, use_before_filter_brs = FALSE, before_filter_brs = NULL)
{
  if(calculate_brs)
  {
    brs <- list()
    for (i in 1:length(sce_list)) 
    {
      brs_result <- try(barcodeRanks(counts(sce_list[[i]]), lower = 200, fit.bounds = c(300,35000)), silent = TRUE  )
      
      # Check if an error occurred
      if (inherits(brs_result, "try-error")) {
          print(paste("Error for sample", names(sce_list)[i]))
        
      } else {
          # If no error, store the result in the list
          brs[[i]] <- brs_result
          names(brs)[[i]] <- names(sce_list)[i]
          print(paste('Success sample', names(sce_list)[i]))
      }
      # brs[[i]] <- barcodeRanks(counts(sce_list[[i]]), lower = 200, fit.bounds = c(300,35000))
    }
  brs <- Filter(Negate(is.null), brs)
  
  # brs <- lapply(sce_list, function(x) {
  #   temp <- counts(x)
  #   print(x)
  #   return(barcodeRanks(temp, lower = 200, fit.bounds = c(300,35000)))} )
  # barcodeRanks function: Compute barcode rank statistics and identify the knee and inflection points on the total count curve. lower = 200, fit.bounds = c(300,35000) are used to alleviate the effect of noise when fit model
  # lower	
  # A numeric scalar specifying the lower bound on the total UMI count, at or below which all barcodes are assumed to correspond to empty droplets.
  # 
  # fit.bounds	
  # A numeric vector of length 2, specifying the lower and upper bounds on the total UMI count from which to obtain a section of the curve for spline fitting.
  }

  # par(mfrow=c(1, 2))
  for(i in 1:length(brs)){
    # returning only unique rank for plotting speed
    uniq <- !duplicated(brs[[i]]$rank)
    plot(brs[[i]]$rank[uniq], brs[[i]]$total[uniq], log="xy",
         xlab="Rank", ylab="Total UMI count", main=unique(names(brs)[[i]]), cex.lab=1.2)

    if(use_before_filter_brs)
    {
    # draw Count by rank plot using knee and inflection points derived from sce_list before filtering empty droplets
    abline(h=metadata(before_filter_brs[[i]])$inflection, col="darkgreen", lty=2)
    abline(h=metadata(before_filter_brs[[i]])$knee, col="dodgerblue", lty=2)
    legend("bottomleft",
           legend=c(paste("Knee",metadata(before_filter_brs[[i]])$knee),
                    paste("Inflection",metadata(before_filter_brs[[i]])$inflection)),
           col=c( "dodgerblue","darkgreen"), lty=2, cex=1)
    }else
    {
    # draw Count by rank plot using knee and inflection points derived from current sce_list
    abline(h=metadata(brs[[i]])$inflection, col="darkgreen", lty=2)
    abline(h=metadata(brs[[i]])$knee, col="dodgerblue", lty=2)
    legend("bottomleft",
           legend=c(paste("Knee",metadata(brs[[i]])$knee),
                    paste("Inflection",metadata(brs[[i]])$inflection)),
           col=c( "dodgerblue","darkgreen"), lty=2, cex=1)
    }

  }

  # Reset to default plotting layout
  par(mfrow=c(1, 1))
  return(brs)

}


# use before_filter_brs to store the value of knee and inflection points derived from sce_list before filtering  empty droplets
before_filter_brs <- Count_by_rank_plot(sce_list, calculate_brs = TRUE)

# sce_list[[13]] <- NULL
# sampNums <- length(sce_list)

```

figure 2. Count by rank plot before filtering empty droplets

an ideal knee point should locate at the curve where its slope significantly shifts and become steep, ideal y axis value of this point should over 1000 since total UMI count lower than 1000 might fail in cluster step


an ideal inflection point should locate at the middle of a steep slope of the curve, ideal value over 500

if under default parameter, due to the interfere of outlier droplets with abnormally high total UMI, estimated knee points of such Sample will become too early.

suspicious samples:
5, 




### 3.2 background knowledge supplement
i think we are less interested in these those kinds of droplets with too low total UMI count.
so i prefer set strict qc metrics first to filter out droplets lower than inflection points.

develop later: 
how much of the filtered droplet can still be used, distinguish droplets have ambient RNAs , droplets have transcription inactive intact droplets.



### 3.3  validate assumption, droplets containing fewer than 100 UMIs do not contain any real droplets

we retrieve droplets whose total UMI count between 0-100, check their P values (possibility of being non-empty droplets) using 'emptyDrops' function, and plot histogram of the predicted P-Value.
default assumption is droplets containing fewer than 100 UMIs do not contain any real droplets, therefore, in theory P Value of these droplets should near 1, and should not peak around 0. 




```{r calculate P values for droplets containing fewer than 100 UMIs and plot histogram of their P values to check how accurate the emptyDrops estimate empty droplets}
if(file.exists("preprocess3.3.RData"))
{
  load("preprocess3.3.RData")
}else
{
  e.out <- vector('list', length=length(sce_list))
  all.out <- vector('list', length=length(sce_list))
  for(i in 1:length(sce_list)){
    set.seed(100)
    e.out[[i]] <- emptyDrops(counts(sce_list[[i]]), niters = n_niters,  lower = 100,
                             test.ambient = FALSE)
    # print(n_niters)
    # Distinguish between droplets containing cell and ambient RNA in a droplet-based single-cell RNA sequencing experiment.


    set.seed(100)
    # obtain all droplets' possibility of being non-empty droplets

    all.out[[i]] <- emptyDrops(counts(sce_list[[i]]), lower=limit, test.ambient=TRUE,  
                               niters = n_niters)
    # print(limit)
    # test.ambient=TRUE, A logical scalar indicating whether results should be returned for barcodes with totals less than or equal to 'lower'.
    # lower=limit, A numeric scalar specifying the lower bound on the total UMI count,  at or below which all barcodes are assumed to correspond to empty droplets.

  # If test.ambient=TRUE, p-values will be reported for all barcodes with positive total counts, including those not greater than lower, which otherwise were reported as non-NA. This is occasionally useful for diagnostics to ensure that the p-values are well-calibrated for barcodes corresponding to (presumably) empty droplets.


  # Note that, when setting test.ambient=TRUE in emptyDrops, barcodes less than or equal to lower will still have NA values in FDR. Such barcodes are still explicitly ignored in the correction as these are considered to be uninteresting.
  }
  
  save.image("preprocess3.3.RData")

}

for(i in 1:length(sce_list))
{
  
  # print(summary(e.out[[i]]$FDR <= 0.001))
  # FDR <= 0.001 indicate droplets containing cell

  # print(table(Sig=e.out[[i]]$FDR <= 0.001, Limited=e.out[[i]]$Limited))
  # niters parameter can be increased in the future if Limited=TRUE and Sig=FALSE  are very high
  table1 <- table(Sig=e.out[[i]]$FDR <= 0.001, Limited=e.out[[i]]$Limited)
  print(table1)
  if(table1[3] == 0)
  {
    print(paste(names(sce_list)[i], 'value of niters parameter is sufficient'))
    # print(paste('number of droplets are estimated contain cell', summary(e.out[[i]]$FDR <= 0.001)[3]))
  }else
  {
    print(paste(names(sce_list)[i], 'insufficient value of niters parameter, inaccurate estimation '))
    # print(paste('number of droplets are estimated contain cell', summary(e.out[[i]]$FDR <= 0.001)[3]))    
  }

  hist(all.out[[i]]$PValue[all.out[[i]]$Total <= limit & all.out[[i]]$Total > 0],
     xlab="P-value", main=unique(sce_list[[i]]$Sample), col="grey80")
  # PValue near 0 means this droplet is inferred to be cell contained, near 1 means this droplet is inferred to be ambient RNA, don't contain cell

}

```

figure 3. P values histogram of droplets containing fewer than 100 UMIs


if Limited=TRUE, it indicates the p-value for a droplet is limited by the number of iterations (niters),  it indicate the true p-value could potentially be lower than computed if more iterations were performed.

"PValue" refers to the statistical p-value obtained from testing each barcode (representing a droplet) for significant deviations from the ambient RNA profile

If any barcodes have Limited=TRUE but does not reject the null hypothesis(Sig=FALSE ), it suggests that niters should be increased.

all droplets with PValue=TRUE are estimated as non-empty droplets. 

Number of iterations for all samples looks appropriate now, as all droplets whose value of Limited=TRUE and Sig=FALSE are always 0.

ideal niters parameter should make Limited=TRUE and Sig=FALSE are 0 for each sample

the code will automatically check whether Limited=TRUE and Sig=FALSE are 0 for each sample, and report fail samples.


in the histogram, P-values of droplets that are assumed to ambient reads sometimes dip near 1 but have no peaks near 0, coincide with the assumption that droplets containing fewer than 100 UMIs do not contain any real droplets.
therefore, samples have ideal quality plots there are: 





### 3.4 filter the estimated empty droplets and plot total UMI histogram
Thus we can filter the droplets

```{r filter the empty droplets based on FDR value and check Count by rank plot after filter the empty droplets }

sce_list_before_filter_empty_droplets <- sce_list
if(file.exists("preprocess3.4.RData"))
{
  load("preprocess3.4.RData")
}else
{

  for(i in 1:length(sce_list))
    {
    sce_list[[i]] <- sce_list[[i]][,which(e.out[[i]]$FDR <= 0.001)]
    # sce_list[[i]] <- sce_list[[i]][,which(e.out[[i]]$Total >= metadata(before_filter_brs[[i]])$inflection )]
    }
    save.image(file = "preprocess3.4.RData")
  }



# lapply(sce_list, dim)
multi_list_bar_plot(sce_list,"after filter estimated empty droplets", seperate_run = seperate_run_value )



brs_after_filtering <- Count_by_rank_plot(sce_list, calculate_brs = TRUE, use_before_filter_brs = TRUE, before_filter_brs)

```
figure 4. Count by rank plot after filtering empty droplets

So the apparent odd slope in Sample 5 has dramatically large number of droplets that have quiet low total UMI (<600), based on FDR value, these droplets are still estimated as cells and been kept. these low total UMI droplets might be an issue, and we will revisit this as we continue to process and QC. maybe later we will filter all droplets that have quiet low total UMI (<600).

if these droplets are real outliers, them tend to cluster together, with abnormal gene expression, should be observed in UMAP plots.



let's have a look at the distribution total UMI, and filter abnormally high total UMI outlier droplets by set the threshold

### 3.5 filter droplets with abnormally high or low total UMI
```{r draw total UMI per droplet histogram before and after filtering droplets with abnormally high or low total UMI }
library("scuttle")
sce_list <- lapply(sce_list, function(x){
  stats <- perCellQCMetrics(x)
  colData(x)$sum_counts <- stats$sum # total UMI per droplet
  return(x)
})

plot_total_UMI_distribution <- function(sce_list)
{
   # Set the plotting area to 1 row and 3 columns
# par(mfrow=c(1, 3))

  lapply(sce_list, function(x){

    hist(x$sum_counts, main=paste0(unique(x$Sample),', total UMI per cell'), xlab='total UMI counts of a cell', ylab = 'number of droplets',
         xlim = c(0,10000), breaks = seq(0, 90000, by=1000)) # breaks = seq(0, 10000, by=1000) need to cover total UMI per cell for all cells 
  abline(v=500, col="red", lwd=1) # Add a vertical red line at x=2000
  abline(v=3000, col="blue", lwd=1)
  # a droplet contain an intact cell always have total UMI counts 2000-5000
  legend("topright", legend=c("500", "3000"),
         col=c("red", "blue"), lty=2, cex=1.2)
# better to combine plots 
})

# Reset to default plotting layout
par(mfrow=c(1, 1))
}

# print('total UMI per droplet histogram before filtering droplets with abnormally high or low total UMI')

# plot_total_UMI_distribution(sce_list)



# remove droplets with extreme high or low Total UMI
 for(i in 1:length(sce_list)){
   # sce_list[[i]]$UMI_counts_outlier <- isOutlier(sce_list[[i]]$sum_counts, type='higher', nmads = 3)
   # since large number of droplets that have quiet low total UMI (<600) in sample 17-18 will make other droplets that have normal total UMI (2000-5000) become ouliers, so there we don't use isOutlier to filter abnormally high total UMI droplets (>50,000)

   # sce_list[[i]]$UMI_counts_outlier <- (sce_list[[i]]$sum_counts >= 50000) | (sce_list[[i]]$sum_counts <= 500)
   sce_list[[i]]$UMI_counts_outlier <- (sce_list[[i]]$sum_counts >= UMI_counts_threshold) 
   # maybe will filter sum_counts <= 800 in the future. sample 18 have too many droplets around 800 total UMI, need to check if them can cluster successfully
   print(paste('TRUE = the number of droplets that will be filtered due to abnormally high UMI counts', UMI_counts_threshold, 'in sample', i ))
   
   print(table(sce_list[[i]]$UMI_counts_outlier))

   sce_list[[i]] <- sce_list[[i]][,!sce_list[[i]]$UMI_counts_outlier]
 }




# print('total UMI per droplet histogram after filtering droplets with abnormally high or low total UMI')
# 
# plot_total_UMI_distribution(sce_list)

# lapply(sce_list, dim)
multi_list_bar_plot(sce_list,"after filter droplets with abnormally high total UMI", seperate_run = seperate_run_value )

```

figure 5. plot total UMI per droplet histogram before and after filtering


issue of most droplets have total UMI lower than 2000, too few droplets have total UMI 2000-5000.
severe: samples
moderate: samples

good quality samples: 

need to check:
shall we filter sum_counts >= 20,000 in the future?

we try produce Count by rank plot again using previously predicted inflation and knee points, them fits the new filtered clean data better in most samples. and we also calculate the percentage of droplets been filtered in this step and the number of droplets left after this filtering step.

```{r plot number of droplets per sample, fig.width=10 }


# Count_by_rank_plot(sce_list, calculate_brs = TRUE, use_before_filter_brs = FALSE)


print('the proportion of droplets remained after filtering empty and high UMI counts droplets')
for(i in 1:length(sce_list))
{
  print(paste(names(sce_list)[i], dim(sce_list[[i]])[2] / dim(sce_list_raw[[i]])[2]))
}
```

figure 6. plot number of droplets per sample

## 4 Removing ambient RNA

While removing ambient RNA is not always necessary, the data we are dealing with comes from different immunological settings of strain 135 infected versus strain ISU-1 infected. As such, there will be some different cell states in each animal and these will contribute to background RNA. The worry is that it might appear that *all* droplets are expressing e.g. specific inflammatory genes as opposed to the main cell type of that cell in their normal condition. Thus, we will remove ambient RNA from each sample. basically, this step change the total UMI counts in each sample by reduce UMI counts of some genes within each droplet.

We start by running a quick clustering.

```{r quick clustering}
# BiocManager::install("scran")
library(scran)
library(scater)
library(uwot)

if(file.exists("preprocess4.RData"))
{
  load("preprocess4.RData")
}else
{
  clust <- quickCluster(sce_list[[1]])
  # library(igraph)
  for(i in 1:length(sce_list))
    {
    set.seed(100)
    clust <- quickCluster(sce_list[[i]])
    colLabels(sce_list[[i]]) <- clust
  }
  save.image('preprocess4.RData')
  }

```

Next we will use this rough clustering to remove ambient RNA.

from documentation of function 'removeAmbience'
Estimate and remove the ambient profile from a count matrix, given pre-existing groupings of similar droplets. This function is largely intended for plot beautification rather than real analysis.

This function will aggregate counts from each group of related droplets into an average profile. For each group, we estimate the contribution of the ambient profile and subtract it from the average. By default, this is done with ambientContribMaximum, but if enough is known about the biological system, users can specify features by using ambientContribNegative instead.

We then perform quantile-quantile mapping of counts in y from the old to new averages. This approach preserves the mean-variance relationship and improves the precision of estimate of the ambient contribution, but relies on a sensible grouping of similar droplets, e.g., unsupervised clusters or cell type annotations. As such, this function is best used at the end of the analysis to clean up expression

from documentation of function 'removeAmbience', it says this function is largely intended for plot beautification after clustering and cell type annotations finished. rather than real analysis. so i am not sure if it's suitable there to do this step.
i don't run this step now.


```{r enter rough clustering to remove ambient RNA using removeAmbience function of DropletUtils package}


# sce_list_raw <- sce_list
# for(i in 1:length(sce_list)){
#   amb <- metadata(e.out[[i]])$ambient[,1]
#   # e.out is the output of function emptyDrops in last step
#   # amb save each gene's contribution as ambient in one sample
#
#   amb_full <- rep(0, nrow((sce_list[[i]])))
#   # amb_full has the length of 33028, the number of all genes measured across all sample
#   names(amb_full) <- rownames(sce_list[[i]])
#   amb_full[names(amb)] <- amb
#   # assign value to some of the elements of amb_full
#   assays(sce_list[[i]], withDimnames=FALSE)$counts <- removeAmbience(counts(sce_list[[i]]),
#                    ambient=amb_full, groups=sce_list[[i]]$label)
#   # removeAmbience function there is from DropletUtils library
# }
# head(amb)
# head(amb_full)
# rm(e.out, amb_full, amb)
#
# for(i in 1:length(sce_list))
# {
#   print(sum(assays(sce_list[[i]], withDimnames=FALSE)$counts) / sum(assays(sce_list_raw[[i]], withDimnames=FALSE)$counts))
#   # cell number doesn't change for each sample, the total UMI of each sample might change
# }
#
# rm(sce_list_raw)


```


```{r remove ambient RNA using SoupX}
library('SoupX')
# Extract raw counts and cell metadata from SingleCellExperiment objects
sce_list_before_remove_ambient_rna <- sce_list


for(i in 1:length(sce_list))
{
  # print(paste('sample', i))
  raw_counts <- counts(sce_list_before_filter_empty_droplets[[i]])
  filtered_counts <- counts(sce_list[[i]])
  # Create a SoupChannel object
  sc <- SoupChannel(raw_counts, filtered_counts)

  # Add cluster information to the SoupChannel object
  # colData(sce_list[[i]])$label
  # unique(colData(sce_list[[i]])$label)
  # colLabels(sce_list[[i]])

  sc = setClusters(sc, setNames(colLabels(sce_list[[i]]), colnames(sce_list[[i]])))
  print(paste('autoEstCont sample', names(sce_list)[i]))
  # Estimate soup profile and contamination fraction
  # pdf(file = "SoupX_output.pdf")

  sc <- try(autoEstCont(sc, doPlot = FALSE), silent = FALSE)
  if (inherits(sc, "try-error"))
  {
    print(paste("Error, autoEstCont step fail", names(sce_list)[i]))
  }else{
  # Output the clean data which removed the predicted ambient reads
  out <- adjustCounts(sc)
  counts(sce_list[[i]]) <- out
  }
  # dev.off()
  # print(paste('test sample', i))
}

# Check the change of adjusted counts
print('the proportion of total read counts remained after remove ambient RNA')

for(i in 1:length(sce_list))
{
  proportion <-  sum(assays(sce_list[[i]], withDimnames=FALSE)$counts) / sum(assays(sce_list_before_remove_ambient_rna[[i]], withDimnames=FALSE)$counts)
  print(paste('proportion of sample', names(sce_list)[i], ':', proportion))
  # cell number doesn't change for each sample, the total UMI of each sample might change
}


```


## 5 Cellranger filtering comparison

For comparison, we will also read in the cellranger filtered output to see how results compare in terms of number of droplets.

```{r load cellranger filtered output and compare with our filtered output}
library(SingleCellExperiment)
library(DropletUtils)

if(file.exists("preprocess5.RData"))
{
  load("preprocess5.RData")
}else
{
  setwd(count_dir)
  dirs <- paste0('muci_', names(sce_list), '_count/outs/filtered_feature_bc_matrix')
  sce_list_cr <- vector('list', length=sampNums)
  names(sce_list_cr) <- names(sce_list)

  for(i in 1:sampNums)
  {

    try(sce_result <- read10xCounts(samples = dirs[i], sample.names = samps_id[i], col.names = TRUE), silent = TRUE)

      # Check if an error occurred
      if (inherits(sce_result, "try-error")) {

          print(paste("Error, there is no filtered_feature_bc_matrix files for sample", samps_id[i], ": demultiplex or count step fail"))
      } else
        {
          # If no error, store the result in the list
          sce_list_cr[[i]] <- sce_result
          print(paste('load filtered data of ',dirs[i], 'successful'))
        }


  }
  sce_list_cr <- Filter(Negate(is.null), sce_list_cr)
  print(names(sce_list_cr))



  save.image(file = paste0(work_dir,"preprocess5.RData"))
}


multi_list_bar_plot(sce_list,"after  filtering extreme high or low total UMI droplets", seperate_run = seperate_run_value )
multi_list_bar_plot(sce_list_cr,"after cellranger filtering", seperate_run = seperate_run_value )

# lapply(sce_list, dim)
# lapply(sce_list_cr, dim)
print('compare the droplets number left after filtering')
# compare the droplets number left after filtering
for(i in 1:length(sce_list))
{
  # print((dim(sce_list[[i]])[2] - dim(sce_list_cr[[i]])[2])/
  #         ((dim(sce_list[[i]])[2] + dim(sce_list_cr[[i]])[2])/2)*100)

  print(paste(names(sce_list)[[i]], dim(sce_list[[i]])[2] / dim(sce_list_cr[[i]])[2]))
}

rm(sce_list_cr)
```
compare remained number of droplets within each sample across cellranger filtered output and our filtered output
sample_05, sample_07, sample_10, sample_14
has quiet big difference between cellranger filtered output and our filtered output




## 6 Removing residual low quality droplets

### 6.1 filter zero counts genes, annotate genes names
Here we will look for outlying droplets in terms of detected gene number and proportion of mitochondrial reads.

In order to look at specific types of genes, we need to annotate the dataset. To make this easier, we'll first remove genes with no counts in every sample

```{r filter zero counts genes and map genes symbols original}
if(bsp_num == "bsp_247")
{
  library(org.Gg.eg.db)
   
}
if(bsp_num == "bsp_248") 
{
  library(org.Ss.eg.db)
}
library(dplyr)

detected <- lapply(sce_list, function(x){rowSums(counts(x)) > 0})

# detected is a list with 20 elements, each element contain all genes with true or false value
names(detected) <- paste0('Sample', as.character(c(1:sampNums )))
detected <- data.frame(detected)
# 33028 row    20 columns

# Remove genes with zero counts in all droplets across all samples
undetected <- which(rowSums(detected) == 0)
# undetected record such gene's id and position
print(paste('number of genes with zero counts in all sample', length(undetected)))

# add gene annotation to ensemble id
if(file.exists("preprocess6.1.RData"))
{
  load("preprocess6.1.RData")
}else
{
  # only delete about 2000 genes out of 30000+ genes
  sce_list <- lapply(sce_list, function(x) {return(x[-undetected,])})
  rm(detected, undetected)
  # lapply(sce_list, dim)
  if(reference_genome_version == "113")
  {
    if(file.exists("references/gene_annotation_map_full_v113.txt") )
    {
      map <- read.table("references/gene_annotation_map_full_v113.txt", header = TRUE)
      colnames(map)[1:5] <- c("chromosome_name", "start_position", "end_position", "ensembl_gene_id",
                              "external_gene_name")
      
      unique_map <- map %>%
        group_by(ensembl_gene_id) %>%
        summarise(
          external_gene_name = paste(unique(external_gene_name[!is.na(external_gene_name)]), collapse = ";"),
          chromosome_name = paste(unique(chromosome_name[!is.na(chromosome_name)]), collapse = ";"),
          start_position = paste(unique(start_position[!is.na(start_position)]), collapse = ";"),
          end_position = paste(unique(end_position[!is.na(end_position)]), collapse = ";")
        ) %>%
        ungroup()
      # head(unique_map)
      
      unique_map <- as.data.frame(unique_map)
      rownames(unique_map) <- unique_map$ensembl_gene_id
      map <- unique_map
      # nrow(unique_map)

    }else
    {
      print('please create ensemblif symbol map files v113 first')
    }    
  }else if(reference_genome_version == "108")
  {
    if(file.exists("references/ensembl_map_v108.csv") && file.exists('references/ensembl_ENTREZ_symbol_map_v108.csv'))
    {
      map <- read.csv("references/ensembl_map_v108.csv", row.names = 1)
      # head(map)
    }else
    {
      print('please create ensemblif symbol map files v108 first')
    }    
  }


  if(bsp_num == "bsp_247")
  {
      # Combine rows with the same 'ensembl_gene_id' and combine their 'chromosome_name', 'start_position', and 'end_position'
    unique_map <- map %>%
      group_by(ensembl_gene_id) %>%
      summarise(
        external_gene_name = paste(unique(external_gene_name[!is.na(external_gene_name)]), collapse = ";"),
        chromosome_name = paste(unique(chromosome_name[!is.na(chromosome_name)]), collapse = ";"),
        start_position = paste(unique(start_position[!is.na(start_position)]), collapse = ";"),
        end_position = paste(unique(end_position[!is.na(end_position)]), collapse = ";")
      ) %>%
      ungroup()
    unique_map <- as.data.frame(unique_map)
    rownames(unique_map) <- unique_map$ensembl_gene_id
    # head(unique_map)
    # dim(unique_map)
      # Update 'sce_list' with the unique 'ensembl_gene_id' data
    sce_list <- lapply(sce_list, function(x) {
      rowData(x)$gene_symbol_ensembl <- unique_map[rownames(x), 'external_gene_name']
      rowData(x)$chr <- unique_map[rownames(x), 'chromosome_name']
      rowData(x)$start <- unique_map[rownames(x), 'start_position']
      rowData(x)$end <- unique_map[rownames(x), 'end_position']
      return(x)
    })
  
     
  }

  if(bsp_num == "bsp_248") 
  {
    sce_list <- lapply(sce_list, function(x){
      rowData(x)$gene_symbol_ensembl <- map[rownames(x), 'hgnc_symbol']
      rowData(x)$chr <- map[rownames(x), 'chromosome_name']
      return(x)})
      # rowData(x)$start <- map[rownames(x), 'start_position']
      # rowData(x)$end <- map[rownames(x), 'end_position']    
    # v113 start too long 3227816;3242750;3242787;3242120;3241473;3237694;3234540;3231571;3230771;3229194;3229095;3228687;3228199 need biomart include v113 to map the sole start position
  }


  head(rowData(sce_list[[1]]))
  save.image(file = paste0(work_dir,"preprocess6.1.RData"))

}
```


```{r examine map quality of original method}
print(paste('number of unique gene_symbol_ensembl', length(unique(rowData(sce_list[[1]])$gene_symbol_ensembl))))
print(paste('total number of contained gene_symbol_ensembl', length(rowData(sce_list[[1]])$gene_symbol_ensembl)))
# check how much gene names are assigned successfully

print(paste('number of unique ensembl_gene_id', length(unique(rownames(sce_list[[1]])))))
print(paste('total number of ensembl_gene_id', length(rownames(sce_list[[1]]))))
# check how much unique gene id exist



```

```{r second map genes symbols method and examine map quality of second method}}
if(ENTREZ_symbol_flag == "yes")
{
  ################################## merge symbols which have same ensembl id
  if(file.exists("references/ensembl_ENTREZ_symbol_map_v108.csv"))
  {
    hgnc_symbol_new <- read.csv("references/ensembl_ENTREZ_symbol_map_v108.csv", row.names = 1)
  }else
  {
    print("fail to find references/ensembl_ENTREZ_symbol_map_v108.csv")
  }
  
  # head(hgnc_symbol_new)
  ################################## add new map symbols in sce_list
  sce_list <- lapply(sce_list, function(x){
    rowData(x)$ENTREZ_symbol <- hgnc_symbol_new[rownames(x), 'ENTREZ_symbol']
    return(x)})

}

 
```


### 6.2 Identify and plot mitochondrial genes UMI counts
```{r  Identify and plot mitochondrial genes counts}
# Identify and plot mitochondrial genes counts

for (i in 1:length(sce_list))
{

  print(paste(length(which(rowData(sce_list[[i]])$chr == 'MT')), 'mitochondrial genes exist in ', names(sce_list)[i]  ))
  # obtain how many mitochondrial genes exist in sample 1

  print(paste('the total count of each mitochondrial genes across all droplets of', names(sce_list)[i]))
  print(rowSums(counts(sce_list[[i]])[which(rowData(sce_list[[i]])$chr == 'MT'),]))
  # obtain the total count of each mitochondrial genes across all droplets
  hist(colSums(counts(sce_list[[i]])[which(rowData(sce_list[[i]])$chr == 'MT'),]),
       main= paste0('Mitochondrial UMI counts sample',i),
       xlab='UMI counts of mitochondrial transcripts per cell',
       ylab= 'Number of droplets', breaks=40)
  # obtain the total count of all mitochondrial genes contained in a droplets


}

```



all samples have 13 annotated mitochondrial genes detected here, but the UMI count for some of mitochondrial genes are much higher than other mitochondrial genes

most samples have most of their droplets contain low UMI count for mitochondrial genes, so most samples in this aspect should be fine.


future work: can check function or metabolic pathway of those highly expressed mitochondrial genes, maybe indicate droplets undergoing stress or in the process of dying


Now we can check various QC plots and metrics (total UMI per droplet, total genes detected per droplet, proportion of UMI counts of mito genes per droplet) of every sample.

### 6.3 check and plot various QC paramters
```{r check and plot various QC paramters lily, fig.width=10}
sum_counts_up <- as.numeric(Sys.getenv("sum_counts_up"))

library("scuttle")
sce_list <- lapply(sce_list, function(x){
  stats <- perCellQCMetrics(x, subsets=list(Mito=which(rowData(x)$chr=="MT")))
  colData(x)$sum_counts <- stats$sum
  colData(x)$genes_detected <- stats$detected
  colData(x)$sum_mito <- stats$subsets_Mito_sum
  colData(x)$mito_proportion <- stats$subsets_Mito_sum/(stats$sum)
  return(x)
})

QC_violin_plot_across_samples <- function(sce_list) {
  QC.tab.all <- NULL
  for(i in 1:length(sce_list)) {
    QC.tab.per <- data.frame(colData(sce_list[[i]])[,c('Sample', 'sum_counts', 'mito_proportion', 'genes_detected')])
    QC.tab.all <- rbind(QC.tab.all, QC.tab.per)
  }
  
  # Create the plot with horizontal lines
  p <- ggplot(QC.tab.all, aes(x=Sample, y=log2(sum_counts))) +
    ggbeeswarm::geom_quasirandom(method="smiley") +
    theme(axis.text.x = element_text(angle = 90)) +
    geom_hline(aes(yintercept=log2(500), color="500"), linetype="dashed", size=1) +  # Add red line at y=
    geom_hline(aes(yintercept=log2(sum_counts_up), color="sum_counts_up"), linetype="dashed", size=1) +  # Add blue line at y=
    scale_color_manual(name="Threshold", 
                       values=c("500"="red", "sum_counts_up"="blue"),  # Set colors for the lines
                       labels=c("sum_counts_up", "500")) +  # Set labels for the legend
    labs(y="log2(sum_counts)", x="Sample")  # Labels for x and y axis
  
  print(p)
  
  # If you want to add additional plots, you can continue as needed:

  p <- ggplot(QC.tab.all, aes(x=Sample, y=log2(genes_detected))) +
    ggbeeswarm::geom_quasirandom(method="smiley") +
    theme(axis.text.x = element_text(angle = 90)) +
    geom_hline(aes(yintercept=log2(500), color="500"), linetype="dashed", size=1) +  # Add red line at y=500
    geom_hline(aes(yintercept=log2(3000), color="3000"), linetype="dashed", size=1) +  # Add blue line at y=1000
    scale_color_manual(name="Threshold", 
                       values=c("500"="red", "3000"="blue"),  # Set colors for the lines
                       labels=c("3000", "500")) +  # Set labels for the legend
    labs(y="log2(genes_detected)", x="Sample")  # Labels for x and y axis
  print(p)

  p <- ggplot(QC.tab.all, aes(x=Sample, y=mito_proportion)) +
    ggbeeswarm::geom_quasirandom(method="smiley") +
    ggtitle("mitochondrial proportion") +
    theme(axis.text.x = element_text(angle = 90)) +
    geom_hline(aes(yintercept=0.2, color="0.2"), linetype="dashed", size=1) +  # Add red line at y=500
    geom_hline(aes(yintercept=0.1, color="0.1"), linetype="dashed", size=1) +  # Add blue line at y=1000
    scale_color_manual(name="Threshold", 
                       values=c("0.1"="red", "0.2"="blue"),  # Set colors for the lines
                       labels=c("0.1", "0.2")) +  # Set labels for the legend
    labs(y="mito_proportion", x="Sample")  # Labels for x and y axis
  print(p)
}



QC_violin_plot_across_samples(sce_list)


# invisible: do not print out the returned variable
invisible(lapply(sce_list, function(x){
  # Set the plotting area to 1 row and 3 columns
  # par(mfrow=c(1, 3))
  # 
  # hist(x$sum_counts, main=paste0(unique(x$Sample),', total UMI per droplet'),
  #      xlab='total UMI counts of a droplet', ylab = 'number of droplet')
  # abline(v=2000, col="red", lwd=1) # Add a vertical red line at x=2000
  # abline(v=5000, col="blue", lwd=1)
  # legend("topright", legend=c("1000", "5000"),
  #        col=c("red", "blue"), lty=2, cex=1.2)
  # hist(x$genes_detected, main=paste(unique(x$Sample),'\n total genes detected per droplet'),
  #      xlab='genes detected of a droplet', ylab = 'number of droplet')
  # abline(v=500, col="red", lwd=1)
  # abline(v=4000, col="blue", lwd=1)
  # legend("topright", legend=c("500", "4000"),
  #        col=c("red", "blue"), lty=2, cex=1.2)
  # 
  # hist(x$mito_proportion, main=paste(unique(x$Sample),'\n proportion of UMI counts of mito genes per droplet'),
  #      xlab='proportion of UMI counts of mito genes per droplet', ylab = 'number of droplet')
  # abline(v=0.2, col="red", lwd=1)
  # legend("topright", legend="0.2",
  #        col="red", lty=2, cex=1.2)
  # 
  # # Reset to default plotting layout
  # par(mfrow=c(1, 1))
  # 
  # plot(x$mito_proportion, x$sum_counts,
  #      main=paste0(unique(x$Sample),', mitochondrial proportion versus total UMI count'),
  #      xlab='proportion of UMI counts of mito genes per droplet', ylab='total UMI counts of a droplet')

}))


# lapply(sce_list, function(x){
#    # Set the plotting area to 1 row and 3 columns
# par(mfrow=c(1, 3))
#
# hist(x$sum_counts, main=paste0(unique(x$Sample),', total UMI per droplet'), xlab='total UMI counts of a droplet', ylab = 'number of droplet')
# abline(v=2000, col="red", lwd=1) # Add a vertical red line at x=2000
# abline(v=5000, col="blue", lwd=1)
# # a droplet contain an intact cell always have total UMI counts 2000-5000
# legend("topright", legend=c("2000", "5000"),
#      col=c("red", "blue"), lty=2, cex=1.2)
#
# hist(x$genes_detected, main='total genes detected per droplet', xlab='genes detected of a droplet', ylab = 'number of droplet')
# abline(v=1000, col="red", lwd=1)
# abline(v=3000, col="blue", lwd=1)
# legend("topright", legend=c("1000", "3000"),
#      col=c("red", "blue"), lty=2, cex=1.2)
#
# hist(x$mito_proportion, main='proportion of UMI counts of mito genes per droplet', xlab='proportion of UMI counts of mito genes per droplet', ylab = 'number of droplet')
# abline(v=0.2, col="red", lwd=1)
# legend("topright", legend="0.2",
#        col="red", lty=2, cex=1.2)
#
# # Reset to default plotting layout
# par(mfrow=c(1, 1))
#
# plot(x$mito_proportion, x$sum_counts, main=paste0(unique(x$Sample),', mitochondrial proportion versus total UMI count'), xlab='proportion of UMI counts of mito genes per droplet', ylab='total UMI counts of a droplet')
# })

```
figure 7. histogram of droplet based on total UMI counts, histogram of droplet based on genes detected, histogram of droplet based on proportion of UMI counts of mito genes, mitochondrial proportion versus total UMI count


While some level of mitochondrial expression is expected, excessively high levels (commonly around 10-20% of total UMI count) can flag a sample or cell for exclusion from downstream analysis

good quality samples have most droplets whose UMI count around 2000-5000, detected genes around 1000-3000, mitochondrial UMI proportion around or below 0.05

suspicious sample:
too few droplets, 3
low total UMI counts in most droplets, too few transcripts captured or sequenced in sample 5, 10



droplets that have high proportion of mitochondrial UMI counts and low total UMI counts are usually poor quality droplets containing lysed cell fragments, and will be filtered. most samples 1-18 only have a small number these droplets, we will filter droplets with outlier mitochondrial UMI proportions later.


### 6.4 quickCluster and visualize the droplets in UMAP and color the plots based on QC metrics
Let's now visualize the droplets in UMAP and color the plots based on QC metrics (detected genes, mitochondrial UMI proportion) and main cell types' markers to see if the suspicious droplets and samples represent doublets or an odd cell type.

```{r run PCA and UMAP individually in each sample}
head(sce_list, 3)



if(file.exists("preprocess6.4.RData"))
{
  load("preprocess6.4.RData")
}else
{

  for(i in 1:length(sce_list)){
    set.seed(100)
    clust <- quickCluster(sce_list[[i]])

    sce_list[[i]] <- computeSumFactors(sce_list[[i]], cluster=clust)
     
    zero_flag <- (sizeFactors(sce_list[[i]]) == 0)
    negative_flag <- (sizeFactors(sce_list[[i]]) < 0)

    if(sum(zero_flag) > 0 || sum(negative_flag) > 0 )
    {
      sizeFactors(sce_list[[i]])[zero_flag] <- 1
      sizeFactors(sce_list[[i]])[negative_flag] <- 1
      print("warning: non positive estimates generated due to low quality cells with few expressed features or low-abundance genes exist, need increase the stringency of the filtering.")
      print(paste("sample", i))
      
    }
    sce_list[[i]] <- logNormCounts(sce_list[[i]])
    
    sce_list[[i]] <- fixedPCA(sce_list[[i]], subset.row=NULL) 
    ## note intentionally not using HVGs here as this is a rough visualization to see all effects on droplets
    sce_list[[i]] <- runUMAP(sce_list[[i]], dimred='PCA')
  }

    save.image(file = paste0(work_dir,"preprocess6.4.RData"))
}
head(sce_list, 3)

```

Let's have a look at samples suspicious samples 3 5 10, and normal samples 1-2, 6-9, 11-18

and check QC parameters and marker genes for typical cell types to determine whether QC-outliers are of one specific cell type.

```{r check samples QC plots and marker genes expression distribution in UMAP, fig.width=12}
# install.packages("patchwork")

library(patchwork)
# c(17:20,1:16)

for(i in 1:length(sce_list))
{
  factor_vector  <- rep('normal_range',dim(sce_list[[i]])[2])
  factor_vector[(sce_list[[i]]$sum_counts <= 1000)] <- 'below_1000'
  factor_vector[(sce_list[[i]]$sum_counts >= 20000)] <- 'above_20000'
  factor_vector <- as.factor(factor_vector)
  sce_list[[i]]$sum_counts_level <- factor_vector

}



for (sample in c(1:sampNums ))
{
  x <- sce_list[[sample]]

  p1 <- plotReducedDim(x, 'UMAP', colour_by = 'sum_counts') + ggtitle(unique(x$Sample))+ coord_fixed()
  p2 <- plotReducedDim(x, 'UMAP', colour_by = 'genes_detected') + ggtitle(unique(x$Sample))+ coord_fixed()
  p3 <- plotReducedDim(x, 'UMAP', colour_by = 'mito_proportion') + ggtitle(unique(x$Sample))+ coord_fixed()


  # Combine the plots side by side
  p_combined <- p1 + p2 + p3 + plot_layout(ncol = 3)

  # Display the combined plot
  print(p_combined)


  p1 <- plotReducedDim(x, 'UMAP', colour_by = 'sum_counts_level') + ggtitle(unique(x$Sample))+ coord_fixed()
  print(p1)

########################################
  
  if(bsp_num == "bsp_247")
  {
      p1 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'CD4')]) + ggtitle(paste(unique(x$Sample),', CD4 gene')) ## CD4
  }
  if(bsp_num == "bsp_248")
  {
      p1 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'HNMT')]) + ggtitle(paste(unique(x$Sample),'myeloid, HNMT')) ## myeloid
  }


  p2 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'CD3E')]) + ggtitle('T cells, CD3E gene') ## T cells

  p3 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'PAX5')]) + ggtitle('B cells, PAX5 gene') ## B cells
  # Set uniform color scale limits
  # color_limits <- c(0, 5)
  color_set <- c("blue", "green", "yellow")

  p1 <- p1 + coord_fixed() +
    scale_color_gradientn(colors = color_set) +
    theme(legend.position = "right", aspect.ratio = 1)

  p2 <- p2 + coord_fixed() +
    scale_color_gradientn( colors = color_set) +
    theme(legend.position = "right", aspect.ratio = 1)

  p3 <- p3 + coord_fixed() +
    scale_color_gradientn( colors = color_set) +
    theme(legend.position = "right", aspect.ratio = 1)

  # Combine plots with patchwork side by side
  p_combined <- (p1 | p2 | p3) &
    plot_layout( guides = "collect")
  # Print the combined plot
  print(p_combined)


}

```

figure 8. visualize the droplets in UMAP and color the droplets based on QC metrics (total UMI counts, detected genes, proportion of mito UMI counts, 3 levels of total UMI counts,  marker gene of B cells, T cells, CD4 )

sample 3 seem fine in this check
sample 5 is mainly composed of droplets below 1000, too few transcripts to cluster and distinguish cell types of these droplets within sample 5.
sample 10, similar issue as 5, less severe.


From this examination, we need to manual check whether any major cell types are specifically characterized by high mitochondrial proportion?
most samples have few high mitochondrial proportion droplets.

so I think that we can safely filter high mitochondrial proportion droplets out.

It would be difficult to justify retaining these outliers. We can always revisit this later to look at marker genes for the droplets if desired.

sample 5 seems fail to cluster due to their low detected genes and total UMI, results in not enough gene expression information to cluster and separate these droplets. namely, the algorithm fail to detect the significant difference between droplets. this will cause difficulty in assign cell types for these droplets. we might remove these low total UMI droplets from downstream analysis.


need to check:
would that be reasonable to filter out droplets with outlier high mitochondrial proportion? or filter out droplets based on absolute mitochondrial proportion > 0.2?

here still filter out droplets with outlier high mitochondrial proportion within each sample and show filtered droplets in UMAP plots

### 6.5 filter out droplets with outlying mitochondrial proportion within each sample 
we also show these filtered droplets in UMAP
```{r filter out droplets with outlying mitochondrial proportion within each sample and show these filtered droplets in UMAP, fig.width=5, fig.height=5}

for(i in 1:length(sce_list)){
   # filter the droplets with mito_proportion larger than 0.2
   sce_list[[i]]$mito_outlier <- (sce_list[[i]]$mito_proportion >= 0.2)
   print(names(sce_list)[[i]])
   print(table(sce_list[[i]]$mito_outlier))
    cols <- rep('black', length(sce_list[[i]]$mito_outlier))
    cols[sce_list[[i]]$mito_outlier] <- 'red'

    plot(sce_list[[i]]$mito_proportion, sce_list[[i]]$sum_counts, main=paste0(i,', mitochondrial proportion versus total UMI count'), col=cols, xlab='proportion of UMI counts of mito genes per droplet', ylab='total UMI counts of a droplet')


    # # filter the droplets with outlier mito_proportion value
    # sce_list[[i]]$mito_outlier <- isOutlier(sce_list[[i]]$mito_proportion, type='higher',nmads = 5)
    # print(table(sce_list[[i]]$mito_outlier))
    # cols <- rep('black', length(sce_list[[i]]$mito_outlier))
    # cols[sce_list[[i]]$mito_outlier] <- 'red'
    # 
    # 
    # p1 <- plot(sce_list[[i]]$mito_proportion, sce_list[[i]]$sum_counts, main=paste0(names(sce_list)[i], '\n mitochondrial proportion versus total UMI count'), col=cols, xlab='proportion of UMI counts of mito genes per droplet', ylab='total UMI counts of a droplet')
    # 
    # print(p1)
    p2 <- plotReducedDim(sce_list[[i]], 'UMAP', colour_by='mito_outlier')
    print(p2)
    # sce_list[[i]] <- sce_list[[i]][,!sce_list[[i]]$mito_outlier]




    # also remove droplets whose mito_proportion higher than 0.2 to reduce noise
    sce_list[[i]] <- sce_list[[i]][, sce_list[[i]]$mito_proportion < 0.2]


  }

```


```{r check marker gene expression again, fig.width=10}
# overview number of droplets per sample
multi_list_bar_plot(sce_list,"after remove droplets \n whose mito_proportion higher than 0.2", seperate_run = seperate_run_value)

visulize_marker_expression_across_samples <- function(sce_list, re_Cluster = FALSE)
{
  sampNums <- length(sce_list)
  if(re_Cluster)
  {
    for(i in 1:length(sce_list)){
      set.seed(100)
      clust <- quickCluster(sce_list[[i]])
  
      sce_list[[i]] <- computeSumFactors(sce_list[[i]], cluster=clust)
      sce_list[[i]] <- logNormCounts(sce_list[[i]])
  
      sce_list[[i]] <- fixedPCA(sce_list[[i]], subset.row=NULL) ## note intentionally not using HVGs here as this is a rough visualization to see all effects on droplets
      sce_list[[i]] <- runUMAP(sce_list[[i]], dimred='PCA')    
  }

  }
  
  
  
  for (sample in c(1:sampNums ))
  {
    x <- sce_list[[sample]]
    
    
    p1 <- plotReducedDim(x, 'UMAP', colour_by = 'sum_counts_level') + ggtitle(unique(x$Sample))+ coord_fixed()
    print(p1)
    
    ########################################
    if(bsp_num == "bsp_247")
    {
          p1 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'CD4')]) + ggtitle(paste(unique(x$Sample),', CD4 gene')) 
    }
    if(bsp_num == "bsp_248")
    {
        p1 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'HNMT')]) + ggtitle(paste(unique(x$Sample),'myeloid, HNMT')) ## myeloid
    }

    
    p2 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'CD3E')]) + ggtitle('T cells, CD3E gene') ## T cells
    
    p3 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'PAX5')]) + ggtitle('B cells, PAX5 gene') ## B cells
    # Set uniform color scale limits
    # color_limits <- c(0, 5)
    color_set <- c("blue", "green", "yellow")
    
    p1 <- p1 + coord_fixed() +
      scale_color_gradientn(colors = color_set) +
      theme(legend.position = "right", aspect.ratio = 1)
    
    p2 <- p2 + coord_fixed() +
      scale_color_gradientn( colors = color_set) +
      theme(legend.position = "right", aspect.ratio = 1)
    
    p3 <- p3 + coord_fixed() +
      scale_color_gradientn( colors = color_set) +
      theme(legend.position = "right", aspect.ratio = 1)
    
    # Combine plots with patchwork side by side
    p_combined <- (p1 | p2 | p3) &
      plot_layout( guides = "collect")
    # Print the combined plot
    print(p_combined)
    
    
  }
  
}
QC_violin_plot_across_samples(sce_list)
visulize_marker_expression_across_samples(sce_list)


```

figure 9. visualize filtered droplets in UMAP and mitochondrial proportion versus total UMI count plot
### 6.6 filter out droplets whose detected gene less than genes_detected_threshold 
```{r remove droplets whose detected gene less than genes_detected_threshold, fig.width=10}
genes_detected_threshold <- as.numeric(Sys.getenv("genes_detected_threshold")) 

for(i in 1:length(sce_list))
{
    sce_list[[i]] <- sce_list[[i]][, sce_list[[i]]$genes_detected > genes_detected_threshold]  
}

multi_list_bar_plot(sce_list, paste("after remove droplets whose detected gene less than", genes_detected_threshold), seperate_run = seperate_run_value )
QC_violin_plot_across_samples(sce_list)
visulize_marker_expression_across_samples(sce_list, re_Cluster = TRUE)


if(bsp_num == "bsp_248")
{
  sum_counts_min <- as.numeric(Sys.getenv("sum_counts_min"))
  for(i in 1:length(sce_list))
  {
      sce_list[[i]] <- sce_list[[i]][, sce_list[[i]]$sum_counts > sum_counts_min]  
  }
  
  multi_list_bar_plot(sce_list, paste("after remove droplets whose sum_counts less than", sum_counts_min), seperate_run = seperate_run_value )
  QC_violin_plot_across_samples(sce_list)
  visulize_marker_expression_across_samples(sce_list, re_Cluster = TRUE)
    
}
```

figure 10. overview number of droplets per sample

## 7 Checking for doublets

Here we annotate suspected doublets within individual samples. We will wait to remove these droplets until after we have merged and clustered everything because it can be the case that a whole cluster is primarily doublets and we want to be sure we detect that.

```{r calculate cxds scores to Check doublets and show cxds scores on UMAP, fig.width=5, fig.height=5 }



# BiocManager::install("scds")
library(scds)
for(i in 1:length(sce_list)){
  sce_list[[i]] <- cxds(sce_list[[i]])
  sce_list[[i]]$query_doublet <- isOutlier(sce_list[[i]]$cxds_score, nmads=5, type='higher')
  p <- plotReducedDim(sce_list[[i]], 'UMAP', colour_by='cxds_score') + ggtitle(paste('sample',i))

  print(p)

  print(i)
  print(table(sce_list[[i]]$query_doublet))
  cols <- rep('black', length(sce_list[[i]]$query_doublet))
  cols[sce_list[[i]]$query_doublet] <- 'red'

  print(names(sce_list)[i])
  print(table(sce_list[[i]]$query_doublet))
  p1 <- plot(sce_list[[i]]$cxds_score, sce_list[[i]]$sum_counts, main=paste0('sample', i,', cxds_score versus total UMI count'), col=cols, xlab='cxds_score per droplet', ylab='total UMI counts of a droplet')
  print(p1)
    
# sce_list[[i]] <- sce_list[[i]][ , !sce_list[[i]]$query_doublet]    

}

```

figure 11. visualize the droplets in UMAP and color the droplets based on cxds scores, visualize droplets with outlying cxds scores in mitochondrial proportion versus total UMI count plot

sample 3 has outlier cxds score, need to check


cxds_score can be affected by sequencing run, cell type constitution, so this varies vastly across samples

there are some high cxds_score droplets in most samples, but the scores are fairly incomparable across samples.
For this reason we've marked those droplets within each sample that are more than 3 MAD above the median. We will revisit whether these are doublets after merging, normalizing and clustering.



# Normalization and merging


Now we will properly normalize the data and merge all of the samples together.

## 8.1 merge samples from same run, exclude catastrophic samples if any
```{r exclude catastrophic samples if any and merge samples }
if(bsp_num == "bsp_248")
{
  for(sample_id in c(c("sample_03", "sample_05", "sample_10" )))
  {
   sce_list[[sample_id]] <- NULL
  
  }
}



sampNums <- length(sce_list)
names(sce_list)
if(file.exists("preprocess8.1.RData"))
{
  load("preprocess8.1.RData")
}else
{

  for(i in 1:length(sce_list)){
    colnames(sce_list[[i]]) <- paste0(colnames(sce_list[[i]]), '-', names(sce_list)[i])
  }

  sce_run_all <- do.call('cbind', sce_list[c(1:sampNums )])
  # sce_run_2 <- do.call('cbind', sce_list[c(9:16)])


  save.image(file = "preprocess8.1.RData")
}


```


```{r add conditions annotation to data}


annot <- read.csv('references/sample_information.csv')
samps_id <- paste("sample", formatC(1:nrow(annot), width=2, flag="0"), sep="_")

annot$sample <- samps_id
rownames(annot) <- samps_id
head(annot)

```

then we perform quick clustering and normalization by pooled factors, correct batch effect between sequencing run.


Now we look at the data to determine whether there are batch effects that ought to be removed between samples. Here we will look for the top 5000 highly variable genes for biological variability and use these for PCA and other dimensional reduction as these are the genes we will ultimately be interested in.


## 8.2 Normalization gene expression counts, conduct batch effect correction
```{r perform quick clustering and normalization for run 1 samples and plot qc metrics before and after batch effect correction, fig.width=10}


normalize_visualize_save <- function(current_run,sce_all, Correct_batch_effects = FALSE)
{
  # cluster across all samples
  # print(head(sce_all))
  clust <- quickCluster(sce_all)
  size_factors <- calculateSumFactors(sce_all, cluster=clust)
  # log transfrom gene counts  
  sce_all <- logNormCounts(sce_all, size_factors=size_factors)
  print(sce_all)

  # get the Top 5000 High Variable Genes
  varmod <- modelGeneVar(sce_all, block=sce_all$Sample)
  # Model the variance of the log-expression profiles for each gene, decomposing it into technical and biological components based on a fitted mean-variance trend.
  vargenes <- getTopHVGs(varmod, n=5000)
  length(vargenes)

  sce_all <- fixedPCA(sce_all, subset.row=vargenes)
  set.seed(100)
  sce_all <- runUMAP(sce_all, dimred='PCA')
  set.seed(100)
  sce_all <- runTSNE(sce_all, dimred='PCA')
  
  head(sce_all$Sample)

 
  
  if(bsp_num == "bsp_247")
  {
  sce_all$condition <- annot[as.character(sce_all$Sample), 'Condition']
  sce_all$run <- annot[as.character(sce_all$Sample), 'sequencing.Run'] 
  sce_all$tissue <- annot[as.character(sce_all$Sample), 'tissue']
  sce_all$cull_date <- annot[as.character(sce_all$Sample), "cull.date..DPI."]
  sce_all$sex <- annot[as.character(sce_all$Sample), 'Sex']  

  }
  if(bsp_num == "bsp_248")
  {
  sce_all$condition <- annot[as.character(sce_all$Sample), 'condition']
  sce_all$run <- annot[as.character(sce_all$Sample), 'run']   
  sce_all$cull_date <- annot[as.character(sce_all$Sample), 'DPI']

  
  # head(sce_all$run)
  }
  
  sce_all$animal_id <- annot[as.character(sce_all$Sample), 'animal.id']      
  sce_all$animal_id <- as.factor(sce_all$animal_id) 
  sce_all$cull_date <- as.factor(sce_all$cull_date)  
  sce_all$run <- as.factor(sce_all$run)

  saveRDS(sce_all, file = paste0('sce_',current_run,'.rds'))



  ###################### plot
  title_suffix_1 = '\n based on PCA componets calculated using top 5000 highly variable genes'
  set.seed(100)
  ord <- sample.int(ncol(sce_all), ncol(sce_all), replace=FALSE)

  # pl <- plotUMAP(sce_all[,ord], colour_by="run") + ggtitle(paste(current_run,'UMAP uncorrected', title_suffix_1))
  # print(pl)    
  if(bsp_num == "bsp_248")
  {
      pl <- plotReducedDim(sce_all, 'UMAP', colour_by=rownames(sce_all)[which(rowData(sce_all)$Symbol == 'HNMT')]) + ggtitle(paste(current_run,'myeloid highlighted (HMNT)','UMAP uncorrected', title_suffix_1)) ## myeloid
  print(pl)
  }
  
  if(bsp_num == "bsp_247")
  {
  pl <- plotReducedDim(sce_all, 'UMAP', colour_by=rownames(sce_all)[which(rowData(sce_all)$Symbol == 'CD4')]) + ggtitle(paste(current_run,'CD4 highlighted (CD4)','UMAP uncorrected', title_suffix_1)) ## CD4
  print(pl)    
  }



  pl <- plotReducedDim(sce_all, 'UMAP', colour_by=rownames(sce_all)[which(rowData(sce_all)$Symbol == 'CD3E')]) + ggtitle(paste(current_run,'T cells highlighted (CD3E)', 'UMAP uncorrected', title_suffix_1)) ## T cells
  print(pl)

  pl <- plotReducedDim(sce_all, 'UMAP', colour_by=rownames(sce_all)[which(rowData(sce_all)$Symbol == 'PAX5')]) + ggtitle(paste(current_run,'B cells highlighted (PAX5)', 'UMAP uncorrected', title_suffix_1)) ## B cells
  print(pl)
  # correct between all samples in same run and plot qc metrics and UMAP and save workspace and rds for run 1 samples



  if(Correct_batch_effects)
  {
    library(batchelor)
    set.seed(100)
    # sce_all_corrected <- fastMNN(sce_all, batch=sce_all$run, subset.row = vargenes, get.variance = TRUE)
    sce_all_corrected <- fastMNN(sce_all, batch=sce_all$run, get.variance = TRUE)
    # batch correct all genes
    
    sce_all_corrected
    
  }else
  {
  # library(batchelor)
  # set.seed(100)
  # sce_all_corrected <- fastMNN(sce_all, batch=sce_all$run, subset.row = vargenes, get.variance = TRUE)
  # sce_all_corrected
  }

  set.seed(100)
  HVGs <- getTopHVGs(sce_all_corrected, n=5000) 
  sce_all_corrected <- fixedPCA(sce_all_corrected, subset.row=HVGs, assay.type = "corrected") 
  #running PCA on batch corrected expression values and just HVGs 
  sce_all_corrected <- runUMAP(sce_all_corrected, dimred="PCA")
  sce_all_corrected <- runTSNE(sce_all_corrected, dimred="PCA")
  
  # sce_all_corrected <- runUMAP(sce_all_corrected, dimred="corrected")
  # sce_all_corrected <- runTSNE(sce_all_corrected, dimred="corrected")

  sce_all_corrected$Sample <- sce_all$Sample
  sce_all_corrected$condition <- sce_all$condition
  sce_all_corrected$run <- sce_all$run
  sce_all_corrected$cull_date <- sce_all$cull_date
  sce_all_corrected$animal_id <- sce_all$animal_id  
  
  if(bsp_num == "bsp_247")
  {
  
  sce_all_corrected$tissue <- sce_all$tissue
  sce_all_corrected$sex <- sce_all$sex  
      
  }

  # sce_all_corrected$cxds_score store cxds_score values for each cell
  sce_all_corrected$cxds_score <- sce_all$cxds_score
  sce_all_corrected$query_doublet <- sce_all$query_doublet
  sce_all_corrected$sum_counts <- sce_all$sum_counts
  sce_all_corrected$genes_detected <- sce_all$genes_detected
  sce_all_corrected$mito_proportion <- sce_all$mito_proportion
  sce_all_corrected$sum_counts_level <- sce_all$sum_counts_level

  # rowData(sce_all) store the ENTREZ_symbol, char, start etc information for each gene (ensemble id)
  # rowData(sce_all_corrected) store only the score of each top 5000 HVG genes in the first 50 PC
  # so we add the ENTREZ_symbol, char, start etc information of each those gene into rowData(sce_all_corrected) again
  rowdat <- rowData(sce_all_corrected)
  rowdat <- cbind(rowData(sce_all)[rownames(sce_all_corrected),], rowdat )
  rowData(sce_all_corrected) <- rowdat


  saveRDS(sce_all_corrected, file = paste0('sce_',current_run,'_corrected.rds'))
  print(sce_all_corrected)
  ###################### plot
  set.seed(100)
  ord <- sample.int(ncol(sce_all_corrected), ncol(sce_all_corrected), replace=FALSE)

  title_suffix_2 = '\n based on top 5000 highly variable genes'

  pl <- plotUMAP(sce_all_corrected[,ord], colour_by="cxds_score") + ggtitle(paste(current_run,'UMAP corrected',title_suffix_2))
  print(pl)

  # pl <- plotUMAP(sce_all_corrected[,ord], colour_by="run") + ggtitle(paste(current_run,'UMAP corrected',title_suffix_2))
  # print(pl)  
  
  pl <- plotUMAP(sce_all_corrected[,ord], colour_by="query_doublet") + ggtitle(paste(current_run,'UMAP corrected',title_suffix_2))
  print(pl)
  # lapply(sce_all_corrected, dim)


}

current_run <- "all"
normalize_visualize_save(current_run,sce_run_all, Correct_batch_effects = TRUE)


# title_suffix = 'corrected by samples in same run'
```

figure 12.1 visualize the droplets in UMAP and color the droplets based on sample;  condition; marker gene expression level of B cells, T cells, CD4;  cxds scores; suspected doublet for sequencing run 1



after batch effect correction. we observe droplets from all samples distribute more uniformly in UMAP plots.
sample 5 does not cluster together with other samples.

We will leave this analysis here and start a new script where we cluster, subcluster, filter doublets, etc.





```{r perform quick clustering and normalization for run 2 samples and plot qc metrics before batch effect correction}
# current_run <- "run_2"
# normalize_visualize_save(current_run,sce_run_2, Correct_batch_effects = FALSE)

```

figure 12.2 visualize the droplets in UMAP and color the droplets based on sample;  condition; marker gene expression level of B cells, T cells, CD4;  cxds scores; suspected doublet for sequencing run 2

```{r perform quick clustering and normalization for run 3 samples and plot qc metrics before batch effect correction}

#
# normalize_visualize_save("run_3_BAL",sce_run_3_BAL, Correct_batch_effects = FALSE)
# normalize_visualize_save("run_3_lung",sce_run_3_lung, Correct_batch_effects = FALSE)



```


figure 12.3 visualize the droplets in UMAP and color the droplets based on sample;  condition; marker gene expression level of B cells, T cells, CD4;  cxds scores; suspected doublet for sequencing run 3
<!-- # 9 save image and show session Info -->
<!-- ```{r sessionInfo} -->
<!-- save.image(file = "preprocess9.RData") -->
<!-- sessionInfo() -->

<!-- ``` -->



