---
title: "Processing scRNA data bsp 248"
author: "lily yang"
date: "`r Sys.Date()`"
output: html_document
---

# setup Environment  
```{r, setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
work_dir <- Sys.getenv("work_dir")
knitr::opts_knit$set(root.dir = work_dir)

work_dir
# work_dir  <- '/mnt/lustre/RDS-ephemeral/bioinformatics/analysis/scrna/proj/bsp_248/lily/stages/'
# work_dir  <- "/mnt/rstudio/lily/bsp248/")
count_dir <- '/mnt/lustre/RDS-ephemeral/bioinformatics/analysis/scrna/proj/bsp_248/lily/stages/count'


sampSet <- Sys.getenv("sampSet")
if (sampSet == "") {
   sampSet <- "all"      # defaulty
   }

sampSet

sampSets <- list( "all" = 1:18,
                  "run1" = 1:9   
                )
              
sampNums <- sampSets[[sampSet]]
run1 <- 1:9
run2 <- 10:18
samps <- paste("sample", formatC(sampNums, width=2, flag="0"), sep="_")
count_filename <- paste0('muci_', samps, '_count')

# load packages
library(SingleCellExperiment)
library(DropletUtils)
library(viridisLite)     

```


# Intro

This code will run an initial processing of the porcine Unsorted lung tissue 10x Genomics scRNAseq data generated at the Pirbright Institute in July and August 2024. bcl files were demultiplexed using `cellranger mkfastq` and aligned and counted using `cellranger count` (cellranger-7.1.0) using default parameters and Sus scrofa genome (genome assembly 11.1, Ensembl release 107).

Note saturation estimated by cellranger as follows:
need to do, write a sh script to extract this values for multiple samples 

# import and remove swapped droplets

We will read in raw counts matrices generated by cellranger to allow us to filter out empty droplets and removed barcode-swapped droplets using tools from DropletUtils. (cellranger filtered pipeline are too loose and can't identify all issues in the data, so it's better to be able to do extra checking and filtering ourselves.)

## 1 Reading in the data

In this data set, we sequenced 18 samples, over 2 sequencing runs. There is some concerns from Graham that run 2 did not run correctly.
experiment design: 
There are 2 time points which will we number as  dpi1 (1 days post-infection) and dpi12 (12 days post-infection).
There are 2 interventions PRCV and H1N1. There are 2 sets of controls: dpi1 (x2) and dpi12 (x2).


```{r import data in cluster and plot overview cell number, fig.width=10}


if(file.exists("preprocess1.RData"))
{
  load("preprocess1.RData")
}else
{
  setwd(count_dir)
  
  dirs <- paste0( count_filename, '/outs/raw_feature_bc_matrix')
  sce_list_raw <- vector('list', length=length(sampNums))
  names(sce_list_raw)  <- samps

  for(i in sampNums) {
      # Use try() to capture potential errors
      sce_result <- try(read10xCounts(samples = dirs[i], sample.names = samps[i], col.names = TRUE), silent = TRUE)
      
      # Check if an error occurred
      if (inherits(sce_result, "try-error")) {
        
          print(paste("Error, there is no fq files for sample", samps[i], "demultiplex or count step fail"))
      } else {
          # If no error, store the result in the list
          sce_list_raw[[i]] <- sce_result
          print(i)
      }
  }
     
  save.image(file = paste0(work_dir,'preprocess1.RData'))
}
# 
lapply(sce_list_raw, dim)
# head(lapply(sce_list_raw, dim))
# head(sce_list_raw)
# Remove NULL elements from the list
sce_list_raw <- Filter(Negate(is.null), sce_list_raw)
sampNums <- length(sce_list_raw)


sce = sce_list_raw
multi_list_bar_plot <- function(sce, my_flag, seperate_run = 9.5)
{
  df <- as.data.frame(lapply(sce, dim))
  df <- as.data.frame(t(df))
  colnames(df) <- c("gene_number","droplet_number")
  df$sample <- as.factor(rownames(df))
  library(ggplot2)
  # Basic barplot
  p1 <- ggplot(data=df, aes(x=sample, y=droplet_number)) +
    geom_bar(stat="identity")+
    geom_text(aes(label=droplet_number), vjust=-0.3, size=3.5) +
    geom_vline(xintercept = seperate_run, color = "red", linetype = "dashed", size = 1)+
    ggtitle(paste0("droplet number summary ", my_flag))+
  theme(axis.text.x = element_text(angle = 90)) 
  # geom_vline(xintercept = 16.5, color = "blue", linetype = "dashed", size = 1)+
  


}

multi_list_bar_plot(sce_list_raw,"before qc filtering", seperate_run = 9.5 )

```

figure 1. droplet number per sample before preprocess 


## 2 Removing swapped droplets

Now we want to look for any droplets that have swapped cell barcodes due to use of patterned flow cell sequencing. 8 samples were sequenced in each run, and so swapping could occur anywhere within the sequencing run but not outside of it.

```{r Removing swapped transcript molecules}

if(file.exists("preprocess2.RData"))
{
  load("preprocess2.RData")}else
{
  getwd()
  setwd(count_dir)
  sce_list <- sce_list_raw
  
  
  ## run 1
  fs <- paste0('muci_sample_0', as.character(1:9), '_count/outs/molecule_info.h5')
  run1 <- swappedDrops(samples=fs)
  
  # samps_path <- paste("muci_sample", formatC(sampNums, width=2, flag="0"), sep="_")

## run 2
  
  fs <- paste0('muci_sample_', as.character(16:18), '_count/outs/molecule_info.h5')
  run2 <- swappedDrops(fs)


  # create a empty list to save the data that have Removed swapped drops
  cleaned <- list()
  
  for (i in 1:9) {
      cleaned[[i]] <- run1[[1]][[i]]
  }
  for (i in 10:12) {
      cleaned[[i]] <- run2[[1]][[i-9]]
      # head(paste0(colnames(cleaned[[i]]), '-1'))
      # typeof(run2[[1]][[1]])
      # names(run2[[1]])
  }
  # add -1 for each droplet barcode 
  for(i in 1:length(cleaned)){
    colnames(cleaned[[i]]) <- paste0(colnames(cleaned[[i]]), '-1')
  }
  
  # lapply(sce_list, dim)
  
  # evaluate how much droplets are removed in this step
  for (i in 1:9) 
  {
    print(i)
    print(sum(colSums(run1$cleaned[[i]]))/sum(colSums(counts(sce_list[[i]]))))
  }
  
  
  for (i in 1:3) 
  {
    print(i+9)
    print(sum(colSums(run2$cleaned[[i]]))/sum(colSums(counts(sce_list[[i+9]]))))
  }
  
  # replace original data with clean data that have Removed swapped drops
  for(i in 1:length(sce_list)){
    counts(sce_list[[i]])[,colnames(cleaned[[i]])] <- cleaned[[i]]
      # take long to run this step
  }
  
  rm(cleaned, run1, run2)  
  save.image(file = paste0(work_dir,'preprocess2.RData'))
}


```

Swapping was extremely minimal (< 0.01). We've removed the swapped reads anyways, but it was hardly consequential.

# QC of droplets
## 3 Filtering out empty droplets
### 3.1 check Count by rank plot
Now we filter empty droplets in each sample separately.
First we'll look at the Count by rank plot for UMI counts within each droplet, just to get a visual overview of whether anything is odd.
Each plot shows the total UMI count of each droplet against their rank. Droplets are ranked based on their total UMI count.

```{r produce Count by rank plot for UMI counts, fig.width=10 }

Count_by_rank_plot <- function(sce_list, calculate_brs = TRUE, use_before_filter_brs = FALSE, before_filter_brs = NULL)
{
  if(calculate_brs)
  {
  brs <- lapply(sce_list, function(x) {
    temp <- counts(x)
    return(barcodeRanks(temp, lower = 200, fit.bounds = c(300,35000)))})
  # barcodeRanks function: Compute barcode rank statistics and identify the knee and inflection points on the total count curve. lower = 200, fit.bounds = c(300,35000) are used to alleviate the effect of noise when fit model
  # lower	
  # A numeric scalar specifying the lower bound on the total UMI count, at or below which all barcodes are assumed to correspond to empty droplets.
  # 
  # fit.bounds	
  # A numeric vector of length 2, specifying the lower and upper bounds on the total UMI count from which to obtain a section of the curve for spline fitting.
  }

  par(mfrow=c(1, 2))
  for(i in 1:length(brs)){
    # returning only unique rank for plotting speed
    uniq <- !duplicated(brs[[i]]$rank)
    plot(brs[[i]]$rank[uniq], brs[[i]]$total[uniq], log="xy",
         xlab="Rank", ylab="Total UMI count", main=unique(sce_list[[i]]$Sample), cex.lab=1.2)
    if(use_before_filter_brs)
    {
    # draw Count by rank plot using knee and inflection points derived from sce_list before filtering empty droplets
    abline(h=metadata(before_filter_brs[[i]])$inflection, col="darkgreen", lty=2)
    abline(h=metadata(before_filter_brs[[i]])$knee, col="dodgerblue", lty=2)
    legend("bottomleft",
           legend=c(paste("Knee",metadata(before_filter_brs[[i]])$knee),
                    paste("Inflection",metadata(before_filter_brs[[i]])$inflection)),
           col=c( "dodgerblue","darkgreen"), lty=2, cex=1)
    }else
    {
    # draw Count by rank plot using knee and inflection points derived from current sce_list 
    abline(h=metadata(brs[[i]])$inflection, col="darkgreen", lty=2)
    abline(h=metadata(brs[[i]])$knee, col="dodgerblue", lty=2)
    legend("bottomleft",
           legend=c(paste("Knee",metadata(brs[[i]])$knee),
                    paste("Inflection",metadata(brs[[i]])$inflection)),
           col=c( "dodgerblue","darkgreen"), lty=2, cex=1)
    }

  }

  # Reset to default plotting layout
  par(mfrow=c(1, 1))

}

Count_by_rank_plot(sce_list, TRUE)


# use before_filter_brs to store the value of knee and inflection points derived from sce_list before filtering  empty droplets
before_filter_brs <- lapply(sce_list, function(x) {
    temp <- counts(x)
    return(barcodeRanks(temp, lower = 200, fit.bounds = c(300,35000)))})


```

figure 2. Count by rank plot before filtering empty droplets

an ideal knee point should locate at the curve where its slope significantly shifts and become steep, ideal y axis value of this point should over 1000 since total UMI count lower than 1000 might fail in cluster step


an ideal inflection point should locate at the middle of a steep slope of the curve, ideal value over 500

if under default parameter, due to the interfere of outlier droplets with abnormally high total UMI, estimated knee points of such Sample will become too early.

suspicious samples:
5, 




### 3.2 background knowledge supplement
i think we are less interested in these those kinds of droplets with too low total UMI count.
so i prefer set strict qc metrics first to filter out droplets lower than inflection points.

develop later: 
how much of the filtered droplet can still be used, distinguish droplets have ambient RNAs , droplets have transcription inactive intact droplets.



### 3.3  validate assumption, droplets containing fewer than 100 UMIs do not contain any real droplets

we retrieve droplets whose total UMI count between 0-100, check their P values (possibility of being non-empty droplets) using 'emptyDrops' function, and plot histogram of the predicted P-Value.
default assumption is droplets containing fewer than 100 UMIs do not contain any real droplets, therefore, in theory P Value of these droplets should near 1, and should not peak around 0. 



```{r calculate P values for droplets containing fewer than 100 UMIs and plot histogram of their P values to check how accurate the emptyDrops estimate empty droplets}
if(file.exists("preprocess3.3.RData"))
{
  load("preprocess3.3.RData")
}else
{
  e.out <- vector('list', length=length(sce_list))
  all.out <- vector('list', length=length(sce_list))
  for(i in 1:length(sce_list)){
    set.seed(100)
    e.out[[i]] <- emptyDrops(counts(sce_list[[i]]), niters = 100000,
                             test.ambient = FALSE)
    # Distinguish between droplets containing cell and ambient RNA in a droplet-based single-cell RNA sequencing experiment.


    set.seed(100)
    limit <- 100   ## set at 100 because this is the default for emptyDrops used above
    # obtain all droplets' possibility of being non-empty droplets

    all.out[[i]] <- emptyDrops(counts(sce_list[[i]]), lower=limit, test.ambient=TRUE, niters = 100000)
    # test.ambient=TRUE, A logical scalar indicating whether results should be returned for barcodes with totals less than or equal to 'lower'.
    # lower=limit, A numeric scalar specifying the lower bound on the total UMI count,  at or below which all barcodes are assumed to correspond to empty droplets.

  # If test.ambient=TRUE, p-values will be reported for all barcodes with positive total counts, including those not greater than lower, which otherwise were reported as non-NA. This is occasionally useful for diagnostics to ensure that the p-values are well-calibrated for barcodes corresponding to (presumably) empty droplets.

  # Specifically, if the null hypothesis were true, p-values for low-count barcodes should have a uniform distribution. Any strong peaks in the p-values near zero indicate that emptyDrops is not controlling the FDR correctly.
  #
  # Note that, when setting test.ambient=TRUE in emptyDrops, barcodes less than or equal to lower will still have NA values in FDR. Such barcodes are still explicitly ignored in the correction as these are considered to be uninteresting.

  save.image("preprocess3.3.RData")


  }

}

for(i in 1:length(sce_list))
{
  print(summary(e.out[[i]]$FDR <= 0.001))
  # FDR <= 0.001 indicate droplets containing cell

  print(table(Sig=e.out[[i]]$FDR <= 0.001, Limited=e.out[[i]]$Limited))
  # niters parameter can be increased in the future if Limited=TRUE and Sig=FALSE  are very high

  hist(all.out[[i]]$PValue[all.out[[i]]$Total <= limit & all.out[[i]]$Total > 0],
     xlab="P-value", main=unique(sce_list[[i]]$Sample), col="grey80")
  # PValue near 0 means this droplet is inferred to be cell contained, near 1 means this droplet is inferred to be ambient RNA, don't contain cell

}

```

figure 3. P values histogram of droplets containing fewer than 100 UMIs


if Limited=TRUE, it indicates whether the p-value for a droplet is limited by the number of iterations (niters),  it indicate the true p-value could potentially be lower than computed if more iterations were performed.

"PValue" refers to the statistical p-value obtained from testing each barcode (representing a droplet) for significant deviations from the ambient RNA profile

If any barcodes have Limited=TRUE but does not reject the null hypothesis(), it suggests that niters should be increased.


Number of iterations for all samples looks appropriate now, as all droplets whose value of Limited=TRUE and Sig=FALSE are always 0.

ideal niters parameter should make Limited=TRUE and Sig=FALSE are 0 for each sample
need to do:
add code to automatically check whether Limited=TRUE and Sig=FALSE are 0 for each sample, and report fail samples.


in the histogram, P-values of droplets that are assumed to ambient reads sometimes dip near 1 but have no peaks near 0, coincide with the assumption that droplets containing fewer than 100 UMIs do not contain any real droplets.
therefore, samples have ideal quality plots there are:  5, 7, 8




### 3.4 filter the estimated empty droplets and plot total UMI histogram
Thus we can filter the droplets

```{r filter the empty droplets based on FDR value and check Count by rank plot after filter the empty droplets, fig.width=10 }

sce_list_before_filter_empty_droplets <- sce_list
if(file.exists("preprocess3.4.RData"))
{
  load("preprocess3.4.RData")
}else
{

  for(i in 1:length(sce_list))
    {
    sce_list[[i]] <- sce_list[[i]][,which(e.out[[i]]$FDR <= 0.001)]
    # sce_list[[i]] <- sce_list[[i]][,which(e.out[[i]]$Total >= metadata(before_filter_brs[[i]])$inflection )]
    }
    save.image(file = "preprocess3.4.RData")
  }



lapply(sce_list, dim)



Count_by_rank_plot(sce_list, calculate_brs = TRUE, use_before_filter_brs = TRUE, before_filter_brs)

```
figure 4. Count by rank plot after filtering empty droplets

So the apparent odd slope in Sample 5 has dramatically large number of droplets that have quiet low total UMI (<600), based on FDR value, these droplets are still estimated as cells and been kept. these low total UMI droplets might be an issue, and we will revisit this as we continue to process and QC. maybe later we will filter all droplets that have quiet low total UMI (<600).

if these droplets are real outliers, them tend to cluster together, with abnormal gene expression, should be observed in UMAP plots.



let's have a look at the distribution total UMI, and filter abnormally high total UMI outlier droplets by set the threshold


```{r plot total UMI per droplet histogram before and after filtering and filter droplets with abnormally high or low total UMI }
library("scuttle")
sce_list <- lapply(sce_list, function(x){
  stats <- perCellQCMetrics(x)
  colData(x)$sum_counts <- stats$sum # total UMI per droplet
  return(x)
})

plot_total_UMI_distribution <- function(sce_list)
{
   # Set the plotting area to 1 row and 3 columns
# par(mfrow=c(1, 3))

  lapply(sce_list, function(x){

    hist(x$sum_counts, main=paste0(unique(x$Sample),', total UMI per cell'), xlab='total UMI counts of a cell', ylab = 'number of droplets',
         xlim = c(0,20000), breaks = seq(0,200000,by=1000))
  abline(v=2000, col="red", lwd=1) # Add a vertical red line at x=2000
  abline(v=5000, col="blue", lwd=1)
  # a droplet contain an intact cell always have total UMI counts 2000-5000
  legend("topright", legend=c("2000", "5000"),
         col=c("red", "blue"), lty=2, cex=1.2)

})

# Reset to default plotting layout
par(mfrow=c(1, 1))
}
plot_total_UMI_distribution(sce_list)

# remove droplets with extreme high or low Total UMI
 for(i in 1:length(sce_list)){
   # sce_list[[i]]$UMI_counts_outlier <- isOutlier(sce_list[[i]]$sum_counts, type='higher', nmads = 5)
   # since large number of droplets that have quiet low total UMI (<600) in sample 17-18 will make other droplets that have normal total UMI (2000-5000) become ouliers, so there we don't use isOutlier to filter abnormally high total UMI droplets (>50,000)

   sce_list[[i]]$UMI_counts_outlier <- (sce_list[[i]]$sum_counts >= 50000) | (sce_list[[i]]$sum_counts <= 500)
   # maybe will filter sum_counts <= 800 in the future. sample 18 have too many droplets around 800 total UMI, need to check if them can cluster successfully
   print(i)
   print(table(sce_list[[i]]$UMI_counts_outlier))

   sce_list[[i]] <- sce_list[[i]][,!sce_list[[i]]$UMI_counts_outlier]
 }





plot_total_UMI_distribution(sce_list)

lapply(sce_list, dim)

```

figure 5. plot total UMI per droplet histogram before and after filtering

from the distribution total UMI plot of sample 17-18, we observe after filtering the low total UMI droplets, the distribution plot looks clearer and the number of droplets with 2000-5000 total UMI is over 3000, which is good and comparable with sample 19-20 in the same run.

need to check:
but in sample 18 the number of droplets with 2000-5000 total UMI is too little. let's check the cluster results of sample 18 to see how severe the issue is and if this sample can provide enough information.

need to check:
shall we filter sum_counts >= 20,000 in the future?

we try produce Count by rank plot again using previously predicted inflation and knee points, them fits the new filtered clean data better in most samples.  and we also calculate the percentage of droplets been filtered in this step and the number of droplets left after this filtering step.

```{r plot number of droplets per sample, fig.width=10 }


# Count_by_rank_plot(sce_list, calculate_brs = TRUE, use_before_filter_brs = FALSE)

#plot number of droplets per sample
multi_list_bar_plot(sce_list,"after filtering empty droplets based on inflection point" )

for(i in 1:length(sce_list))
{
  print(dim(sce_list[[i]])[2] / dim(sce_list_raw[[i]])[2])
}
```

figure 6. plot number of droplets per sample

## 4 Removing ambient RNA

While removing ambient RNA is not always necessary, the data we are dealing with comes from different immunological settings of strain 135 infected versus strain ISU-1 infected. As such, there will be some different cell states in each animal and these will contribute to background RNA. The worry is that it might appear that *all* droplets are expressing e.g. specific inflammatory genes as opposed to the main cell type of that cell in their normal condition. Thus, we will remove ambient RNA from each sample. basically, this step change the total UMI counts in each sample by reduce UMI counts of some genes within each droplet.

We start by running a quick clustering.

```{r quick clustering}
# BiocManager::install("scran")
library(scran)
library(scater)
library(uwot)

if(file.exists("preprocess4.RData"))
{
  load("preprocess4.RData")
}else
{
  clust <- quickCluster(sce_list[[1]])
  # library(igraph)
  for(i in 1:length(sce_list))
    {
    set.seed(100)
    clust <- quickCluster(sce_list[[i]])
    colLabels(sce_list[[i]]) <- clust
  }
  save.image('preprocess4.RData')
  }

```

Next we will use this rough clustering to remove ambient RNA.

<!-- from documentation of function 'removeAmbience' -->
<!-- Estimate and remove the ambient profile from a count matrix, given pre-existing groupings of similar droplets. This function is largely intended for plot beautification rather than real analysis. -->

<!-- This function will aggregate counts from each group of related droplets into an average profile. For each group, we estimate the contribution of the ambient profile and subtract it from the average. By default, this is done with ambientContribMaximum, but if enough is known about the biological system, users can specify features by using ambientContribNegative instead. -->

<!-- We then perform quantile-quantile mapping of counts in y from the old to new averages. This approach preserves the mean-variance relationship and improves the precision of estimate of the ambient contribution, but relies on a sensible grouping of similar droplets, e.g., unsupervised clusters or cell type annotations. As such, this function is best used at the end of the analysis to clean up expression -->

<!-- from documentation of function 'removeAmbience', it says this function is largely intended for plot beautification after clustering and cell type annotations finished. rather than real analysis. so i am not sure if it's suitable there to do this step. -->
<!-- i don't run this step now. -->


<!-- ```{r use rough clustering to remove ambient RNA} -->


<!-- # sce_list_raw <- sce_list -->
<!-- # for(i in 1:length(sce_list)){ -->
<!-- #   amb <- metadata(e.out[[i]])$ambient[,1] -->
<!-- #   # e.out is the output of function emptyDrops in last step -->
<!-- #   # amb save each gene's contribution as ambient in one sample -->
<!-- # -->
<!-- #   amb_full <- rep(0, nrow((sce_list[[i]]))) -->
<!-- #   # amb_full has the length of 33028, the number of all genes measured across all sample -->
<!-- #   names(amb_full) <- rownames(sce_list[[i]]) -->
<!-- #   amb_full[names(amb)] <- amb -->
<!-- #   # assign value to some of the elements of amb_full -->
<!-- #   assays(sce_list[[i]], withDimnames=FALSE)$counts <- removeAmbience(counts(sce_list[[i]]), -->
<!-- #                    ambient=amb_full, groups=sce_list[[i]]$label) -->
<!-- #   # removeAmbience function there is from DropletUtils library -->
<!-- # } -->
<!-- # head(amb) -->
<!-- # head(amb_full) -->
<!-- # rm(e.out, amb_full, amb) -->
<!-- # -->
<!-- # for(i in 1:length(sce_list)) -->
<!-- # { -->
<!-- #   print(sum(assays(sce_list[[i]], withDimnames=FALSE)$counts) / sum(assays(sce_list_raw[[i]], withDimnames=FALSE)$counts)) -->
<!-- #   # cell number doesn't change for each sample, the total UMI of each sample might change -->
<!-- # } -->
<!-- # -->
<!-- # rm(sce_list_raw) -->


<!-- ``` -->


<!-- ```{r remove ambient RNA using SoupX} -->
<!-- library('SoupX') -->
<!-- # Extract raw counts and cell metadata from SingleCellExperiment objects -->
<!-- sce_list_before_remove_ambient_rna <- sce_list -->


<!-- for(i in 1:length(sce_list)) -->
<!-- { -->
<!--   raw_counts <- counts(sce_list_before_filter_empty_droplets[[i]]) -->
<!--   filtered_counts <- counts(sce_list[[i]]) -->
<!--   # Create a SoupChannel object -->
<!--   sc <- SoupChannel(raw_counts, filtered_counts) -->

<!--   # Add cluster information to the SoupChannel object -->
<!--   # colData(sce_list[[i]])$label -->
<!--   # unique(colData(sce_list[[i]])$label) -->
<!--   # colLabels(sce_list[[i]]) -->

<!--   sc = setClusters(sc, setNames(colLabels(sce_list[[i]]), colnames(sce_list[[i]]))) -->
<!--   # Estimate soup profile and contamination fraction -->
<!--   # pdf(file = "SoupX_output.pdf") -->
<!--   print(paste('sample', i)) -->
<!--   sc <- autoEstCont(sc) -->
<!--   # dev.off() -->

<!--   # Output the clean data which removed the predicted ambient reads -->
<!--   out <- adjustCounts(sc) -->
<!--   counts(sce_list[[i]]) <- out -->

<!-- } -->

<!-- # Check the change of adjusted counts -->

<!-- for(i in 1:length(sce_list)) -->
<!-- { -->
<!--   print(sum(assays(sce_list[[i]], withDimnames=FALSE)$counts) / sum(assays(sce_list_before_remove_ambient_rna[[i]], withDimnames=FALSE)$counts)) -->
<!--   # cell number doesn't change for each sample, the total UMI of each sample might change -->
<!-- } -->


<!-- ``` -->


<!-- ## 5 Cellranger filtering comparison -->

<!-- For comparison, we will also read in the cellranger filtered output to see how results compare in terms of number of droplets. -->

<!-- ```{r load cellranger filtered output and compare with our filtered output} -->
<!-- library(SingleCellExperiment) -->
<!-- library(DropletUtils) -->

<!-- if(file.exists("preprocess5.RData")) -->
<!-- { -->
<!--   load("preprocess5.RData") -->
<!-- }else -->
<!-- { -->
<!--   setwd(rehash_dir) -->
<!--   dirs <- paste0( samps, '/filtered_feature_bc_matrix') -->
<!--   sce_list_cr <- vector('list', length=20) -->

<!--   for(i in 1:20) -->
<!--   { -->
<!--     sce_list_cr[[i]] <- read10xCounts(samples=dirs[i], sample.names=samps[i], col.names=TRUE) -->
<!--   } -->

<!--   save.image(file = paste0(work_dir,"preprocess5.RData")) -->
<!-- } -->



<!-- lapply(sce_list, dim) -->

<!-- lapply(sce_list_cr, dim) -->

<!-- for(i in 1:length(sce_list)) -->
<!-- { -->
<!--   # print((dim(sce_list[[i]])[2] - dim(sce_list_cr[[i]])[2])/ -->
<!--   #         ((dim(sce_list[[i]])[2] + dim(sce_list_cr[[i]])[2])/2)*100) -->
<!--   print(dim(sce_list[[i]])[2] / dim(sce_list_cr[[i]])[2]) -->
<!-- } -->

<!-- rm(sce_list_cr) -->
<!-- ``` -->


<!--  compare remained number of droplets within each sample across cellranger filtered output and our filtered output -->



<!-- ## 6 Removing residual low quality droplets -->

<!-- ### 6.1 filter zero counts genes and annotate genes names -->
<!-- Here we will look for outlying droplets in terms of detected gene number and proportion of mitochondrial reads.  -->

<!-- In order to look at specific types of genes, we need to annotate the dataset. To make this easier, we'll first remove genes with no counts in every sample -->

<!-- ```{r filter zero counts genes and map genes symbols original} -->


<!-- # Remove genes with zero counts -->
<!-- detected <- lapply(sce_list, function(x){rowSums(counts(x)) > 0}) -->
<!-- # detected is a list with 20 elements, each element contain all genes with true or false value -->
<!-- names(detected) <- paste0('Sample', as.character(c(1:20))) -->
<!-- detected <- data.frame(detected) -->
<!-- # 33028 row    20 columns -->
<!-- undetected <- which(rowSums(detected) == 0) -->
<!-- # undetected record such gene's id and position  -->
<!-- print(paste('number of genes with zero counts in every sample', length(undetected))) -->


<!-- if(file.exists("preprocess6.1.RData")) -->
<!-- { -->
<!--   load("preprocess6.1.RData") -->
<!-- }else -->
<!-- { -->
<!--   # only delete about 2000 genes out of 30000+ genes -->
<!--   sce_list <- lapply(sce_list, function(x) {return(x[-undetected,])}) -->
<!--   rm(detected, undetected) -->
<!--   # lapply(sce_list, dim) -->
<!--   if(file.exists("references/ensembl_map.csv")) -->
<!--   { -->
<!--     map <- read.csv("references/ensembl_map.csv", row.names = 1) -->
<!--     # head(map) -->
<!--   }else -->
<!--   { -->

<!--   # retrieve hgnc_symbol for the remaining non-zero genes -->
<!-- # unloadNamespace("tidyselect") -->
<!--     unloadNamespace("dplyr") -->
<!--     library(biomaRt) -->

<!--     success <- FALSE   -->
<!--     while(!success )  -->
<!--     { -->
<!--       try({ -->
<!--         ensembl <- useEnsembl(biomart = "genes", dataset = "sscrofa_gene_ensembl", mirror = "www", verbose = TRUE) -->
<!--         # If the above line runs without error, set success to TRUE to exit the loop -->
<!--         success <- TRUE}, silent = TRUE)  -->
<!--       # 'silent = TRUE' suppresses the error message in the console -->

<!--       if(!success)  -->
<!--       { -->
<!--         message(paste("Attempt failed. Retrying...")) -->
<!--         Sys.sleep(5) # Wait for 5 seconds before retrying to give some time for the internet issue to resolve -->
<!--       } -->

<!--     }   -->

<!--     map <- getBM(attributes=c('ensembl_gene_id',  -->
<!--                               'hgnc_symbol', 'chromosome_name', -->
<!--                               'start_position', 'end_position'), -->
<!--                  filters='ensembl_gene_id',  -->
<!--                  values=rownames(sce_list[[1]]), -->
<!--                  mart=ensembl) -->
<!--     rownames(map) <- as.character(map$ensembl_gene_id) -->
<!--     write.csv(map, "ensembl_map.csv") -->
<!--   } -->



<!--   sce_list <- lapply(sce_list, function(x){ -->
<!--     rowData(x)$hgnc_symbol_old <- map[rownames(x), 'hgnc_symbol'] -->
<!--     rowData(x)$chr <- map[rownames(x), 'chromosome_name'] -->
<!--     rowData(x)$start <- map[rownames(x), 'start_position'] -->
<!--     rowData(x)$end <- map[rownames(x), 'end_position'] -->
<!--     return(x)}) -->



<!--   head(rowData(sce_list[[1]])) -->
<!--   save.image(file = paste0(work_dir,"preprocess6.1.RData")) -->

<!-- } -->
<!-- ``` -->


<!-- ```{r examine map quality of original method} -->
<!-- print(paste('number of unique hgnc_symbol_old', length(unique(rowData(sce_list[[1]])$hgnc_symbol_old)))) -->
<!-- print(paste('total number of hgnc_symbol_old', length(rowData(sce_list[[1]])$hgnc_symbol_old))) -->
<!-- # check how much gene names are assigned successfully -->

<!-- print(paste('number of unique ensembl_gene_id', length(unique(rownames(sce_list[[1]]))))) -->
<!-- print(paste('total number of ensembl_gene_id', length(rownames(sce_list[[1]])))) -->
<!-- # check how much unique gene id exist -->

<!-- # check if CD8B assigned successfully -->
<!-- grep('CD8B',rowData(sce_list[[1]])$hgnc_symbol_old) -->

<!-- rowData(sce_list[[1]])$hgnc_symbol_old[grep('CD8',rowData(sce_list[[1]])$hgnc_symbol_old)] -->

<!-- # check if CD8B exist according its ensembl_gene_id -->
<!-- grep('ENSSSCG00000008213',rownames(sce_list[[1]])) -->


<!-- map['ENSSSCG00000008213',] -->

<!-- map[map$ensembl_gene_id == 'ENSSSCG00000060380',] -->
<!-- map[map$ensembl_gene_id == 'ENSSSCG00000008213',] -->


<!-- ``` -->

<!-- ```{r second map genes symbols method and examine map quality of second method}} -->

<!-- if(file.exists("references/ensembl_entrez_map.csv")) -->
<!-- { -->
<!--   ensembl_entrez <- read.csv("references/ensembl_entrez_map.csv", row.names = 1)   -->
<!-- }else -->
<!-- { -->
<!--   ################################## Get Entrez Gene IDs corresponding to each Ensembl ID -->
<!--   # library(biomaRt) -->
<!--   success <- FALSE -->
<!--   while(!success )  -->
<!--   { -->
<!--     try({ -->
<!--       ensembl <- useEnsembl(biomart = "genes", dataset = "sscrofa_gene_ensembl", mirror = "www", verbose = TRUE) -->
<!--       # If the above line runs without error, set success to TRUE to exit the loop -->
<!--       success <- TRUE}, silent = TRUE)  -->
<!--     # 'silent = TRUE' suppresses the error message in the console -->

<!--     if(!success)  -->
<!--     { -->
<!--       message(paste("Attempt failed. Retrying...")) -->
<!--       Sys.sleep(5) # Wait for 5 seconds before retrying to give some time for the internet issue to resolve -->
<!--     } -->

<!--   }  -->

<!--   entrez_ids <- getBM(attributes=c('ensembl_gene_id', 'entrezgene_id'), -->
<!--                       filters='ensembl_gene_id', -->
<!--                       values=rownames(sce_list[[1]]), -->
<!--                       mart=ensembl) -->
<!--   dim(entrez_ids) -->
<!--   # head(entrez_ids) -->

<!--   entrez_ids[entrez_ids$ensembl_gene_id == 'ENSSSCG00000008213',] -->

<!--   ################################## Get symbols corresponding to each Entrez Gene IDs -->
<!--   # BiocManager::install("org.Ss.eg.db") -->
<!--   library(org.Ss.eg.db) -->
<!--   vector <- as.character(entrez_ids$entrezgene_id) -->

<!--   Named_vector_of_HGNC_symbols <- mapIds(org.Ss.eg.db,  -->
<!--                                          keys=vector,  -->
<!--                                          column="SYMBOL",  -->
<!--                                          keytype="ENTREZID",  -->
<!--                                          multiVals = "first") -->

<!--   # the key of the output list are the ENTREZID IDs of the input  -->
<!--   # vector. The values of the output vector are the HGNC symbols.  -->

<!--   # see all the info and ID types stored in the annotation database using: -->
<!--   columns(org.Ss.eg.db) -->
<!--   # this database package can also been used for gene ontology enrichment  -->




<!--   # Convert the map list to a data frame -->
<!--   df <- data.frame(entrezgene_id = names(Named_vector_of_HGNC_symbols), -->
<!--                    symbols = sapply(Named_vector_of_HGNC_symbols, function(x) paste(x, collapse = "; ")), -->
<!--                    stringsAsFactors = FALSE) -->

<!--   # Printing the resulting data frame -->
<!--   print(dim(df)) -->

<!--   # check how much symbols mapped to Entrez Gene IDs -->

<!--   # keytypes(org.Ss.eg.db) -->
<!--   length(Named_vector_of_HGNC_symbols) -->
<!--   head(Named_vector_of_HGNC_symbols) -->
<!--   # ensembl_gene_id entrezgene_id -->
<!--   #  ENSSSCG00000008213        396636 -->
<!--   Named_vector_of_HGNC_symbols[['396636']] -->

<!--   # check how much entrez_id map symbols successfully  -->
<!--   print(sum(!is.na(df$symbols) & df$symbols != "")) -->
<!--   print(length(unique(df$symbols) )) -->
<!--   # a few entrez_id map same symbols -->

<!--   # df$entrez_ids == entrez_ids$entrezgene_id -->
<!--   ensembl_entrez <- cbind(df,entrez_ids) -->
<!--   # rownames(ensembl_entrez) <- ensembl_entrez$ensembl_gene_id -->
<!--   ensembl_entrez$entrezgene_id = NULL -->

<!--   # ensembl_entrez[ensembl_entrez$entrezgene_id == 'ENSSSCG00000008213',] -->

<!--   write.csv(ensembl_entrez, "ensembl_entrez_map.csv") -->

<!-- } -->


<!-- ################################## merge symbols which have same ensembl id -->
<!-- if(file.exists("references/ensembl_hgnc_symbol_new_map.csv")) -->
<!-- { -->
<!--   hgnc_symbol_new <- read.csv("references/ensembl_hgnc_symbol_new_map.csv", row.names = 1) -->
<!-- }else -->
<!-- { -->
<!--   library(dplyr) -->
<!--   hgnc_symbol_new <- ensembl_entrez %>% -->
<!--     group_by(ensembl_gene_id) %>% -->
<!--     summarise(symbols = paste(unique(symbols), collapse = ", ")) %>% -->
<!--     ungroup() -->
<!--   hgnc_symbol_new <- as.data.frame(hgnc_symbol_new) -->
<!--   rownames(hgnc_symbol_new) <- hgnc_symbol_new$ensembl_gene_id -->
<!--   write.csv(hgnc_symbol_new, "ensembl_hgnc_symbol_new_map.csv") -->

<!-- } -->




<!-- # print(sum(!is.na(hgnc_symbol_new$symbols) & hgnc_symbol_new$symbols != "")) -->
<!-- # entrez_ids[entrez_ids$ensembl_gene_id == 'ENSSSCG00000008213',] -->
<!-- ensembl_entrez[ensembl_entrez$ensembl_gene_id == 'ENSSSCG00000008213',] -->
<!-- hgnc_symbol_new['ENSSSCG00000008213',] -->
<!-- head(rowData(sce_list[[1]])$hgnc_symbol) -->


<!-- ################################## add new map symbols in sce_list -->
<!--   sce_list <- lapply(sce_list, function(x){ -->
<!--     rowData(x)$hgnc_symbol <- hgnc_symbol_new[rownames(x), 'symbols'] -->
<!--     return(x)}) -->

<!-- ################################# compare the mapping results of two methods   -->
<!-- # show rows which have same ensembl_gene_id, check if they have same symbol annotation  -->
<!-- df <- ensembl_entrez -->

<!-- # Find rows with duplicate 'ensembl_gene_id' values -->
<!-- duplicates <- df[duplicated(df$ensembl_gene_id) | duplicated(df$ensembl_gene_id, fromLast = TRUE), ] -->

<!-- # Output the rows with duplicate 'ensembl_gene_id' -->
<!-- # print(duplicates) -->


<!-- # grep('TNFRSF13B',rowData(sce_list[[1]])$hgnc_symbol_new) -->
<!-- #  -->
<!-- # test = rowData(sce_list[[1]])$hgnc_symbol -->
<!-- # test_new = rowData(sce_list[[1]])$hgnc_symbol_new -->
<!-- # flag = (test==test_new) -->
<!-- # sum(!flag) -->
<!-- #  -->
<!-- # df1 <- data.frame(hgnc_symbol = test[!flag], -->
<!-- #                   hgnc_symbol_new = test_new[!flag]) -->
<!-- #  -->
<!-- #  -->
<!-- # install.packages('rapport') -->
<!-- # library(rapport) -->
<!-- # library(rapportools) -->
<!-- # sum(is.empty(df1$hgnc_symbol)) -->
<!-- # sum(is.empty(df1$hgnc_symbol_new)) -->

<!-- ``` -->


<!-- ### 6.2 Identify and plot mitochondrial genes UMI counts -->
<!-- ```{r  Identify and plot mitochondrial genes counts} -->
<!-- # Identify and plot mitochondrial genes counts -->

<!-- for (i in 1:length(sce_list)) -->
<!-- { -->

<!--   print(length(which(rowData(sce_list[[i]])$chr == 'MT'))) -->
<!--   # obtain how many mitochondrial genes exist in sample 1 -->

<!--   print(rowSums(counts(sce_list[[i]])[which(rowData(sce_list[[i]])$chr == 'MT'),])) -->
<!--   # obtain the count of each mitochondrial genes across all droplets -->
<!--   hist(colSums(counts(sce_list[[i]])[which(rowData(sce_list[[i]])$chr == 'MT'),]), -->
<!--        main= paste0('Mitochondrial UMI counts sample',i), -->
<!--        xlab='UMI counts of mitochondrial transcripts per cell', -->
<!--        ylab= 'Number of droplets', breaks=40) -->
<!--   # obtain the total count of all mitochondrial genes contained in a droplets -->


<!-- } -->

<!-- ``` -->



<!-- all samples have 13 annotated mitochondrial genes detected here, but the UMI count for some of mitochondrial genes are much higher than other mitochondrial genes -->

<!-- most samples have most of their droplets contain low UMI count for mitochondrial genes, so most samples could be fine. -->

<!-- after tuning the parameter in empty droplets filtering step, the new parameter result in reasonable droplets numbers. and the mitochondrial genes counts distribution of sample 2,7 also become reasonable and similar to other samples, which is a great improvement.  -->



<!-- future work: can check function or metabolic pathway of those highly expressed mitochondrial genes, maybe indicate droplets undergoing stress or in the process of dying  -->


<!-- Now we can check various QC plots and metrics (total UMI per droplet, total genes detected per droplet, proportion of UMI counts of mito genes per droplet) of every sample.  -->
<!-- ### 6.3 check and plot various QC paramters  -->
<!-- ```{r check and plot various QC paramters lily, fig.width=10} -->
<!-- library("scuttle") -->
<!-- sce_list <- lapply(sce_list, function(x){ -->
<!--   stats <- perCellQCMetrics(x, subsets=list(Mito=which(rowData(x)$chr=="MT"))) -->
<!--   colData(x)$sum_counts <- stats$sum -->
<!--   colData(x)$genes_detected <- stats$detected -->
<!--   colData(x)$sum_mito <- stats$subsets_Mito_sum -->
<!--   colData(x)$mito_proportion <- stats$subsets_Mito_sum/(stats$sum) -->
<!--   return(x) -->
<!-- })  -->

<!--  lapply(sce_list, function(x){ -->


<!--    # Set the plotting area to 1 row and 3 columns -->
<!-- par(mfrow=c(1, 3)) -->

<!--   hist(x$sum_counts, main=paste0(unique(x$Sample),', total UMI per droplet'), xlab='total UMI counts of a droplet', ylab = 'number of droplet') -->
<!-- abline(v=2000, col="red", lwd=1) # Add a vertical red line at x=2000 -->
<!-- abline(v=5000, col="blue", lwd=1) -->
<!-- # a droplet contain an intact cell always have total UMI counts 2000-5000 -->
<!-- legend("topright", legend=c("2000", "5000"),  -->
<!--        col=c("red", "blue"), lty=2, cex=1.2) -->

<!--   hist(x$genes_detected, main='total genes detected per droplet', xlab='genes detected of a droplet', ylab = 'number of droplet') -->
<!-- abline(v=1000, col="red", lwd=1) -->
<!-- abline(v=3000, col="blue", lwd=1) -->
<!-- legend("topright", legend=c("1000", "3000"),  -->
<!--        col=c("red", "blue"), lty=2, cex=1.2) -->

<!--   hist(x$mito_proportion, main='proportion of UMI counts of mito genes per droplet', xlab='proportion of UMI counts of mito genes per droplet', ylab = 'number of droplet') -->
<!-- abline(v=0.2, col="red", lwd=1)   -->
<!-- legend("topright", legend="0.2",  -->
<!--        col="red", lty=2, cex=1.2) -->

<!-- # Reset to default plotting layout -->
<!-- par(mfrow=c(1, 1))  -->

<!--   plot(x$mito_proportion, x$sum_counts, main=paste0(unique(x$Sample),', mitochondrial proportion versus total UMI count'), xlab='proportion of UMI counts of mito genes per droplet', ylab='total UMI counts of a droplet') -->
<!-- }) -->

<!-- ``` -->
<!-- figure 7. histogram of droplet based on total UMI counts, histogram of droplet based on genes detected, histogram of droplet based on proportion of UMI counts of mito genes, mitochondrial proportion versus total UMI count  -->

<!-- While some level of mitochondrial expression is expected, excessively high levels (commonly around 10-20% of total UMI count) can flag a sample or cell for exclusion from downstream analysis -->

<!-- most samples have most droplets whose UMI count around 2000-5000, detected genes around 1000-3000, mitochondrial UMI proportion around or below 0.05 -->

<!-- samples from run 1 often have a little higher total UMI count than samples from run 2, since we exclude high total UMI count outliers, the distribution of droplets are clearer to visualize and compared.  -->

<!-- droplets that have high proportion of mitochondrial UMI counts and low total UMI counts are usually poor quality droplets containing lysed cell fragments, and will be filtered. -->

<!-- most samples 1-18 only have a small number these droplets, but samples 19-20 contain a lot of these droplets, we will filter droplets with outlier mitochondrial UMI proportions later. -->

<!-- need to check: -->
<!-- still most droplets in Sample 18 have low number of detected genes and low total UMI counts but with low mitochondrial proportion, seem are not lysed cells might be low sequencing depth droplets. droplets in Sample 17 are better than 18.  -->


<!-- we will check more qc metrics of sample 17-18 to decide if discard these low total UMI counts droplets from downstream analysis. -->


<!-- ### 6.4 quickCluster and visualize the droplets in UMAP and color the plots based on QC metrics -->
<!-- Let's now visualize the droplets in UMAP and color the plots based on QC metrics (detected genes, mitochondrial UMI proportion) and main cell types' markers to see if the suspicious droplets and samples represent doublets or an odd cell type. -->

<!-- ```{r run PCA and UMAP individually in each sample} -->
<!-- sce_list -->
<!-- if(file.exists("preprocess6.4.RData")) -->
<!-- { -->
<!--   load("preprocess6.4.RData") -->
<!-- }else -->
<!-- { -->
<!--   library(scran) -->
<!--   library(scater) -->
<!--   library(uwot) -->
<!--   for(i in 1:length(sce_list)){ -->
<!--     set.seed(100)  -->
<!--     clust <- quickCluster(sce_list[[i]]) -->

<!--     sce_list[[i]] <- computeSumFactors(sce_list[[i]], cluster=clust) -->
<!--     sce_list[[i]] <- logNormCounts(sce_list[[i]]) -->

<!--     sce_list[[i]] <- fixedPCA(sce_list[[i]], subset.row=NULL) ## note intentionally not using HVGs here as this is a rough visualization to see all effects on droplets -->
<!--     sce_list[[i]] <- runUMAP(sce_list[[i]], dimred='PCA') -->
<!--   } -->

<!--     save.image(file = paste0(work_dir,"preprocess6.4.RData")) -->
<!-- } -->

<!-- sce_list -->
<!-- ``` -->

<!-- Let's have a look at samples suspicious samples 17-18, and normal samples 1-16, 19-20 and check QC parameters and marker genes for typical cell types to determine whether QC-outliers are of one specific cell type. -->

<!-- ```{r check suspicious and normal samples QC plots and marker genes expression distribution, fig.width=12} -->
<!-- # install.packages("patchwork") -->

<!-- library(patchwork) -->
<!-- # c(17:20,1:16) -->

<!-- for(i in 1:length(sce_list)) -->
<!-- { -->
<!--   factor_vector  <- rep('normal_range',dim(sce_list[[i]])[2]) -->
<!--   factor_vector[(sce_list[[i]]$sum_counts <= 1000)] <- 'below_1000' -->
<!--   factor_vector[(sce_list[[i]]$sum_counts >= 20000)] <- 'above_20000' -->
<!--   factor_vector <- as.factor(factor_vector) -->
<!--   sce_list[[i]]$sum_counts_level <- factor_vector -->

<!-- } -->




<!-- for (sample in c(17:20,1:16))  -->
<!-- { -->
<!--   x <- sce_list[[sample]] -->

<!--   p1 <- plotReducedDim(x, 'UMAP', colour_by = 'sum_counts') + ggtitle(unique(x$Sample))+ coord_fixed() -->
<!--   p2 <- plotReducedDim(x, 'UMAP', colour_by = 'genes_detected') + ggtitle(unique(x$Sample))+ coord_fixed() -->
<!--   p3 <- plotReducedDim(x, 'UMAP', colour_by = 'mito_proportion') + ggtitle(unique(x$Sample))+ coord_fixed() -->


<!--   # Combine the plots side by side -->
<!--   p_combined <- p1 + p2 + p3 + plot_layout(ncol = 3) -->

<!--   # Display the combined plot -->
<!--   print(p_combined)  -->


<!--   p1 <- plotReducedDim(x, 'UMAP', colour_by = 'sum_counts_level') + ggtitle(unique(x$Sample))+ coord_fixed() -->
<!--   print(p1) -->

<!-- ########################################   -->
<!--   p1 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'HNMT')]) + ggtitle(paste(unique(x$Sample),'myeloid, HNMT')) ## myeloid -->

<!--   p2 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'CD3E')]) + ggtitle('T cells, CD3E gene') ## T cells -->

<!--   p3 <- plotReducedDim(x, 'UMAP', colour_by=rownames(sce_list[[1]])[which(rowData(sce_list[[1]])$Symbol == 'PAX5')]) + ggtitle('B cells, PAX5 gene') ## B cells   -->
<!--   # Set uniform color scale limits  -->
<!--   # color_limits <- c(0, 5) -->
<!--   color_set <- c("blue", "green", "yellow") -->

<!--   p1 <- p1 + coord_fixed() + -->
<!--     scale_color_gradientn(colors = color_set) +  -->
<!--     theme(legend.position = "right", aspect.ratio = 1)  -->

<!--   p2 <- p2 + coord_fixed() + -->
<!--     scale_color_gradientn( colors = color_set) + -->
<!--     theme(legend.position = "right", aspect.ratio = 1) -->

<!--   p3 <- p3 + coord_fixed() + -->
<!--     scale_color_gradientn( colors = color_set) + -->
<!--     theme(legend.position = "right", aspect.ratio = 1) -->

<!--   # Combine plots with patchwork side by side -->
<!--   p_combined <- (p1 | p2 | p3) & -->
<!--     plot_layout( guides = "collect")  -->
<!--   # Print the combined plot -->
<!--   print(p_combined) -->


<!-- } -->

<!-- ``` -->

<!-- figure 8. visualize the droplets in UMAP and color the droplets based on QC metrics (total UMI counts, detected genes, proportion of mito UMI counts, 3 levels of total UMI counts,  marker gene of B cells, T cells, myeloid ) -->

<!-- histogram of droplet based on total UMI counts, histogram of droplet based on genes detected, histogram of droplet based on proportion of UMI counts of mito genes, mitochondrial proportion versus total UMI count  -->

<!-- need to check: -->
<!-- we can observe three kinds of droplets distribution in UMAP, whose total UMI above 20,000, within normal range, or below 1000 to check if abnormal total UMI droplets cluster together. these droplets often locate in several different clusters, it seems fine? would  these abnormal total UMI droplets be an issue ?would filtering them necessary? -->

<!-- From this examination, it appears that none of the major cell types are specifically characterized by high mitochondrial proportion, and most samples have few high mitochondrial proportion droplets. -->

<!-- so I think that we can safely filter high mitochondrial proportion droplets out.  -->

<!-- It would be difficult to justify retaining these outliers. We can always revisit this later to look at marker genes for the droplets if desired. -->

<!-- need to check: -->
<!-- sample 19-20 looks fine in their plots above, but sample 17-18 seems fail to cluster due to their low detected genes and total UMI, results in not enough gene expression information to cluster and separate these droplets. namely, the algorithm fail to detect the significant difference between droplets. this will cause difficulty in assign cell types for these droplets. we might remove these low total UMI droplets from downstream analysis.  -->


<!-- need to check: -->
<!-- in a few samples, the B cells number is quiet low. not sure it is okay? -->

<!-- need to check: -->
<!-- would that be reasonable to filter out droplets with outlier high mitochondrial proportion? or filter out droplets based on absolute mitochondrial proportion > 0.2? -->

<!-- here still filter out droplets with outlier high mitochondrial proportion within each sample and show filtered droplets in UMAP plots -->

<!-- ### 6.5 filter out droplets with outlying mitochondrial proportion within each sample and show these filtered droplets in UMAP -->
<!-- ```{r filter out droplets with outlying mitochondrial proportion within each sample and show these filtered droplets in UMAP, fig.width=5, fig.height=5}  -->
<!-- # if(file.exists("preprocess6.5.RData")) -->
<!-- # { -->
<!-- #   load("preprocess6.5.RData") -->
<!-- # }else -->
<!-- # { -->
<!-- # } -->

<!--   for(i in 1:length(sce_list)){ -->
<!--    # sce_list[[i]]$mito_outlier <- (sce_list[[i]]$mito_proportion >= 0.2)  -->
<!--    # print(i) -->
<!--    # print(table(sce_list[[i]]$mito_outlier)) -->
<!--    #  cols <- rep('black', length(sce_list[[i]]$mito_outlier)) -->
<!--    #  cols[sce_list[[i]]$mito_outlier] <- 'red' -->
<!--    #  -->
<!--    #  plot(sce_list[[i]]$mito_proportion, sce_list[[i]]$sum_counts, main=paste0(i,', mitochondrial proportion versus total UMI count'), col=cols, xlab='proportion of UMI counts of mito genes per droplet', ylab='total UMI counts of a droplet') -->


<!--     sce_list[[i]]$mito_outlier <- isOutlier(sce_list[[i]]$mito_proportion, type='higher',nmads = 5) -->
<!--     print(table(sce_list[[i]]$mito_outlier)) -->
<!--     cols <- rep('black', length(sce_list[[i]]$mito_outlier)) -->
<!--     cols[sce_list[[i]]$mito_outlier] <- 'red' -->


<!--     p1 <- plot(sce_list[[i]]$mito_proportion, sce_list[[i]]$sum_counts, main=paste0('sample', i,', mitochondrial proportion versus total UMI count'), col=cols, xlab='proportion of UMI counts of mito genes per droplet', ylab='total UMI counts of a droplet') -->

<!--     print(p1) -->
<!--     p2 <- plotReducedDim(sce_list[[i]], 'UMAP', colour_by='mito_outlier') -->
<!--     print(p2) -->




<!--     sce_list[[i]] <- sce_list[[i]][,!sce_list[[i]]$mito_outlier] -->
<!--   } -->

<!-- # save.image(file = paste0(work_dir,"preprocess6.5.RData")) -->



<!-- ``` -->

<!-- figure 9. visualize filtered droplets in UMAP and mitochondrial proportion versus total UMI count plot -->

<!-- ```{r  overview number of droplets per sample } -->
<!-- multi_list_bar_plot(sce_list,"after qc filtering") -->

<!-- ``` -->

<!-- figure 10. overview number of droplets per sample -->

<!-- ## 7 Checking for doublets -->

<!-- Here we annotate suspected doublets within individual samples. We will wait to remove these droplets until after we have merged and clustered everything because it can be the case that a whole cluster is primarily doublets and we want to be sure we detect that. -->

<!-- ```{r calculate cxds scores to Check doublets and show cxds scores on UMAP, fig.width=5, fig.height=5 } -->



<!-- # BiocManager::install("scds") -->
<!-- library(scds) -->
<!-- for(i in 1:length(sce_list)){ -->
<!--   sce_list[[i]] <- cxds(sce_list[[i]]) -->
<!--   sce_list[[i]]$query_doublet <- isOutlier(sce_list[[i]]$cxds_score, nmads=5, type='higher') -->
<!--   p <- plotReducedDim(sce_list[[i]], 'UMAP', colour_by='cxds_score') + ggtitle(paste('sample',i)) -->

<!--   print(p) -->

<!--   print(i) -->
<!--   print(table(sce_list[[i]]$query_doublet)) -->
<!--   cols <- rep('black', length(sce_list[[i]]$query_doublet)) -->
<!--   cols[sce_list[[i]]$query_doublet] <- 'red' -->


<!--     p1 <- plot(sce_list[[i]]$cxds_score, sce_list[[i]]$sum_counts, main=paste0('sample', i,', cxds_score versus total UMI count'), col=cols, xlab='cxds_score per droplet', ylab='total UMI counts of a droplet') -->

<!--     print(p1)   -->
<!-- } -->

<!-- ``` -->

<!-- figure 11. visualize the droplets in UMAP and color the droplets based on cxds scores, visualize droplets with outlying cxds scores in mitochondrial  -->
<!-- proportion versus total UMI count plot -->

<!-- need to check: -->
<!-- sample 18 have quiet compact cluster distribution in UMAP compared with remaining samples. -->


<!-- cxds_score can be affected by sequencing run, cell type constitution, so this varies vastly across samples  -->

<!-- there are some high cxds_score droplets in most samples, but the scores are fairly incomparable across samples.  -->
<!-- For this reason we've marked those droplets within each sample that are more than 3 MAD above the median. We will revisit whether these are doublets after merging, normalizing and clustering. -->



<!-- # Normalization and merging -->

<!-- need to check: -->
<!-- Now we will properly normalize the data and merge all of the samples together. We will keep Sample 17-18 tentatively even though they have lots of low total UMIs droplets. will check whether low total UMIs droplets cluster together.  -->


<!-- ## 8.1 merge samples from same run, exclude catastrophic samples if any -->
<!-- ```{r exclude catastrophic samples if any and merge samples from same run} -->

<!-- if(file.exists("preprocess8.1.RData")) -->
<!-- { -->
<!--   load("preprocess8.1.RData") -->
<!-- }else -->
<!-- { -->

<!--   for(i in 1:length(sce_list)){ -->
<!--     colnames(sce_list[[i]]) <- paste0(colnames(sce_list[[i]]), '-', as.character(i)) -->
<!--   } -->
<!--   names(sce_list) <- paste0('sample', c(1:20)) -->

<!--   sce_run_1 <- do.call('cbind', sce_list[c(1:8)]) -->
<!--   sce_run_2 <- do.call('cbind', sce_list[c(9:16)]) -->
<!--   sce_run_3_lung <- do.call('cbind', sce_list[c(19:20)]) -->
<!--   sce_run_3_BAL <- do.call('cbind', sce_list[c(17:18)]) -->


<!--      save.image(file = "preprocess8.1.RData") -->
<!-- } -->



<!-- ``` -->


<!-- ```{r add conditions annotation to data} -->

<!-- setwd(work_dir) -->
<!-- annot <- read.csv('references/sample_biology.csv') -->
<!-- samps <- paste("sample", formatC(1:20, width=2, flag="0"), sep="_") -->
<!-- annot$sample <- samps -->
<!-- rownames(annot) <- annot$sample -->
<!-- ``` -->

<!-- then we perform quick clustering and normalization by pooled factors.  -->


<!-- Now we look at the data to determine whether there are batch effects that ought to be removed between samples. Here we will look for the top 5000 highly variable genes for biological variability and use these for PCA and other dimensional reduction as these are the genes we will ultimately be interested in. -->

<!-- Whether or not sample should be specified in the design for determining biological variability is an interesting question.  -->
<!-- In some ways, sample corresponds to a batch effect in that technical effects of (e.g. sample preparation procedures) will be captured here.  -->
<!-- In other ways, it is a biological effect as different samples underwent different experimental treatments (here i suspect our samples are from inbreeding individual? ) and these would be expected to impact gene expression in quite a profound way. -->

<!-- 1-8 samples from inbreed individual. -->
<!-- Given that 1-8 samples come from the same sequencing run and presuming consistent sample preparation and processing, 1-4 under condition 1, 5-8 under condition 2, so we expect the sample 1-4 ideally should be same, sample 5-8 ideally should be same.  -->

<!-- since we want to preserve the true biological signals related to the different conditions so that we can research the biological difference between these 2 conditions. -->

<!-- so we shouldn't batch correct between 2 conditions (ISU1 vs 135) to avoid -->
<!-- diminishing the biological differences between these 2 conditions. so we don't apply Batch Effect Correction here.  -->

<!-- maybe in the future. when we merge sample 1-18, we could apply Batch Effect Correction based on batch variables sequencing run. -->


<!-- after we cluster, we can check whether droplets from ISU1 are assigned into different cluster compared with droplets from 135. -->

<!-- if same cell type are assigned into different cluster due to condition, then it would be interesting, indicating same immune cell type behave differently under these 2 conditions. -->


<!-- i previously apply Batch Effect correct between all samples in same run to exclude the possible genome difference between different individuals and random effect. but i noticed this cause the difference between 2 conditions reduced. so i don't risk diminishing the biological differences between these 2 conditions in order to apply Batch Effect correct between all samples. -->


<!-- ## 8.2 Normalization gene expression counts, conduct batch effect correction  -->
<!-- ```{r perform quick clustering and normalization for run 1 samples and plot qc metrics before and after batch effect correction} -->


<!-- normalize_visualize_save <- function(current_run,sce_run_1, Correct_batch_effects = FALSE) -->
<!-- { -->
<!--   if(file.exists(paste0('sce_',current_run,'.rds'))) -->
<!-- { -->
<!--   sce_run_1 <- readRDS(paste0('sce_',current_run,'.rds')) -->
<!-- }else -->
<!-- { -->
<!--   # log transfrom gene counts -->

<!--   print(sce_run_1) -->
<!--   clust <- quickCluster(sce_run_1) -->
<!--   size_factors <- calculateSumFactors(sce_run_1, cluster=clust) -->
<!--   sce_run_1 <- logNormCounts(sce_run_1, size_factors=size_factors) -->
<!--   print(sce_run_1) -->

<!--   # get the Top 5000 High Variable Genes -->
<!--   varmod <- modelGeneVar(sce_run_1, block=sce_run_1$Sample) -->
<!--   # Model the variance of the log-expression profiles for each gene, decomposing it into technical and biological components based on a fitted mean-variance trend. -->
<!--   vargenes <- getTopHVGs(varmod, n=5000) -->
<!--   length(vargenes) -->

<!--   sce_run_1 <- fixedPCA(sce_run_1, subset.row=vargenes)  -->
<!--   set.seed(100) -->
<!--   sce_run_1 <- runUMAP(sce_run_1, dimred='PCA') -->
<!--   set.seed(100) -->
<!--   sce_run_1 <- runTSNE(sce_run_1, dimred='PCA') -->
<!--   sce_run_1$condition <- annot[as.character(sce_run_1$Sample), 'Condition'] -->
<!--   sce_run_1$run <- current_run -->


<!--   saveRDS(sce_run_1, file = paste0('sce_',current_run,'.rds')) -->

<!-- } -->

<!--   ###################### plot -->
<!--   title_suffix_1 = '\n based on PCA componets calculated using top 5000 highly variable genes' -->


<!--   set.seed(100) -->
<!--   ord <- sample.int(ncol(sce_run_1), ncol(sce_run_1), replace=FALSE) -->
<!--   pl <- plotUMAP(sce_run_1[,ord], colour_by="Sample") + ggtitle(paste(current_run,'UMAP uncorrected', title_suffix_1)) -->
<!--   print(pl) -->


<!--   pl <- plotUMAP(sce_run_1[,ord], colour_by="condition") + ggtitle(paste(current_run,'UMAP uncorrected', title_suffix_1)) -->
<!--   print(pl) -->

<!--   pl <- plotReducedDim(sce_run_1, 'UMAP', colour_by=rownames(sce_run_1)[which(rowData(sce_run_1)$Symbol == 'HNMT')]) + ggtitle(paste(current_run,'myeloid highlighted (HMNT)','UMAP uncorrected', title_suffix_1)) ## myeloid -->
<!--   print(pl) -->

<!--   pl <- plotReducedDim(sce_run_1, 'UMAP', colour_by=rownames(sce_run_1)[which(rowData(sce_run_1)$Symbol == 'CD3E')]) + ggtitle(paste(current_run,'T cells highlighted (CD3E)', 'UMAP uncorrected', title_suffix_1)) ## T cells -->
<!--   print(pl) -->

<!--   pl <- plotReducedDim(sce_run_1, 'UMAP', colour_by=rownames(sce_run_1)[which(rowData(sce_run_1)$Symbol == 'PAX5')]) + ggtitle(paste(current_run,'B cells highlighted (PAX5)', 'UMAP uncorrected', title_suffix_1)) ## B cells -->
<!--   print(pl) -->
<!--   # correct between all samples in same run and plot qc metrics and UMAP and save workspace and rds for run 1 samples -->

<!--   p1 <- plotReducedDim(sce_run_1, 'UMAP', colour_by = 'sum_counts_level') + ggtitle(paste(current_run, 'UMAP uncorrected', title_suffix_1)) -->
<!--   print(p1) -->


<!--   varmod <- modelGeneVar(sce_run_1, block=sce_run_1$Sample) -->
<!--   vargenes <- getTopHVGs(varmod, n=5000) -->

<!--   if(Correct_batch_effects) -->
<!--   { -->
<!--     library(batchelor) -->
<!--     set.seed(100) -->
<!--     sce_run_1_corrected <- fastMNN(sce_run_1, batch=sce_run_1$Sample, subset.row = vargenes, get.variance = TRUE) -->
<!--   # # produce batch effects Corrected data -->
<!--   #   if(file.exists(paste0('sce_',current_run,'_corrected.rds'))) -->
<!--   # { -->
<!--   #   sce_run_1_corrected <- readRDS(paste0('sce_',current_run,'_corrected.rds')) -->
<!--   #   print(paste0("skip save ",'sce_',current_run,'_corrected.rds','since it already exist to save running time')) -->
<!--   # }else -->
<!--   #   { -->
<!--   #    -->
<!--   #   } -->
<!--   }else -->
<!--   { -->
<!--   library(batchelor) -->
<!--   set.seed(100) -->
<!--   sce_run_1_corrected <- fastMNN(sce_run_1, batch=sce_run_1$run, subset.row = vargenes, get.variance = TRUE) -->
<!--   sce_run_1_corrected -->
<!--   } -->

<!--   set.seed(100) -->
<!--   sce_run_1_corrected <- runUMAP(sce_run_1_corrected, dimred="corrected") -->
<!--   sce_run_1_corrected <- runTSNE(sce_run_1_corrected, dimred="corrected") -->

<!--   sce_run_1_corrected$Sample <- sce_run_1$Sample -->
<!--   sce_run_1_corrected$condition <- sce_run_1$condition -->
<!--   sce_run_1_corrected$cxds_score <- sce_run_1$cxds_score -->
<!--   sce_run_1_corrected$query_doublet <- sce_run_1$query_doublet -->
<!--   sce_run_1_corrected$sum_counts <- sce_run_1$sum_counts -->
<!--   sce_run_1_corrected$genes_detected <- sce_run_1$genes_detected -->
<!--   sce_run_1_corrected$mito_proportion <- sce_run_1$mito_proportion -->
<!--   sce_run_1_corrected$hgnc_symbol <- sce_run_1$hgnc_symbol -->
<!--   sce_run_1_corrected$sum_counts_level <- sce_run_1$sum_counts_level -->

<!--   rowdat <- rowData(sce_run_1_corrected) -->
<!--   rowdat <- cbind(rowdat, rowData(sce_run_1_corrected)[rownames(sce_run_1_corrected),]) -->

<!--   rowData(sce_run_1_corrected) <- rowdat -->


<!--   if(file.exists(paste0('sce_',current_run,'_corrected.rds'))) -->
<!--   { -->
<!--     print(paste0("skip save ",'sce_',current_run,'_corrected.rds','since it already exist to save running time')) -->
<!--   }else -->
<!--   { -->
<!--     saveRDS(sce_run_1_corrected, file = paste0('sce_',current_run,'_corrected.rds'))    -->
<!--   }     -->
<!--   print(sce_run_1_corrected)     -->
<!--   ###################### plot  -->
<!--   set.seed(100) -->
<!--   ord <- sample.int(ncol(sce_run_1_corrected), ncol(sce_run_1_corrected), replace=FALSE) -->

<!--   title_suffix_2 = '\n based on top 5000 highly variable genes' -->

<!--   pl <- plotUMAP(sce_run_1_corrected[,ord], colour_by="Sample") + ggtitle(paste(current_run,'UMAP uncorrected',title_suffix_2)) -->
<!--   print(pl) -->

<!--   pl <- plotUMAP(sce_run_1_corrected[,ord], colour_by="condition") + ggtitle(paste(current_run,'UMAP uncorrected',title_suffix_2)) -->
<!--   print(pl) -->

<!--   pl <- plotUMAP(sce_run_1_corrected[,ord], colour_by="cxds_score") + ggtitle(paste(current_run,'UMAP uncorrected',title_suffix_2)) -->
<!--   print(pl) -->

<!--   pl <- plotUMAP(sce_run_1_corrected[,ord], colour_by="query_doublet") + ggtitle(paste(current_run,'UMAP uncorrected',title_suffix_2)) -->
<!--   print(pl) -->
<!--   # lapply(sce_run_1_corrected, dim) -->

<!--   p1 <- plotUMAP(sce_run_1_corrected[,ord], colour_by = 'sum_counts_level') + ggtitle(paste(current_run, 'UMAP uncorrected', title_suffix_2)) -->
<!--   print(p1)   -->
<!-- } -->

<!-- current_run <- "run_1" -->
<!-- normalize_visualize_save(current_run,sce_run_1, Correct_batch_effects = FALSE) -->


<!-- # title_suffix = 'corrected by samples in same run' -->
<!-- ``` -->

<!-- figure 12.1 visualize the droplets in UMAP and color the droplets based on sample;  condition; marker gene expression level of B cells, T cells, myeloid;  cxds scores; suspected doublet for sequencing run 1 -->



<!-- after batch effect correction. we observe droplets from 8 samples distribute more uniformly in UMAP plots. droplets from 2 conditions distribute apprantly more uniformly. -->

<!-- We will leave this analysis here and start a new script where we cluster, subcluster, filter doublets, etc. -->





<!-- ```{r perform quick clustering and normalization for run 2 samples and plot qc metrics before batch effect correction} -->
<!-- current_run <- "run_2" -->
<!-- normalize_visualize_save(current_run,sce_run_2, Correct_batch_effects = FALSE) -->

<!-- ``` -->

<!-- figure 12.2 visualize the droplets in UMAP and color the droplets based on sample;  condition; marker gene expression level of B cells, T cells, myeloid;  cxds scores; suspected doublet for sequencing run 2 -->

<!-- ```{r perform quick clustering and normalization for run 3 samples and plot qc metrics before batch effect correction} -->


<!-- normalize_visualize_save("run_3_BAL",sce_run_3_BAL, Correct_batch_effects = FALSE) -->
<!-- normalize_visualize_save("run_3_lung",sce_run_3_lung, Correct_batch_effects = FALSE) -->




<!-- ``` -->

<!-- figure 12.3 visualize the droplets in UMAP and color the droplets based on sample;  condition; marker gene expression level of B cells, T cells, myeloid;  cxds scores; suspected doublet for sequencing run 3 -->
<!-- # 9 save image and show session Info -->
<!-- ```{r sessionInfo} -->
<!-- save.image(file = "preprocess9.RData") -->
<!-- sessionInfo() -->

<!-- ``` -->



